---
title: "Neural_Network"
output: html_document
date: "2025-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Set personal working directory
setwd("~/Documents/Repos/GRP-6_DS-Project")
renv::activate()
```

```{r safe-caffeinate, echo=FALSE}
# SAFE WAKE LOCK
# 1. -i : Prevents System Sleep (CPU stays on), but allows Screen Sleep.
# 2. -w : Waits for the current R Process ID (Sys.getpid()).
# 3. Result: When this R session ends (or crashes), the wake lock dies automatically.
system(sprintf("caffeinate -i -w %d", Sys.getpid()), wait = FALSE)
```
## Type of Network
The dataset is tabular data with numerical features and a categorical target variable. A feedforward neural network is the most suitable type of network for this classification task.

A convolutional neural network (CNN) is typically used for image data, while a recurrent neural network (RNN) is more suited for sequential data such as time series or text. Since our data does not have spatial or sequential characteristics, a feedforward neural network is the best choice.

```{r libraries}
library(tfruns)
library(tictoc)
library(beepr)
```

## Changed metric
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_no_Fold.r")
beep(sound=5)
toc()
```

## Changed metric
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_metric.r")
beep(sound=5)
toc()
```

## Oversample
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_oversample.r")
beep(sound=5)
toc()
```

## SMOTE
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_SMOTE.r")
beep(sound=5)
toc()
```

## Moderate Weight 
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_moderate_weight.r")
beep(sound=5)
toc()
```

## Focal Loss
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_focal_loss.r")
beep(sound=5)
toc()
```

## SMOTE (+ Class weight)
```{r run the experiment}
tic()
training_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_SMOTE_weight.r")
beep(sound=5)
toc()
```




# --- Architecture Search ---
# Goal: Find the balance between Model Size (Units) and Overfitting (Dropout)
```{r}
runs_grid <- list(
  # --- VARY THESE (Architecture) ---
  
  # 1. Width (Number of Neurons)
  # Testing a "Narrow" vs "Wide" first layer
  units1 = c(128,256),
  
  # Testing tapering down in the second layer
  units2 = c(64,128),
  
  # 2. Depth (Number of Layers)
  # 0 = 2-Layer Model (The logic in your script handles this)
  # 64/128 = 3-Layer Model
  units3 = c(0, 64),
  
  # --- KEEP THESE CONSTANT (Controls) ---
  
  # Fix dropout to a baseline (e.g., 0) so we compare architectures fairly
  dropout1 = c(0.0),
  dropout2 = c(0.0), 
  dropout3 = c(0.0),
  
  # Fix optimization dynamics
  learning_rate = c(0.001), 
  batch_size = c(128)
)
```


# Execute the Tuning Run
```{r}
tic()
tuning_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_no_Fold.r", 
           runs_dir = "tuning_architecture_search", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = FALSE)

beep(sound=5)
toc()
```


# --- View Results ---
# List top 5 models sorted by Validation Accuracy
```{r}
# Retrieve runs, ordered by lowest Validation Loss first
results <- ls_runs(runs_dir = "tuning_architecture_search", 
                   order = metric_val_loss, 
                   decreasing = FALSE) # FALSE because lower loss is better

print(results)
# Display the winning hyperparameters
# I added flag_units3 since you are testing depth now as well
head(results[, c("flag_units1", "flag_units2", "flag_units3",
                 "metric_val_loss", "metric_val_accuracy")], 5)
```



# Execute the Tuning Run
```{r}
tic()
tuning_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_SMOTE_weight.r", 
           runs_dir = "tuning_architecture_search", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = FALSE)

beep(sound=5)
toc()
```


# --- View Results ---
# List top 5 models sorted by Validation Accuracy
```{r}
# Retrieve runs, ordered by lowest Validation Loss first
results <- ls_runs(runs_dir = "tuning_architecture_search", 
                   order = metric_val_loss, 
                   decreasing = FALSE) # FALSE because lower loss is better

# Display the winning hyperparameters
# I added flag_units3 since you are testing depth now as well
head(results[, c("flag_units1", "flag_units2", "flag_units3", 
                 "metric_val_loss", "metric_val_accuracy")], 5)
```

# Execute the Tuning Run
```{r}
tic()
tuning_run("/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/Assignment_2_MLP_Moderate_weight.r", 
           runs_dir = "tuning_architecture_search", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = FALSE)

beep(sound=5)
toc()
```


# --- View Results ---
# List top 5 models sorted by Validation Accuracy
```{r}
# Retrieve runs, ordered by lowest Validation Loss first
results <- ls_runs(runs_dir = "tuning_architecture_search", 
                   order = metric_val_loss, 
                   decreasing = FALSE) # FALSE because lower loss is better

# Display the winning hyperparameters
# I added flag_units3 since you are testing depth now as well
head(results[, c("flag_units1", "flag_units2", "flag_units3", 
                 "metric_val_loss", "metric_val_accuracy")], 5)
```










