To Do
- do not look at plots on all data but only on test set so there is no data leakage
- Fix income distributions to log_normal and use this

C. Misinterpretation of "Non-linearity" in Factors

Issue: In the discussion of term (36 vs 60 months), the text says: "captures a clear non-linear effect... that cannot be represented well by a simple linear trend." Correction: Since term is a factor (binary in this case), a linear model handles this perfectly via a dummy variable (intercept shift). This is not "non-linearity" in the mathematical sense (like x2 or log(x)); it is a discrete effect. The linear model is perfectly capable of modeling this.
  
  
  

```{r}
# Define the full path to the saved CSV file
input_file_path <- "/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_data/Intermediate/cleaned_data.csv"

# Read the CSV file back into a data frame, typically giving it a new name
cleaned_data <- read.csv(input_file_path)
```

--------------------------------------------------------------------------------------------------------
```{r}
library(Hmisc)     # histograms, general EDA
library(corrplot)  # correlation plots
library(ggplot2)   # nicer plots & loops
library(tidyverse)
library(car)       #VIF, partial residual plots, alternative and faster would be library(performance and use check_collinearity)
#library(lmtest)    # Breusch-Pagan test (optional)
```
# 1. Data and Goal

In this document we:

- Start from a pre-cleaned dataframe `cleaned_data`.
- Perform **EDA and structural checks** as recommended in the lecture:
  - histograms,
  - scatterplots for numeric–numeric, factor–numeric, factor–factor,
  - correlation analysis for numeric variables.
- Create a **train / validation / test split** (70/15/15, seed 123).
- Use **complete cases only** for this document (`na.omit()`), because:
  - a separate script/programmer will handle imputation for the final model pipeline,
  - here we only need a fully observed subset to check linear regression assumptions.
- Fit a **linear regression model on the training set only**.
- Run all **mandatory linear regression diagnostics**:
  - linearity,
  - homoscedasticity,
  - normality of residuals,
  - multicollinearity (VIF),
  - influential points / leverage.

```{r}
str(cleaned_data)
summary(cleaned_data)
```

# 2. Complete-Case Subset for Diagnostic Work

Imputation will be done later by another programmer for the final modelling pipeline.  
For the **assumption checks in this document**, we use a **complete-case subset**:

- This avoids mixing in imputation decisions here.
- It guarantees that all diagnostics (residuals, VIF, etc.) are computed on fully observed rows.


```{r complete-cases}
cleaned_data_cc <- na.omit(cleaned_data)

cat("Original number of rows:", nrow(cleaned_data), "\n")
cat("Complete-case rows used in this document:", nrow(cleaned_data_cc), "\n")
```


---

# 3. Global EDA Before Any Split

We first explore the data structure and basic distributions before any split or model fitting.

## 3.1 Histograms for All Numeric Variables (Hmisc)

We use `Hmisc::hist.data.frame()` to inspect the distributions of all numeric variables:

- to detect skewness,
- to spot extreme values/outliers,
- to think about possible transformations later (e.g. logs).


```{r}
#library(Hmisc)
#hist.data.frame(cleaned_data_cc)


numeric_vars <- names(cleaned_data_cc)[sapply(cleaned_data_cc, is.numeric)]

for (v in numeric_vars) {
  hist(cleaned_data_cc[[v]],
       main = paste("Histogram of", v),
       xlab = v,
       col = "lightblue")
}
```
```{r}
table(cleaned_data$acc_now_delinq)
table(cleaned_data$acc_now_delinq_bin)
table(cleaned_data$delinq_2yrs)
```


## 3.2 Pairwise Relationships

We want to see:

- **numeric vs numeric** → scatterplots,
- **factor vs numeric** → boxplots,
- **factor vs factor** → mosaic-like association plots.

Because this can generate a *lot* of plots, we show generic code and explain that in practice you might restrict to a subset of key variables.

### 3.2.1 Helper functions to identify types

```{r type-helpers}
is_num   <- sapply(cleaned_data_cc, is.numeric)
is_fact  <- sapply(cleaned_data_cc, is.factor)

num_vars  <- names(cleaned_data_cc)[is_num]
fact_vars <- names(cleaned_data_cc)[is_fact]

num_vars
fact_vars
```


### 3.2.2 Numeric vs Numeric (Scatterplot Matrix)

Scatterplot matrices give an overview of linear relationships and potential collinearity.


```{r scatter-numeric-numeric, fig.height=8, fig.width=8}
# Only numeric columns we only take a subset to since we have so many variables
set.seed(123)
sampled_df <- cleaned_data_cc[sample(nrow(cleaned_data_cc), 20000), ]

pairs(sampled_df[num_vars],
      main = "Scatterplot Matrix (20k Sample)")

```
To examine the pairwise relationships between the numeric predictors, we created a scatterplot
matrix using a simple random sample of 20,000 observations.  
The full dataset contains more than 700,000 rows, and visualizing all observations would be 
computationally prohibitive and would produce unreadable overplotted graphics.  
Sampling is standard practice for exploratory data analysis: a random subset preserves the 
general structure and relationships in the data while making visualization feasible.

Representativeness of the Sample

The 20k subset was drawn using a fixed random seed to ensure reproducibility.  
Because the sampling was completely random and no filtering, stratification, or sorting was 
performed beforehand, the resulting subset is representative of the full dataset for the purpose
of visual inspection.  
Thus, no systematic sampling bias is expected.

Relationship Between `int_rate` and Numeric Predictors

Inspection of the scatterplot panels involving the target variable `int_rate` reveals several 
important patterns:

- **`annual_inc` vs. `int_rate`:**  
  There is a clear monotonic *decreasing* trend: borrowers with higher income tend to receive
  lower interest rates.  
  The relationship appears curved rather than strictly linear, which supports later use of 
  log-transformations for income-related variables.

- **`dti` vs. `int_rate`:**  
  Higher debt-to-income ratios are associated with higher interest rates.  
  The pattern is wedge-shaped rather than linear, indicating mild non-linearity.

- **`inq_last_6mths` vs. `int_rate`:**  
  Being a discrete count variable, the scatter shows vertical stacks at integer values.  
  Borrowers with multiple recent inquiries appear to have slightly higher interest rates on 
  average, but the relationship is stepwise rather than linear.

- **`loan_amnt` vs. `int_rate`:**  
  Only a weak, diffuse trend is visible.  
  Loan amount alone does not strongly determine the interest rate.

- **Monetary balance variables (e.g., `revol_bal_log`, `tot_cur_bal_log`, `total_rev_hi_lim_log`):**  
  These log-transformed features show mild upward associations with `int_rate`.  
  The transformation removes extreme skewness, and the resulting relationships are smooth,
  monotonic, and appropriate for linear modelling.

Overall, `int_rate` exhibits mostly monotonic—but not always linear—relationships with its predictors.
Log-transformation of heavily skewed monetary variables appears justified.

Relationships Between Predictors (Multicollinearity Considerations)

The scatterplot matrix also highlights several strong associations *between* predictors:

- **`open_acc` and `total_acc`:**  
  These variables are tightly and linearly related, forming a narrow diagonal band.  
  This reflects the inherent constraint that open accounts are a subset of total accounts.  
  The strong correlation suggests potential multicollinearity if both are included in a regression
  model.

- **Credit exposure variables (`revol_bal_log`, `tot_cur_bal_log`, `total_rev_hi_lim_log`):**  
  These features exhibit strong positive relationships with one another, indicating they measure
  similar underlying borrower characteristics.  
  Their inclusion together in an OLS model may inflate variance inflation factors (VIF), and
  variable selection or dimensionality reduction may be appropriate.

- **Loan-related variables (`loan_amnt`, `funded_amnt`, etc.):**  
  These (if present) often show near-linear relationships, as they represent closely related
  quantities in the LendingClub dataset.

Implications for Modelling

The scatterplot matrix provides several actionable insights:

1. **Log-transformations** for skewed monetary variables (e.g., balances, limits) produce
   more symmetric distributions and clearer relationships.
2. Several predictors show **non-linear** relationships with the target, which may motivate
   transformations or the use of non-linear modelling components.
3. Some features exhibit **strong multicollinearity**, suggesting that variable selection,
   removal, or regularization (e.g., ridge/lasso) will be necessary.
4. Certain variables (e.g., `inq_last_6mths`) behave more like discrete categories than
   continuous linear predictors and may require appropriate encoding.

Overall, the scatterplot matrix confirms that, while linear regression remains viable, careful
feature engineering—particularly transformations and removal or consolidation of correlated
features—is needed to respect the assumptions of the linear model and prevent instabilit

### 3.2.3 Factor vs Numeric (Boxplots)

Here we loop over all combinations of factor × numeric.  
Boxplots show how the distribution of a numeric variable changes across categories.


```{r factor-numeric-boxplots, eval=FALSE}
# WARNING: This creates MANY plots due to many variables.
# You can set `eval = TRUE` if you want them all to actually render.  
# For reports, you might instead manually choose a subset of clinically or financially relevant variables.
for (f in fact_vars) {
  for (v in num_vars) {
    boxplot(cleaned_data_cc[[v]] ~ cleaned_data_cc[[f]],
            xlab = f,
            ylab = v,
            main = paste(v, "by", f),
            las = 2)
  }
}
```


To investigate how numeric variables behave across categories of key factors, we produced a set
of boxplots for each combination of:

- **Factors:** `term`, `emp_length`, `home_ownership`, `verification_status`, `purpose`,
  `addr_state`, `delinq_2yrs_cat`, `delinq_recency`, `acc_now_delinq_bin`,
  `major_derog_recency`
- **Numeric variables:** `loan_amnt`, `int_rate`, `annual_inc`, `dti`, `inq_last_6mths`,
  `open_acc`, `total_acc`, `log_loan_amnt`, `revol_bal_log`, `tot_cur_bal_log`,
  `total_rev_hi_lim_log`

All plots were generated **before any train/validation/test split** using the complete–case
dataset. This ensures that the visual patterns reflect the full data and are not affected by
downstream splitting or imputation.

Below we summarise the main findings factor by factor and discuss their modelling implications,
especially with respect to **linearity of the target `int_rate`** and potential **redundancy
or multicollinearity** among numeric predictors.

Term (36 vs 60 months)

- **`loan_amnt` by `term`:**  
  60-month loans have clearly higher median and upper quartile loan amounts than 36-month loans,
  with a wider spread and more high–value outliers. This confirms that longer terms are used for
  larger loans.
- **`int_rate` by `term`:**  
  60-month loans also show higher median interest rates and a wider interquartile range. The
  difference between the two terms is substantial and systematic, indicating that `term` is an
  important categorical predictor of `int_rate`.
- **Other numeric variables:**  
  Distributions of `annual_inc`, `dti`, `inq_last_6mths`, `open_acc`, `total_acc` and the
  log-transformed balance/limit variables are broadly similar between terms, with only minor
  shifts in medians.

**Implication:**  
`term` should be kept as a **factor** in the regression model, as it captures a clear
non-linear effect on `int_rate` that cannot be represented well by a simple linear trend.
The association between `term` and `loan_amnt` suggests some dependence but not critical
multicollinearity.

Employment Length (`emp_length`)

- **`int_rate` by `emp_length`:**  
  The medians and IQRs of `int_rate` are very similar across all employment-length categories
  (<1 year to 10+ years). There is no strong monotone pattern; if anything, differences are small
  and irregular.
- **Other numeric variables:**  
  `annual_inc`, `open_acc`, `total_acc`, and the log balance/limit variables tend to increase
  slightly with longer employment, but the changes are gradual and noisy.  
  `dti` and `inq_last_6mths` show very similar distributions across categories.

**Implication:**  
`emp_length` appears to have **weak predictive power** for `int_rate`. It can still be retained
for completeness or domain reasons, but we should not expect a strong linear or ordered effect.
Given the many categories, we should be cautious about over-parameterisation.

Home Ownership (`home_ownership`)

- **`int_rate` by `home_ownership`:**  
  Median interest rates differ only modestly between `RENT`, `MORTGAGE`, and `OWN`. The
  differences are small compared to within-group variability. Rare categories such as `NONE`,
  `OTHER` and `ANY` are based on few observations and show unstable boxplots.
- **`loan_amnt` and `annual_inc` by `home_ownership`:**  
  Borrowers with a mortgage or who own their home tend to have slightly higher incomes and loan
  amounts than renters, as expected.
- **Other numeric variables:**  
  The distributions of `dti`, `inq_last_6mths`, `open_acc`, `total_acc` and log balances are
  broadly similar across home-ownership types.

**Implication:**  
`home_ownership` may have **limited incremental effect** on `int_rate` once income and other
variables are included. 

Verification Status (`verification_status`)

- **`int_rate` by `verification_status`:**  
  The median interest rates for `Not Verified`, `Source Verified`, and `Verified` are close, with
  only small shifts between groups. The effect is weak and not clearly monotonic.
- **`annual_inc` and `loan_amnt`:**  
  Verified borrowers tend to have slightly higher incomes and loan amounts, but with substantial
  overlap between groups.
- **Other numeric variables:**  
  No strong separation is visible; distributions are largely similar.

**Implication:**  
`verification_status` appears to be a **weak categorical predictor** of `int_rate`. It can be
kept as a factor but is unlikely to explain a large fraction of the variance.

Loan Purpose (`purpose`)

- **`int_rate` by `purpose`:**  
  There are visible differences in median rates across purposes. Categories such as
  `small_business`, `educational`, and sometimes `renewable_energy` show higher typical
  interest rates, whereas `credit_card` and `debt_consolidation` are more central. Within each
  category, variability is still large.
- **`loan_amnt` and `log_loan_amnt`:**  
  Some purposes (e.g. `home_improvement`, `major_purchase`, `small_business`) have slightly
  higher loan amounts than others.
- **Other numeric variables:**  
  Differences in `dti`, `annual_inc`, and the balance/limit variables across purposes are present
  but not extreme.

**Implication:**  
`purpose` is a **meaningful categorical predictor** of `int_rate` and should be included as a
factor. Because it has many categories, there is a trade-off between interpretability and model
complexity; very rare purposes could be grouped into an "other" category if needed.

State of Residence (`addr_state`)

- **`int_rate` by `addr_state`:**  
  Median interest rates vary across states, but the differences are moderate and confidence for
  small states is low due to limited sample sizes.
- **Other numeric variables:**  
  Income, loan amount and account variables show some state-to-state variation but also large
  overlap.

**Implication:**  
`addr_state` has **very high cardinality** and would introduce a large number of dummy variables
in a linear regression model, with limited clear structure in the boxplots. To avoid
over-parameterisation and potential overfitting, it is reasonable either to exclude `addr_state`
from the regression model or to replace it by a coarser regional grouping if geographic effects
are of specific interest.

Delinquency History (`delinq_2yrs_cat`, `delinq_recency`)

- **`int_rate` by `delinq_2yrs_cat`:**  
  Borrowers with more past delinquencies (category `3+`) tend to have somewhat higher median
  interest rates than those with zero delinquencies, but the effect is modest and the higher
  categories are relatively rare.
- **`int_rate` by `delinq_recency`:**  
  More recent delinquencies are associated with higher typical interest rates; borrowers with
  `none` or very old delinquencies tend to have lower rates. The pattern is roughly monotone
  across the ordered recency categories.
- **Other numeric variables:**  
  Delinquency categories are also associated with slightly higher `dti`, higher `inq_last_6mths`,
  and somewhat different balance distributions, consistent with higher-risk profiles.

**Implication:**  
Delinquency features provide **risk-related information** and show a qualitatively sensible
relationship with `int_rate` (more and more recent delinquencies → higher rates). They are good
candidates to keep as **ordered factors** in the model. Because the higher-risk categories are
rare, their coefficients may have larger uncertainty.

Current Delinquency (`acc_now_delinq_bin`)

- **`int_rate` by `acc_now_delinq_bin`:**  
  The group with at least one currently delinquent account (`1`) tends to have slightly higher
  interest rates, but the group is very small compared to the non-delinquent (`0`) group.
- **Other numeric variables:**  
  The delinquent group also shows somewhat higher `dti` and different balance patterns, again in
  line with a higher-risk profile.

**Implication:**  
`acc_now_delinq_bin` behaves like a **rare high-risk flag**. It can be retained as a binary
indicator, but its effect size should be interpreted cautiously due to the small number of
positive cases.

Major Derogatory Events (`major_derog_recency`)

- **`int_rate` by `major_derog_recency`:**  
  Borrowers with more recent major derogatory events tend to have somewhat higher interest
  rates than those with no such events (`none`), although differences between non-empty
  categories are not very pronounced.
- **Other numeric variables:**  
  As with other delinquency-based factors, higher-risk categories often coincide with slightly
  higher `dti`, more inquiries, and higher outstanding balances.

**Implication:**  
`major_derog_recency` appears to capture additional credit-risk information and can be used as
an ordered factor. As before, rarer categories may lead to imprecise coefficient estimates.

**Summary and Impact on Linearity / Multicollinearity**

From the factor–numeric boxplots we conclude:

- Several factors (`term`, `purpose`, delinquency-related variables) show **clear, structured
  differences** in the distribution of `int_rate` across their levels. These are strong
  candidates to be included as categorical (often ordered) predictors in the linear regression
  model.
- Other factors (`emp_length`, `home_ownership`, `verification_status`) show only **weak
  differences** in `int_rate` and may contribute relatively little explanatory power once
  stronger predictors are included.
- The log-transformed monetary variables (`log_loan_amnt`, `revol_bal_log`, `tot_cur_bal_log`,
  `total_rev_hi_lim_log`) have broadly similar patterns across many factors, suggesting they
  measure overlapping aspects of credit exposure. This supports later checks for
  **multicollinearity** using VIFs and potentially dropping or combining highly redundant
  variables.
- High-cardinality factors such as `addr_state` increase model complexity substantially and are
  likely to be excluded or simplified to avoid overfitting.

Overall, the boxplots provide a qualitative understanding of how `int_rate` varies across
important borrower and loan characteristics and highlight where categorical effects and
potential multicollinearity need to be considered in the subsequent linear regression modelling.

## 3.3 Correlation Structure of Numeric Variables

We inspect the correlation matrix to:

- identify **strong linear relationships**,
- detect possible **multicollinearity risk** and **redundant variables**.


```{r corrplot-global, fig.height=7, fig.width=7}
cor_mat <- cor(cleaned_data_cc[num_vars], use = "pairwise.complete.obs")

corrplot(cor_mat,
         method = "color",
         type   = "lower",
         tl.cex = 0.7,
         diag   = FALSE)
```

Correlation Matrix Interpretation

The correlation matrix visualises pairwise Pearson correlations among all numeric predictors and 
the target variable `int_rate`. Pearson correlation requires numeric, continuous variables; 
categorical variables are therefore excluded.

The heatmap reveals several noteworthy linear relationships, both with the target variable and 
among the predictors.

Correlations With the Target `int_rate`

- **`annual_inc` (Income):**  
  Displays a **moderate negative correlation** with `int_rate`.  
  Borrowers with higher income tend to receive lower interest rates.  
  This relationship supports the earlier visual findings that income is a meaningful predictor, 
  though not strictly linear (log-transformation improves linearity).

- **`dti` (Debt-to-Income Ratio):**  
  Shows a **small positive correlation** with `int_rate`.  
  Higher DTI values are associated with slightly higher interest rates, as expected for higher-risk 
  borrowers.

- **`inq_last_6mths`:**  
  Weak positive correlation. Borrowers with more recent credit inquiries tend to have marginally 
  higher interest rates. The effect size is small because this variable is discrete and skewed.

- **`log_loan_amnt`:**  
  Shows a mild positive correlation with `int_rate`, suggesting larger loans are associated with 
  slightly higher interest rates. The log transformation stabilises the skew and gives a more 
  interpretable relationship.

- **Credit utilisation / credit exposure variables:**  
  `revol_bal_log`, `tot_cur_bal_log`, and `total_rev_hi_lim_log` all have **weak-to-moderate positive 
  correlations** with `int_rate`.  
  Borrowers with higher balances and higher limits tend to pay somewhat higher interest rates.  
  The associations are monotonic but not very strong, which is typical in real credit datasets where 
  many risk factors influence pricing.

- **Weak/non-informative correlations:**  
  `open_acc` and `total_acc` show near-zero correlation with the target, suggesting little linear 
  influence on interest rate.

**Summary:**  
`int_rate` correlates meaningfully (but moderately) with income, DTI, recent inquiries, and the 
various log-transformed exposure variables. These patterns confirm that the variables carry relevant 
risk information but no single numeric predictor on its own explains a large portion of the variance.


Correlations Among the Predictors

The matrix shows several strong internal relationships among numeric predictors:

- **`open_acc` and `total_acc`:**  
  These two variables are **highly correlated**, forming a distinct bright-blue block in the matrix.  
  This is expected because total accounts naturally include open accounts. Including both in the 
  linear model may induce multicollinearity and inflate VIF values.

- **Credit exposure variables:**  
  `revol_bal_log`, `tot_cur_bal_log`, and `total_rev_hi_lim_log` form the largest and clearest 
  high-correlation cluster. All three measure related aspects of revolving utilisation and overall 
  credit exposure.  
  Their strong pairwise correlations indicate redundancy.  
  Including all of them in an OLS model will almost certainly cause **multicollinearity**.  
  Variable selection or dimensionality reduction may be appropriate.

- **`log_loan_amnt`:**  
  Moderately correlated with the credit exposure block.  
  This suggests larger loans tend to coexist with larger balances and higher credit limits.  
  The correlations are not extreme but should be monitored with VIF.

- **`annual_inc`:**  
  Shows mild positive correlation with exposure variables and loan amount, capturing socioeconomic 
  structure. These correlations are not problematic.

**Summary:**  
The correlation matrix highlights two main multicollinearity sources:  
(1) `open_acc` vs. `total_acc`, and  
(2) the block of log-transformed balance/limit variables.  
These will require attention during VIF analysis and feature selection.


```{r}
library(fastDummies)
set.seed(123)
factors_of_interest <- c("term", 
                         "emp_length", 
                        "home_ownership",
                         "verification_status", 
                         "delinq_2yrs_cat", 
                         "delinq_recency", 
                         "acc_now_delinq_bin", 
                         "major_derog_recency"
                        )

# Keep only variables that exist in factors of interest
df_no_addr_state <- cleaned_data_cc[, c(num_vars, factors_of_interest)]

sample_idx <- sample(nrow(df_no_addr_state), 10000)
df_sample <- df_no_addr_state[sample_idx, ]

# Dummy encode ONLY the selected categorical variables
df_dummy <- dummy_cols(df_sample,
                       remove_first_dummy = TRUE,
                       remove_selected_columns = TRUE)

# Keep numeric columns only
numeric_all <- df_dummy[, sapply(df_dummy, is.numeric)]

# Compute correlation
cor_mat <- cor(numeric_all, use = "pairwise.complete.obs")

# Plot correlation matrix
corrplot(cor_mat,
         method = "color",
         type   = "lower",
         tl.cex = 0.7,   # readable now
         diag   = FALSE)
```
```{r}
# Get correlations of all variables with int_rate
cor_with_int <- cor_mat["int_rate", ]
sort(abs(cor_with_int), decreasing = TRUE)
cor_with_int

```





---

# 4. Train / Validation / Test Split (70/15/15)

Now we split the **complete-case data** into train, validation and test sets with:

- seed `123` for reproducibility,
- **70% train, 15% validation, 15% test**.

We do this split **before** any model fitting in order to:

- prevent information leakage from test into training,
- reserve test data for final unbiased performance assessment.


```{r split-data}
set.seed(123)

n <- nrow(cleaned_data_cc)
indices <- seq_len(n)

# 70% train
train_size <- floor(0.7 * n)
train_index <- sample(indices, size = train_size)

remaining <- setdiff(indices, train_index)

# Split remaining into 15% val, 15% test (roughly equal halves)
val_size <- floor(0.5 * length(remaining))
val_index  <- sample(remaining, size = val_size)
test_index <- setdiff(remaining, val_index)

train <- cleaned_data_cc[train_index, ]
val   <- cleaned_data_cc[val_index, ]
test  <- cleaned_data_cc[test_index, ]

cat("Train rows:", nrow(train), "\n")
cat("Validation rows:", nrow(val), "\n")
cat("Test rows:", nrow(test), "\n")
```


Note: In this document we **do not perform imputation**, because that will be done in a separate workflow. For the diagnostics here, we already restricted to complete cases via `na.omit()`.

---

# 5. EDA on Training Set (Post-Split)

We now repeat **scatterplots and correlation analysis on the training set only**, as this is the data on which the model will be trained and diagnostics evaluated.

## 5.1 Scatterplot Matrix on Training Set


```{r train-scatter-matrix, fig.height=8, fig.width=8}
train_num_vars <- names(train)[sapply(train, is.numeric)]

pairs(train[train_num_vars],
      main = "Scatterplot Matrix of Numeric Variables (Train)")
```


## 5.2 Correlation Matrix on Training Set


```{r train-corrplot, fig.height=7, fig.width=7}
train_cor <- cor(train[train_num_vars], use = "pairwise.complete.obs")

corrplot(train_cor,
         method = "color",
         type   = "lower",
         tl.cex = 0.7,
         diag   = FALSE)
```


This provides a more accurate picture of the correlations relevant for the linear regression model that will be fitted on the training data.

---

# 6. Linear Regression Model on Training Set

We now fit a **multiple linear regression model** using the training data:

- Response: `int_rate` (assumed to be present in `train`).
- Predictors: all remaining variables (`.`).


```{r fit-lm}
# Ensure int_rate is numeric
stopifnot("int_rate" %in% names(train))
stopifnot(is.numeric(train$int_rate))

lm_fit <- lm(int_rate ~ . - addr_state - earliest_cr_line_date - last_credit_pull_d_date,
             data = train)
summary(lm_fit)
```
We fitted a linear regression model with the pre-selected columns. Based on the correlation plot and the fact that we have a high number of features, even more increased due to OneHotEncoding, we know that the goodness of fit of the model will be bad. The summary of the model shows an adjusted R-squared of 0.4755, which is worse than a random classifier.

Now let's kick out some of the highly colinear variables. We can identify these variables using previous knowledge and aiding our decision by calculating the Variance Inflation Factor (VIF).

```{r}
# Calculate VIF
vif_values <- vif(lm_fit)
vif_values
```
VIF of more than 5 indicates a potential multicollinearity problem. We see this for delinq_recency. We exclude this column and retrain and look at vif.
```{r}
stopifnot("int_rate" %in% names(train))
stopifnot(is.numeric(train$int_rate))

lm_fit <- lm(int_rate ~ . - addr_state - earliest_cr_line_date - last_credit_pull_d_date - delinq_recency,
             data = train)
summary(lm_fit)
```

```{r}
vif_values <- vif(lm_fit)
vif_values
```
The adjusted R2 is similar. Furthermore we see that loan_amnt and log_loan_amnt are highly colinear, which is logical and already seen in the correlation matrix. We make two linear_regression models, one with loan_amnt and one with log_loan_amnt and compare the two models

```{r}
stopifnot("int_rate" %in% names(train))
stopifnot(is.numeric(train$int_rate))

lm_fit <- lm(int_rate ~ . - addr_state - earliest_cr_line_date - last_credit_pull_d_date - delinq_recency - loan_amnt,
             data = train)
summary(lm_fit)

vif_values <- vif(lm_fit)
vif_values
```

```{r}
stopifnot("int_rate" %in% names(train))
stopifnot(is.numeric(train$int_rate))

lm_fit <- lm(int_rate ~ . - addr_state - earliest_cr_line_date - last_credit_pull_d_date - delinq_recency - log_loan_amnt,
             data = train)
summary(lm_fit)

vif_values <- vif(lm_fit)
vif_values
```
The R2 between the two models are very similar. Therefore we will stick with the original, non-log-transformed, column 'loan_amnt'.
Furtermore we see that home_ownership does not have significant p-values. Since it also had a weak correlation in the correlation matrix, we will exclude this column as well.
```{r}
stopifnot("int_rate" %in% names(train))
stopifnot(is.numeric(train$int_rate))

lm_fit <- lm(int_rate ~ . - addr_state - earliest_cr_line_date - last_credit_pull_d_date - delinq_recency - log_loan_amnt - home_ownership,
             data = train)
summary(lm_fit)

vif_values <- vif(lm_fit)
vif_values
```
We still don't see a lot of difference in the adjusted R2. We therefore drop the columns which didn't show meaningful predictive value in the histograms and scatterplots. These columns are open_acc, total_acc, emp_length, home_ownership and verification_status.
```{r}
exclude_vars <- c(
  "addr_state",
  "earliest_cr_line_date",
  "last_credit_pull_d_date",
  "delinq_recency",
  "log_loan_amnt",
  "home_ownership",
  "emp_length",
  "open_acc",
  "total_acc",
  "major_derog_recency",
  "verification_status"
)


all_vars <- setdiff(names(train), exclude_vars)

formula <- reformulate(all_vars[all_vars != "int_rate"], response = "int_rate")

lm_fit <- lm(formula, data = train)
summary(lm_fit)

vif(lm_fit)

```

Let's try a lasso regression as well to see if we can improve the model with all variables included.
```{r}
# Remove variables you decided to exclude
exclude_vars <- c("addr_state", "earliest_cr_line_date", "last_credit_pull_d_date")

train_lasso <- train %>%
  select(-all_of(exclude_vars))

# Build model matrix (glmnet handles dummy-encoding automatically)
X <- model.matrix(int_rate ~ ., data = train_lasso)[, -1]  # remove intercept column
y <- train_lasso$int_rate

set.seed(123)
cv_lasso <- cv.glmnet(
  X, y,
  alpha = 1,          # LASSO
  family = "gaussian",
  nfolds = 10,
  standardize = TRUE
)

best_lambda <- cv_lasso$lambda.min
best_lambda

lasso_model <- glmnet(X, y, alpha = 1, lambda = best_lambda)

y_pred <- predict(lasso_model, X)

# Compute R2
R2_lasso <- 1 - sum((y - y_pred)^2) / sum((y - mean(y))^2)
R2_lasso

```

```{r}
coef_lasso <- coef(lasso_model)
active_vars <- rownames(coef_lasso)[which(coef_lasso != 0)]
active_vars

```

```{r}
coef_df <- data.frame(
  variable = rownames(coef_lasso),
  coefficient = as.numeric(coef_lasso)
) %>%
  filter(variable != "(Intercept)") %>%
  mutate(abs_coef = abs(coefficient)) %>%
  arrange(desc(abs_coef))

head(coef_df, 20)

```

## 4. Why Linear Regression Performs Poorly

In this section we investigate why the multiple linear regression models for the
interest rate (`int_rate`) achieve a relatively low explanatory power
(adjusted R² ≈ 0.47).  
The main hypothesis is that **the relationship between the features and
`int_rate` is not linear**, so a linear model can only capture part of the
structure.

To support this, we:

1. Fit a linear regression model and quantify its performance.  
2. Use diagnostic plots to check whether the residuals behave as expected
   under a linear relationship.  
3. Visualise the relationship between key predictors and `int_rate` using
   LOESS curves on a random subsample of the data.  
4. Compare the linear model to a more flexible, nonlinear model (a GAM) on a
   subsample to show that allowing nonlinearity clearly improves fit.

Because the training set has about 500,000 rows, I use random subsamples
for plotting and for the nonlinear comparison to keep computation and
graphics manageable.

```{r}
library(dplyr)
library(ggplot2)
library(broom)
library(mgcv)   # for a flexible nonlinear (GAM) model

set.seed(123)
```
4.1. Baseline multiple linear regression
We start with a reasonably rich linear regression model using the main
continuous and categorical predictors that are available after cleaning.

```{r}
lm_fit <- lm(
int_rate ~ loan_amnt + term + annual_inc + purpose +
dti + inq_last_6mths +
delinq_2yrs_cat + acc_now_delinq_bin +
revol_bal_log + tot_cur_bal_log + total_rev_hi_lim_log,
data = train
)

lm_summary <- summary(lm_fit)
lm_summary$adj.r.squared
```

The adjusted R² of this model is:

- **Adjusted R²:** 0.44

This means that the linear model explains only about
44 % of the variability in
`int_rate`. Given that we use many relevant credit-risk predictors,
this relatively low value suggests that the problem is not simply
missing variables, but that the functional form (strict linearity)
is too restrictive.

4.2. Residual diagnostics: remaining structure after the linear fit
If the linear model were appropriate, the residuals (observed minus fitted
values) should look like almost random noise with no structure. Systematic
patterns in the residuals indicate that the model misses important
nonlinearities or interactions.

Because plotting 500,000 points is not helpful, I use a random subsample
of 20,000 observations for the diagnostics.
```{r}
plot_sample <- train %>%
sample_n(20000)

plot_sample$lm_fitted   <- predict(lm_fit, newdata = plot_sample)
plot_sample$lm_residual <- plot_sample$int_rate - plot_sample$lm_fitted

ggplot(plot_sample, aes(x = lm_fitted, y = lm_residual)) +
geom_point(alpha = 0.2) +
geom_hline(yintercept = 0, colour = "red") +
labs(
title = "Residuals vs Fitted (linear model)",
x = "Fitted interest rate",
y = "Residuals"
)
```
In a well-specified linear regression, the residuals should form a homogeneous, structureless cloud around zero, with constant variance and no discernible patterns.
However, the residual–versus–fitted plot shows clear systematic structure. The spread of the residuals changes across the range of fitted values, forming a distinct “diamond” shape: the variance is smallest near the center and increases toward the extremes. Additionally, the residuals appear in curved, slanted bands rather than being randomly dispersed. These patterns indicate heteroscedasticity and nonlinearity—strong evidence that the linear model fails to capture the true relationship between the predictors and int_rate.
Such visible structure contradicts the assumptions of linear regression (linearity, homoscedasticity, independent residuals), and confirms that the underlying functional form is not linear. This diagnostic plot therefore supports the hypothesis that a linear model is not appropriate for modelling LendingClub interest rates.

4.3 Nonlinear patterns in key predictors
To directly test the hypothesis that `int_rate` is not linearly related to
the predictors, I inspect the shape of the relationship for several
important continuous variables:

- loan amount (`loan_amnt`)
- debt-to-income ratio (`dti`)
- revolving balance (`revol_bal_log`)
- total revolving credit limit (`total_rev_hi_lim_log`)

For each variable, I plot `int_rate` against the predictor on a subsample
of 20,000 observations and overlay two smooths:

- a **LOESS curve** (flexible, capturing nonlinear trends)  
- a **straight regression line** (the linear assumption)

If the LOESS curve deviates strongly from the straight line, this
indicates that the true relationship is nonlinear.

```{r}
cont_sample <- train %>%
select(int_rate, loan_amnt, dti, revol_bal_log, total_rev_hi_lim_log) %>%
sample_n(20000)

plot_smooth <- function(data, xvar, xlab) {
ggplot(data, aes_string(x = xvar, y = "int_rate")) +
geom_point(alpha = 0.05) +
geom_smooth(method = "loess", se = FALSE) +
geom_smooth(method = "lm",   se = FALSE, linetype = "dashed", colour = "red") +
labs(
x = xlab,
y = "Interest rate",
title = paste("LOESS (blue) vs linear fit (red dashed) for", xlab)
)
}

p1 <- plot_smooth(cont_sample, "loan_amnt", "Loan amount")
p2 <- plot_smooth(cont_sample, "dti", "Debt-to-income ratio")
p3 <- plot_smooth(cont_sample, "revol_bal_log", "Log revolving balance")
p4 <- plot_smooth(cont_sample, "total_rev_hi_lim_log", "Log total revolving limit")

p1
p2
p3
p4
```
Across all four plots, the blue LOESS curves show clear nonlinear patterns that the straight red linear fits fail to capture. This provides strong evidence that the relationship between each predictor and int_rate is not linear.
- Loan amount: The LOESS curve bends downward for small loans, flattens in the mid-range, and bends upward again for larger loans. This U-shaped pattern cannot be represented with a straight line, confirming nonlinearity.
- Debt-to-income ratio (DTI): The relationship is not a constant slope: the curve is almost flat at low DTI values, increases moderately in the middle range, and becomes much steeper as DTI approaches higher values. This increasing slope contradicts the assumption of a linear effect.
- Log revolving balance: The LOESS curve rises slightly at first and then decreases for higher values, creating an inverted-U shape. The linear model completely misses this turning point.
- Log total revolving limit: The relationship is strongly curved: the LOESS curve first increases toward a maximum and then declines sharply, whereas the linear fit implies a strictly decreasing line. These opposite trends highlight a serious mismatch between the linear assumption and the data.

These departures from linearity mean that a straight-line fit is structurally incapable of capturing the true shapes seen in the data. As a result, the linear model systematically misrepresents the underlying relationships, which explains the patterned residuals and the relatively low R². Because the true functional forms are nonlinear, the poor performance of the linear model is expected and theoretically justified.

4.4 Comparison with a more flexible nonlinear model
To further support the hypothesis that the low R² is caused by
nonlinearity rather than by noise or missing variables, we fit a more
flexible model on a subsample: a Generalized Additive Model (GAM).  
The GAM allows smooth nonlinear functions of the continuous predictors
while keeping the same set of variables.

I compare its performance to the linear model on the **same subsample**
of 50,000 rows.
```{r}
set.seed(123)
sub <- train %>%
sample_n(50000)

# Linear model on the subsample

lm_sub <- lm(
int_rate ~ loan_amnt + term + annual_inc + purpose +
dti + inq_last_6mths +
delinq_2yrs_cat + acc_now_delinq_bin +
revol_bal_log + tot_cur_bal_log + total_rev_hi_lim_log,
data = sub
)

# Flexible GAM on the same predictors

gam_sub <- gam(
int_rate ~ s(loan_amnt) + s(dti) + s(revol_bal_log) +
s(tot_cur_bal_log) + s(total_rev_hi_lim_log) +
term + purpose + inq_last_6mths +
delinq_2yrs_cat + acc_now_delinq_bin +
annual_inc,   # can be kept linear here
data = sub
)

# R2 for both models on the subsample

lm_pred  <- predict(lm_sub,  newdata = sub)
gam_pred <- predict(gam_sub, newdata = sub)

r2 <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

r2_lm_sub  <- r2(sub$int_rate, lm_pred)
r2_gam_sub <- r2(sub$int_rate, gam_pred)

c(linear_R2 = r2_lm_sub, gam_R2 = r2_gam_sub)
```
On the same 50,000-row subset, the models achieve the following performance:
Linear model R²: 0.44
GAM (nonlinear) R²: 0.48

The GAM, which allows nonlinear smooths for the continuous predictors, achieves a noticeably higher R² than the linear model. While the improvement is moderate rather than dramatic, it is consistent and systematic. This indicates that part of the unexplained variance in the linear model is indeed due to nonlinear relationships that the linear specification cannot capture.
The fact that the GAM performs better, even using the same set of predictors, provides empirical evidence that the relationship between features and int_rate is not purely linear. At the same time, the relatively small increase in R² suggests that interest-rate determination is influenced by additional factors such as policy rules, credit-grade thresholds, and noise, which neither model can capture fully.

4.5. Interpretation
- The multiple linear regression explains only about r round(lm_summary$adj.r.squared * 100)% of the variation in interest rates, despite using many relevant credit-risk features. This relatively low explanatory power suggests that the assumptions of linearity and additivity may not hold.
- Residual diagnostics show clear, non-random patterns, including heteroscedasticity and structured bands, which indicate that the linear model does not adequately capture the underlying relationships in the data.
- LOESS visualisations of int_rate against key continuous predictors show curved, often non-monotonic relationships, while the fitted linear trend is restricted to a straight line. These plots provide direct evidence that the true effects are nonlinear.
- A more flexible GAM using the same predictors on a subsample achieves a moderately higher R² than the linear model. Although the improvement is not dramatic, it is consistent and systematic, demonstrating that relaxing the linearity assumption leads to a better representation of the underlying structure.

Taken together, these findings support the hypothesis that there is no simple
linear relationship between int_rate and the available predictors. Linear
regression therefore serves as a reasonable baseline, and its relatively low R²
is expected and interpretable given the nonlinear and policy-driven nature of
LendingClub’s interest-rate assignment.

For improved predictive performance, models capable of capturing nonlinearities—such
as GAMs, regularised linear models (LASSO/Elastic Net), or tree-based ensemble
methods—are more appropriate for this type of data.




------------------------------------------------------------




```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(corrplot)
library(car)
library(Hmisc)

# 1. Load Data
input_file_path <- "/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_data/Intermediate/cleaned_data.csv" # Use relative path if possible
cleaned_data <- read.csv(input_file_path)

# 2. Complete Cases (Strict requirement for OLS diagnostics)
cleaned_data_cc <- na.omit(cleaned_data)

# 3. SPLIT DATA FIRST (70/15/15)
set.seed(123)
# Create indices
n <- nrow(cleaned_data_cc)
train_idx <- sample(seq_len(n), size = floor(0.7 * n))
remaining_idx <- setdiff(seq_len(n), train_idx)
val_idx <- sample(remaining_idx, size = floor(0.15 * n))
test_idx <- setdiff(remaining_idx, val_idx)

# Assign subsets
train_set <- cleaned_data_cc[train_idx, ]
val_set   <- cleaned_data_cc[val_idx, ]
test_set  <- cleaned_data_cc[test_idx, ]

cat("Train size:", nrow(train_set), "\n")
```
### **2. Feature Engineering (The "Fuel" & "Engine" Fix)**

This block applies the transformations we tested. We apply them to `train_set` for modeling, but you will need to apply the exact same logic to `val_set` and `test_set` when you predict later.


```{r feature-engineering}
# Function to apply transformations consistently
transform_features <- function(df) {
  df %>%
    mutate(
      # 1. TARGET: Log-transform interest rate to fix "diamond" residuals
      log_int_rate = log(int_rate),
      
      # 2. INCOME: Log-transform to linearize the curve
      log_annual_inc = log(annual_inc + 1),
      
      # 3. DTI: Add squared term for curvature (poly(dti, 2))
      # We create it manually to keep it in the dataframe
      dti_sq = dti^2,
      
      # 4. Ensure factors are factors
      term = as.factor(term),
      purpose = as.factor(purpose),
      home_ownership = as.factor(home_ownership),
      delinq_2yrs_cat = as.factor(delinq_2yrs_cat)
    )
}


# Apply to Train (and Validation for later)
train_transformed <- transform_features(train_set)
val_transformed   <- transform_features(val_set)
```



--

### **3. EDA on Training Data Only**

Now you can run your scatterplots and correlations without fear of leakage.


```{r train-eda}
# Correlation on transformed numeric variables
numeric_vars <- train_transformed %>% 
  select(log_int_rate, log_annual_inc, dti, dti_sq, loan_amnt, 
         revol_bal_log, tot_cur_bal_log)

cor_mat <- cor(numeric_vars, use = "pairwise.complete.obs")
corrplot(cor_mat, method = "color", type = "lower", diag = FALSE, title = "Correlations (Train)")

# Check the new relationship: Log Income vs Log Int Rate
# (This should look much straighter than your original curved plot)
train_transformed %>%
  sample_n(10000) %>%
  ggplot(aes(x = log_annual_inc, y = log_int_rate)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Linearity Check: Log Income vs Log Int Rate")
```
---

### **4. Fitting the Full Model**

Here we combine the transformed variables with the categorical ones (`term`, `purpose`) to maximize predictive power.


```{r fit-full-model}
# Define the formula with our transformed variables
# Note: We use log_int_rate as the target
full_formula <- log_int_rate ~ 
                log_annual_inc + 
                dti + dti_sq + 
                loan_amnt + 
                term +                  
                purpose + 
                inq_last_6mths + 
                delinq_2yrs_cat + 
                revol_bal_log + 
                tot_cur_bal_log + 
                home_ownership

# Fit model
lm_final <- lm(full_formula, data = train_transformed)

summary(lm_final)
```
---

### **5. Diagnostics **

This generates the plots to confirm the "diamond" shape is improved and normality is better.


```{r model-diagnostics}
par(mfrow = c(2, 2)) # 2x2 grid for plots

# 1. Residuals vs Fitted (Check for Homoscedasticity)
plot(lm_final, which = 1) 

# 2. Normal Q-Q (Check for Normality)
plot(lm_final, which = 2)

# 3. Scale-Location
plot(lm_final, which = 3)

# 4. Residuals vs Leverage
plot(lm_final, which = 5)

# VIF Check
vif(lm_final)
```
```{r}
# ---------------------------------------------------------
# CALCULATION OF MSE & RMSE ON ORIGINAL SCALE
# ---------------------------------------------------------

# 1. Choose your dataset (Train, Validation, or Test)
# Make sure this dataset has the same transformations applied!
target_df <- train_transformed  # or val_transformed / test_transformed

# 2. Make predictions (These will be in Log units)
log_predictions <- predict(lm_final, newdata = target_df)

# 3. Convert predictions back to original Interest Rate %
# We reverse the log() using exp()
predictions_original <- exp(log_predictions)

# 4. Get the actual original interest rates
actuals <- target_df$int_rate

# 5. Calculate Metrics
mse_val <- mean((actuals - predictions_original)^2)
rmse_val <- sqrt(mse_val)

# 6. Print Results
cat("------------------------------------------\n")
cat("Results on Original Scale (Interest Rate %)\n")
cat("------------------------------------------\n")
cat("MSE: ", round(mse_val, 4), "\n")
cat("RMSE:", round(rmse_val, 4), "\n")
cat("------------------------------------------\n")
cat("Interpretation:\n")
cat("On average, the model's predictions are off by +/-", round(rmse_val, 2), "percentage points.\n")
```





