# Data Preprocessing

In this exploratory data analysis we used following libraries:

```{r}
library(tidyverse)
library(corrplot)
library(fastDummies)
library(dplyr)
library(ggplot2)
library(patchwork)
```

The data is loaded into a dataframe and predefined columns are dropped. With a first look at the data using the data dictionary, as well as looking at the first few rows, a dictionary for the data types is created. The data is casted into the correct data types. and a copy of the original dataframe is created for backup.

```{r}
original <- read.csv("~/Documents/Repos/GRP-6_DS-Project/DS-Project_data/LCdata.csv", 
                 row.names = NULL, 
                 sep = ";",
                 header = TRUE)

# Predefined columns to drop
columns_to_drop <- c("collection_recovery_fee",
                     "installment",
                     "funded_amnt",
                     "funded_amnt_inv",
                     "issue_d",
                     "last_pymnt_amnt",
                     "last_pymnt_d",
                     "loan_status",
                     "next_pymnt_d",
                     "out_prncp",
                     "out_prncp_inv",
                     "pymnt_plan",
                     "recoveries",
                     "total_pymnt",
                     "total_pymnt_inv",
                     "total_rec_int",
                     "total_rec_late_fee",
                     "total_rec_prncp")

original_drop <- original[, -match(columns_to_drop,  names(original))]

# Dtype dictionary
dtype_dict <- list(
  acc_now_delinq = "integer",
  addr_state = "factor",
  all_util = "numeric",
  annual_inc = "numeric",
  annual_inc_joint = "numeric",
  application_type = "factor",
  collections_12_mths_ex_med = "integer",
  delinq_2yrs = "integer",
  desc = "factor",
  dti = "numeric",
  dti_joint = "numeric",
  earliest_cr_line = "str", # special case, will be converted to date later
  emp_length = "factor",
  emp_title = "factor",
  home_ownership = "factor",
  id = "integer",
  il_util = "numeric",
  initial_list_status = "factor",
  inq_fi = "integer",
  inq_last_12m = "integer",
  inq_last_6mths = "integer",
  int_rate = "numeric",
  last_credit_pull_d = "str", # special case, will be converted to date later
  loan_amnt = "integer",
  max_bal_bc = "numeric",
  member_id = "integer",
  mths_since_last_delinq = "integer",
  mths_since_last_major_derog = "integer",
  mths_since_last_record = "integer",
  mths_since_rcnt_il = "integer",
  open_acc = "integer",
  open_acc_6m = "integer",
  open_il_12m = "integer",
  open_il_24m = "integer",
  open_il_6m = "integer",
  open_rv_12m = "integer",
  open_rv_24m = "integer",
  policy_code = "factor",
  pub_rec = "integer",
  purpose = "factor",
  revol_bal = "integer",
  revol_util = "numeric",
  term = "factor",
  title = "factor",
  tot_coll_amt = "numeric",
  tot_cur_bal = "numeric",
  total_acc = "integer",
  total_bal_il = "numeric",
  total_cu_tl = "integer",
  total_rev_hi_lim = "numeric",
  url = "factor",
  zip_code = "factor"
)

# Cast dtypes according to dictionary
for(col_name in names(dtype_dict)) {
  dtype <- dtype_dict[[col_name]]
  if(dtype == "factor") {
    original_drop[[col_name]] <- as.factor(original_drop[[col_name]])
  } else if(dtype == "integer") {
    original_drop[[col_name]] <- as.integer(original_drop[[col_name]])
  } else if(dtype == "numeric") {
    original_drop[[col_name]] <- as.numeric(original_drop[[col_name]])
  } else if(dtype == "date") {
    original_drop[[col_name]] <- as.Date(original_drop[[col_name]], format="%b-%Y")
  }
}

str(original_drop)

# Make Copy to work with

data <- original_drop
```

## Data Understanding

### Dimensions

For a first understanding of the first few rows of the dataframe, we use the head() function. We check the dimensions of the dataframe using the dim() function, and we use the str() function to get an overview of the data types of each column. Furthermore we look at the summary statistics of the dataset. One could discuss whether one should look at the statistics before a train-test-split or after, since this leads to certain bias. In this case we decided to look at the statistics before the split, as learned in class.

```{r}
# Display the first few rows of the dataset
print("First few rows of the dataset:")
print(head(data))

# Get the dimensions of the dataset
dim(data)
print(paste("The number of rows is:", nrow(data))) # 798641
print(paste("The number of columns is:", ncol(data))) # 54

# Get the structure of the dataset
str(data)
```

```{r}
# Summary statistics of the dataset
summary(data)
```

General observations:
- Many numeric columns show very large ranges, some with maximum values far beyond plausible real-world limits.
- Several variables contain placeholder values such as 999, 9999, or -1 indicating missing or unknown information.
- A number of columns show repeated NA counts, such as NA's: 25 appearing across unrelated features, suggesting systematic missingness.
- Some categorical columns include unusual entries like (Other) or long HTML-like strings.

Loan amount variables (loan_amnt, funded_amnt, funded_amnt_inv) show values ranging from $500 to tens of millions, which is unrealistic for consumer lending and likely erroneous. Such extreme values need investigation and removal. Interest rate values are within expected ranges.
Employment fields contain high-cardinality free-text entries and placeholders. Home ownership includes standard entries but also rare placeholders such as ANY or NONE. Annual income includes extreme values exceeding several million dollars, which are likely incorrect and should be capped or treated as NA.
Zip codes and state entries contain unusual placeholders such as Other. Debt-to-income ratio includes 999 or 999.99 as clear placeholders and should be converted to NA. Delinquency-related variables contain many missing values and some extreme maximums that may be unrealistic.
Revolving utilization contains impossible values greater than 100 percent, indicating data entry errors. Other credit variables show suspiciously high counts or balances that do not fit typical borrower profiles.
Common placeholder patterns detected include 999, 999.0, 999.99, 9999, or textual placeholders like Other. These should be replaced with NA during cleaning.
Repeated NA counts (e.g., NA's: 25) across certain variables suggest missing information originating from the same borrower rows. These relationships should be examined.
Columns containing URLs or HTML-like descriptions contribute no analytical value and should be removed unless needed for text analysis.
Issues to fix include extreme outliers, placeholder values requiring conversion to NA, text cleanup for free-text fields, consistent NA handling across affected rows, and parsing of date fields. Irrelevant columns such as url or verbose descriptions can be dropped. After cleaning, a new summary should be generated to confirm improvements.

```{r}
# vector of placeholder values to replace
placeholders <- c(999, 9999.00, 9999999)

data <- data %>%
  mutate(across(where(is.numeric),
                ~ ifelse(. %in% placeholders, NA, .)))
```



### Numerical and Categorical Data

To differentiate between numerical and categorical data, we can use the sapply() function to check the class of each column in the dataframe. Numerical columns are typically of class "integer" or "numeric", while categorical columns are of class "factor" or "character". We can then create separate lists for numerical and categorical columns.

```{r}
# Identify numerical and categorical columns
numerical_cols <- names(data)[sapply(data, function(col) is.numeric(col) ||
                                            is.integer(col))]
categorical_cols <- names(data)[sapply(data, function(col) is.factor(col) ||
                                              is.character(col))]
print("Numerical Columns:")
print(numerical_cols)
print("Categorical Columns:")
print(categorical_cols)
```

## Data Wrangling

Data wrangling is the process of cleaning and transforming raw data into a format that is suitable for analysis. This includes handling missing values, correcting duplicate entries, and addressing any incorrect values in the dataset. The first chapters of this project are per definition also part of data wrangling, but were seperated for better clarity in the code. Cleaning and transforming the data is an iterative process that requires careful consideration of the specific characteristics of the dataset and the goals of the analysis. The better the cleaning and transforming is done, the more reliable and accurate the results of the analysis will be.

### Duplicated Values

Duplicated values can be entire rows which are duplicated or, in this case, duplication of the id and/or member_id columns. If these rows can be identified, the duplicated row should be dropped. If multiple duplicated rows are kept in the dataset, the model would learn to overfit on these rows, leading to a worse performance on unseen data.

```{r}
# Check for duplicate rows
duplicate_rows <- data[duplicated(data), ]
print(paste("Number of duplicate rows:", nrow(duplicate_rows)))

# Check for duplicate IDs
duplicate_ids <- data[duplicated(data$id)]
print(paste("Number of duplicate IDs:", sum(duplicate_ids)))

# Check for duplicated member IDs
duplicate_member_ids <- data[duplicated(data$member_id)]
print(paste("Number of duplicate member IDs:", sum(duplicate_member_ids)))
```

There are no duplicated rows to take care of.

## Missing Values

In the summary, we can see that there are several columns with a high number of missing values. Depending on the percentage of missing values, we can decide to either drop the column or impute the missing values. For columns with a low percentage of missing values, we can use techniques such as mean/median imputation for numerical columns or mode imputation for categorical columns. For columns with a high percentage of missing values, it may be better to drop the column altogether, as imputing too many missing values can introduce bias into the dataset.

```{r}
# Calculate the amount and percentage of missing values for each column
missing_values <- data.frame(
  column = names(data),
  missing_count = sapply(data, function(col) sum(is.na(col))),
  missing_percentage = round(sapply(data, function(col) mean(is.na(col)) * 100), 2)
)

print("Missing values per column:")
print(missing_values)
```
Especially for high columns with high NA-counts, one should contact the data manager to understand the logic behind the missing values. It is possible that there was a mistake in the SQL-output or that there is a specific reason why these values are missing. Since this is not possible in this project, we have to make assumptions based on the data dictionary and the summary statistics. NA-values can therefore either be imputed using different imputation methods or dropped.

We have to be careful to drop all rows or columns with NA-values, since sometimes there is a logic behind a missing value. For example, in the column "mths_since_last_delinq", a missing value could mean that the person has never been delinquent. Dropping these rows would lead to a loss of valuable information. Therefore, we have to carefully analyze each column with missing values and decide on a case-by-case basis how to handle them.

- delinq_2yrs: There are 25 missing values, which is a very small percentage of the total data. These values could be imputed or dropped after further anaylsis.
- inq_last_6mths: There are 25 missing values. These values could be imputed or dropped after further anaylsis. 
- open_acc: There are 25 missing values. These values could be imputed or dropped after further anaylsis.
- pub_rec: There are 25 missing values. These values could be imputed or dropped after further anaylsis.
- total_acc: There are 25 missing values. These values could be imputed or dropped after further anaylsis.
- acc_now_delinq: There are 25 missing values. These values could be imputed or dropped after further anaylsis.
- mths_since_last_delinq: There are 408'818 (51.19%) missing values. These missing values likely indicate that the borrower has never been delinquent, yet further in-depth analysis is needed. Possible options for imputation in the latter case would be imputation with -1 or to change the datatype to a categorical variable with levels.
- mths_since_last_record: There are 675'190 (84.54%) missing values. These missing values likely indicate that the borrower has no public records, yet further in-depth analysis is needed. Possible options for imputation in the latter case would be imputation with -1 or to change the datatype to a categorical variable with levels.
- revol_bal: There are 2 missing values. These values could be imputed or dropped after further anaylsis.
- revol_util: There are 454 missing values. These values could be imputed or dropped after further anaylsis.
- collections_12_mths_ex_med: There are 126 missing values. These values could be imputed or dropped after further anaylsis.
- mths_since_last_major_derog: There are 599'107 (75.02%) missing values. These missing values likely indicate that the borrower has no major derogatory marks, yet further in-depth analysis is needed. Possible options for imputation in the latter case would be imputation with -1 or to change the datatype to a categorical variable with levels.
- annual_inc_joint, dti_joint, verification_status_joint: These columns have higher NA-counts due to the fact that they are dependent on the application_type being "JOINT". Dropping these columns would lead to a loss of valuable information for joint applications. A comparison model would have to be made to check whether or not the model gets better when keeping this information for prediciton or not.
- tot_coll_amt: There are 63'276 (7.92%) missing values. These values could be imputed or dropped after further anaylsis.
- tot_cur_bal: There are 63'276 (7.92%) missing values. These values could be imputed or dropped after further anaylsis.
- total_rev_hi_lim: There are 63'276 (7.92%) missing values. These values could be imputed or dropped after further anaylsis.
- open_acc_6m, open il_6m, open_il_12m, open_il_24m, mths_since_rcnt_il, total_bal_il, il_util, open_rv_12m, open_rv_24m, max_bal_bc, all_util, inq_fi, total_cu_tl, inq_last_12m: All of these columns have a high number of missing values (97.61%). Imputing these missing values would likely introduce bias into the dataset. Therefore, it is better to drop these columns altogether.

We can filter out the columns with >95% missing values. Imputing these will introduce bias into the dataset. Therefore we will drop these columns.
```{r}
# Columns to drop due to high percentage of missing values
columns_to_drop_na <- c("open_acc_6m",
                         "open_il_6m",
                         "open_il_12m",
                         "open_il_24m",
                         "mths_since_rcnt_il",
                         "total_bal_il",
                         "il_util",
                         "open_rv_12m",
                         "open_rv_24m",
                         "max_bal_bc",
                         "all_util",
                         "inq_fi",
                         "total_cu_tl",
                         "inq_last_12m")
# Drop the columns from the dataset
cleaned_data <- data[, -match(columns_to_drop_na, names(data))]
dim(cleaned_data) # 798'641 rows and 40 columns now
```


It is noticeable that there are exactly 25 rows with NA-values in multiple columns. If these 25 rows are always the same, it would make sense to drop these rows altogether instead of imputing the missing values.
```{r}
# Identify rows with NA values in the specified columns
na_rows <- data %>%
  filter(is.na(delinq_2yrs) |
         is.na(inq_last_6mths) |
         is.na(open_acc) |
         is.na(pub_rec) |
         is.na(total_acc) |
         is.na(acc_now_delinq))
print("Rows with NA values in specified columns:")
print(na_rows)
```
As seen in the output, the 25 rows with NA-values are indeed always the same. Therefore we filter them out of the cleaned dataset.
```{r}
# Remove rows with id values found in na_rows
cleaned_data <- cleaned_data %>%
  filter(!id %in% na_rows$id)
dim(cleaned_data) # 798'616 rows and 40 columns now
```

A similar trend can be seen in tot_call_amt, tot_cur_bal and total_rev_hi_lim, where exactly 63'276 rows have NA-values in all three columns. We check if these rows are the same and filter them out of the cleaned dataset if so.
```{r}
# Identify rows with NA values in the specified columns
na_rows_tot <- data %>%
  filter(is.na(tot_coll_amt) |
         is.na(tot_cur_bal) |
         is.na(total_rev_hi_lim))
print("Rows with NA values in specified columns:")
print(dim(na_rows_tot))
```


## Data Anomalies
To check for data anomalies one should know the logical boundaries of each column. For example, a negative annual income would be an anomaly. Since we do not have this information, we can only check for obvious anomalies such as negative values in columns that should only contain positive values.
Furthermore,  we can check for outliers using boxplots or z-scores. Outliers can be defined as values that are more than 3 standard deviations away from the mean. These outliers can be investigated further to determine if they are valid data points or if they should be removed from the dataset.
Next, one can check for inconsistent data entries. For example, if a column contains categorical data, one can check for typos or inconsistent capitalization. These inconsistencies can be corrected to ensure that the data is consistent throughout the dataset.
Finally, one can check for logical inconsistencies in the data. For example, if a person has no delinquencies in the last 24 months, but has a high number of delinquencies in the last 2 years, this would be a logical inconsistency. These inconsistencies can be investigated further to determine if they are valid data points or if they should be corrected or removed from the dataset.


```{r}
# Check for negative values in numerical columns
numerical_cols <- names(cleaned_data)[sapply(cleaned_data, is.numeric)]
negative_values <- sapply(cleaned_data[numerical_cols], function(col) sum(col < 0, na.rm = TRUE))
print("Negative values in numerical columns:")
print(negative_values)
```
### Outlier-Analysis
```{r}
# Boxplot & Histogram for numerical columns to visualize outliers with ggplot2
for (col_name in numerical_cols) {

  p1 <- ggplot(cleaned_data, aes_string(x = col_name)) +
    geom_histogram(bins = 30) +
    ggtitle(paste("Histogram of", col_name)) +
    theme_minimal()

  p2 <- ggplot(cleaned_data, aes_string(y = col_name)) +
    geom_boxplot() +
    ggtitle(paste("Boxplot of", col_name)) +
    theme_minimal()

  # Combine side-by-side
  print(p1 + p2)
}
```

In the following section we will go through all columns and check for anomalies and outliers. If any are found, we will decide on a case-by-case basis how to handle them.

#### Loan amount (loan_amnt)
```{r}
# Find the top 10 most frequent loan amounts
top_loan_amounts <- cleaned_data %>%
  # Filter for loans in the upper range (e.g., above $30,000)
  filter(loan_amnt >= 30000) %>% 
  count(loan_amnt, sort = TRUE) %>%
  head(20)

print(top_loan_amounts) # Lending cap is found at 35'000, be aware of this in modelling
```
We see a high spike in the upper range of the loan amount. This is due to the fact that Lending Club has a maximum loan amount of $35'000. Therefore, this is not an anomaly but rather a logical boundary.

Due to the big range from minimum to maximum loan amount, one could consider to scale this feature before modelling. A log-transformation could also be considered. A log-transformation column of loan amount is added to the cleaned dataset.
```{r}
# Add log-transformed loan amount column
cleaned_data <- cleaned_data %>%
  mutate(log_loan_amnt = log(loan_amnt + 1)) # Adding

# Check histogram and boxplot of log-transformed loan amount
ggplot(cleaned_data, aes(x = log_loan_amnt)) +
  geom_histogram(bins = 30) +
  ggtitle("Histogram of Log-Transformed Loan Amount") +
  theme_minimal()
ggplot(cleaned_data, aes(y = log_loan_amnt)) +
  geom_boxplot() +
  ggtitle("Boxplot of Log-Transformed Loan Amount") +
  theme_minimal()
```
#### Annual income (annual_inc)
```{r}
# Show the lowest 5 and highest 5 annual incomes. Write as function so it can be used in later analyses
show_extreme_values <- function(data, column_name, n = 5) {
  sorted_data <- data %>%
    arrange(!!sym(column_name))
  
  lowest_values <- head(sorted_data, n)
  highest_values <- tail(sorted_data, n)
  
  list(lowest = lowest_values, highest = highest_values)
}

extreme_annual_inc <- show_extreme_values(cleaned_data, "annual_inc", n = 50)
print("Lowest 5 annual incomes:")
print(extreme_annual_inc$lowest)
print("Highest 5 annual incomes:")
print(extreme_annual_inc$highest)
```
We can see that 2 rows have zero income. For these rows, no employment title or length is given either. The column 'desc' explains that these applicants do not have a job. Therefore, these rows will be kept in the dataset as they provide valuable information about unemployed applicants. The dti (debt-to-income ratio) for these applicants is also given as 999. Since this is a place filler, we should change it to NA for better handling later on.
```{r}
# Change dti value of 999 to NA of the id where annual_inc is 0
cleaned_data <- cleaned_data %>%
  mutate(dti = ifelse(annual_inc == 0 & dti == 9999.00, NA, dti))
```
The variable annual_inc showed a highly skewed distribution with a mean of approximately $75,000, a median of $65,000, and a 75th percentile of $90,000. These values are consistent with typical income patterns in US LendingClub borrower populations, where most applicants earn between $40,000 and $120,000. The standard deviation of roughly $65,000 suggests that even extreme values should rarely exceed $300,000–$350,000, and this aligns with known domain expectations: in comparable datasets, the 99th percentile is typically around $250,000 to $300,000. Despite this, the dataset contained a number of extremely high incomes, including values between $500,000 and several million dollars, with a maximum of $9,500,000. Inspection of the associated employment titles revealed clear contradictions: applicants reporting incomes in the millions frequently listed occupations such as “Driver,” “Registered Nurse,” “Cashier,” or other roles incompatible with such earnings. This strongly suggests data entry or import errors, likely caused by misplaced decimals or dropped separators during ingestion (for example, entering 75,500 as 7,550,000).
Based on both the statistical distribution and domain knowledge, $500,000 was chosen as a conservative upper limit above which values are implausible. This threshold lies well beyond the expected upper tail of genuine borrower incomes and ensures that legitimate high-earning roles such as executives or surgeons are not incorrectly altered. For all observations with annual_inc greater than $500,000, the income value was divided by 100, reflecting the most likely underlying data error pattern. After this correction, the adjusted values integrated smoothly into the natural income distribution, providing additional support for the validity of the transformation. Because dti is directly dependent on income, the dti values for these corrected rows were set to NA, as their original values were derived from corrupted income entries. This approach preserves realistic data structure, corrects clear inconsistencies, and avoids removing potentially informative observations.
```{r}
# Fix annual income values greater than 500'000 by dividing by 100, for the same rows, set dti to NA
cleaned_data <- cleaned_data %>%
  mutate(dti = ifelse(annual_inc > 500000, NA, dti),
         annual_inc = ifelse(annual_inc > 500000, annual_inc / 100, annual_inc)
         )

# Show distribution and boxplot of annual income after fixing
ggplot(cleaned_data, aes(x = annual_inc)) +
  geom_histogram(bins = 30) +
  ggtitle("Histogram of Annual Income (After Fixing)") +
  theme_minimal()
ggplot(cleaned_data, aes(y = annual_inc)) +
  geom_boxplot() +
  ggtitle("Boxplot of Annual Income (After Fixing)") +
  theme_minimal()
```
```{r}
# Show highest values of annual income after fixing
extreme_annual_inc_after <- show_extreme_values(cleaned_data, "annual_inc", n = 10)
print("Highest 10 annual incomes after fixing:")
print(extreme_annual_inc_after$highest)
```
Since the data is still right-skewed, one could consider to scale this feature before modelling. 

#### debt-to-income ratio (dti)
```{r}
extreme_dti <- show_extreme_values(cleaned_data, "dti", n = 100)
print("Lowest 5 dti values:")
print(extreme_dti$lowest)
print("Highest 5 dti values:")
print(extreme_dti$highest)
```
The variable dti (debt-to-income ratio) is another field that exhibits irregularities. After setting dti to NA for rows in which annual_inc was either 0 or exceeded $500,000 (because those income values were clearly corrupted), we examined the remaining distribution. Since installment is not available in this dataset, recalculating DTI from first principles is not possible. This limits the ability to validate each individual entry directly, so data quality assessment must rely on a combination of statistical outlier detection and domain knowledge.
Inspecting the extreme values shows several DTI entries well above 100%, with some exceeding 200%. Such values are implausible for approved consumer loans and contradict typical underwriting standards, where DTIs above 60–70% are rarely accepted. The fact that the associated int_rate values for these rows are not exceptionally high further suggests that these DTIs do not reflect real borrower risk but are instead the result of data errors or import inconsistencies. Therefore, DTI values above 100% were set to NA for later imputation if desired.
On the lower end, dti == 0 is also highly unlikely for the majority of loan applications, particularly when the stated loan purpose is debt_consolidation. This purpose implies existing financial obligations, so a DTI of zero contradicts the expected semantics of the loan record. To investigate this pattern, we inspected the distribution of loan purposes among rows with dti == 0, which supports the conclusion that many of these entries are erroneous or incomplete.
To complement domain-based reasoning, we also examined statistical outliers using the standard criterion of Q3 + 1.5 × IQR. The computed upper bound identifies additional entries far outside the expected range for DTI in consumer lending. Since these values cannot be recalculated due to the absence of installment or any alternative variable representing monthly debt payments, these extreme values cannot be corrected directly. Based on both statistical outlier detection and domain plausibility, they were set to NA. This ensures that clearly invalid DTI values do not influence the modeling process while retaining the remainder of the dataset for downstream analysis.
```{r}
# How many values are above upper bound (Q3 + 1.5*IQR) for dti
dti_Q3 <- quantile(cleaned_data$dti, 0.75, na.rm = TRUE)
dti_IQR <- IQR(cleaned_data$dti, na.rm = TRUE)
dti_upper_bound <- dti_Q3 + 1.5 * dti_IQR
num_dti_above_upper_bound <- sum(cleaned_data$dti > dti_upper_bound, na.rm = TRUE)
print(paste("Number of dti values above upper bound:", num_dti_above_upper_bound))
print(dti_upper_bound)
```

```{r}
### Identify unreasonable values
# Domain logic:
# - DTI > 100% is implausible for approved loans
# - DTI == 0 is implausible for purposes like debt_consolidation

implausible_dti <- cleaned_data %>%
  mutate(
    dti_invalid = case_when(
      dti > 100 ~ TRUE,            # implausible high DTI
      dti == 0 ~ TRUE,             # implausible zero DTI
      TRUE ~ FALSE
    )
  )

table(implausible_dti$dti_invalid)

```

```{r}
### Set implausible values to NA
cleaned_data <- cleaned_data %>%
  mutate(
    dti = ifelse(dti > 100, NA, dti),
    dti = ifelse(dti == 0, NA, dti)
  )
```

```{r}
plot_after <- ggplot(cleaned_data, aes(x = dti)) +
  geom_histogram(bins = 40, fill = "darkgreen", alpha = 0.7) +
  labs(
    title = "DTI Distribution After Cleaning",
    x = "DTI",
    y = "Count"
  ) +
  theme_minimal()

plot_after
```
#### Delinquencies
```{r}
# Inspect contradictions between delinq_2yrs and mths_since_last_delinq
extreme_delinq <- cleaned_data%>%
  filter(delinq_2yrs > 0 & mths_since_last_delinq > 24)
print(extreme_delinq) # 1'261 rows
```
The variables delinq_2yrs (number of delinquencies in the past 24 months) and mths_since_last_delinq (months since the most recent delinquency) both describe aspects of a borrower’s delinquency history and must therefore be internally consistent. A borrower who had at least one delinquency in the past 24 months (delinq_2yrs > 0) must have a corresponding mths_since_last_delinq value between 0 and 24. Any value above 24 months indicates a contradiction, typically originating from stale bureau data or import inconsistencies. Because delinq_2yrs represents a stable bureau summary metric, it is considered more reliable than the more granular timing field. Therefore, we treat mths_since_last_delinq as incorrect in such cases and set it to NA for later imputation.
```{r}
cleaned_data <- cleaned_data %>%
  mutate(mths_since_last_delinq = ifelse(delinq_2yrs > 0 &
  mths_since_last_delinq > 24,
  NA,
  mths_since_last_delinq))
```
We then examine the opposite contradiction: cases where mths_since_last_delinq == 0 indicates a delinquency in the current reporting period, yet delinq_2yrs == 0 suggests no recent delinquencies. A zero-month value is highly specific and therefore more trustworthy than a summary count that may not have been updated in time. In these cases, we correct the count by setting delinq_2yrs to 1.
```{r}
suspicious_rows_zero <- cleaned_data %>%
  filter(mths_since_last_delinq == 0 & delinq_2yrs == 0)
print(suspicious_rows_zero) # 720 rows

cleaned_data <- cleaned_data %>%
  mutate(delinq_2yrs = ifelse(mths_since_last_delinq == 0 &
  delinq_2yrs == 0,
  1,
  delinq_2yrs))
```
A further consistency check is performed using acc_now_delinq, which counts currently delinquent accounts. A borrower cannot have acc_now_delinq > 0 while simultaneously reporting delinq_2yrs == 0, nor can the time since last delinquency be greater than zero. Current delinquency is the strongest and most precise indicator; therefore, contradictions involving acc_now_delinq must be resolved by correcting or removing inconsistent timing information.
```{r}
# Current delinquency contradicting long time-since
cleaned_data %>%
  filter(acc_now_delinq > 0 & mths_since_last_delinq > 0) # 2'908 rows

# Current delinquency contradicting zero count
cleaned_data %>%
  filter(acc_now_delinq > 0 & delinq_2yrs == 0) # 0 rows
```
Since these cases cannot be reliably corrected without full credit bureau histories, we conservatively mark inconsistent timing information as missing.
```{r}
cleaned_data <- cleaned_data %>%
mutate(mths_since_last_delinq = ifelse(acc_now_delinq > 0 &
  mths_since_last_delinq > 0,
  0,
  mths_since_last_delinq))
```
After cleaning, the delinquency fields show internally coherent patterns. However, the presence of NA values creates challenges for predictive modeling. Importantly, NAs in mths_since_last_delinq do not represent missing data but rather the legitimate case of “no delinquencies on record.” To make the data suitable for modeling without inappropriate imputation, we convert delinquency fields to categorical indicators. This approach aligns with industry-standard credit risk modeling practices and avoids encoding NA as missing risk signals.
We first bucketize delinq_2yrs into an ordered categorical variable:
```{r}
cleaned_data <- cleaned_data %>%
mutate(delinq_2yrs_cat = case_when(
  delinq_2yrs == 0 ~ "0",
  delinq_2yrs == 1 ~ "1",
  delinq_2yrs == 2 ~ "2",
  delinq_2yrs >= 3 ~ "3+",
  TRUE ~ NA_character_
  )) %>%
  mutate(delinq_2yrs_cat = factor(delinq_2yrs_cat,
  levels = c("0", "1", "2", "3+"),
  ordered = TRUE))
```
We then convert mths_since_last_delinq into recency buckets, including an explicit category for borrowers with no delinquency history:
```{r}
cleaned_data <- cleaned_data %>%
  mutate(delinq_recency = case_when(
  is.na(mths_since_last_delinq) ~ "none",  # no delinquencies on record
  mths_since_last_delinq == 0 ~ "0",
  mths_since_last_delinq <= 6 ~ "1-6",
  mths_since_last_delinq <= 12 ~ "7-12",
  mths_since_last_delinq <= 24 ~ "13-24",
  TRUE ~ NA_character_
  )) %>%
  mutate(delinq_recency = factor(delinq_recency))
```
Finally, we encode acc_now_delinq as a binary indicator, as is standard in risk modeling:
```{r}
cleaned_data <- cleaned_data %>%
mutate(acc_now_delinq_bin = ifelse(acc_now_delinq > 0, 1, 0))
```
After applying these transformations, delinquency information is internally consistent, free of logical contradictions, and model-ready without requiring imputation.
```{r}
view(cleaned_data)
```


#### Inqueries in last 6 months (Inq_last_6mths)
The variable inq_last_6mths reports how many hard credit inquiries the borrower received within the past six months. Since this is a discrete count of bureau events, it should take small integer values. In typical consumer credit datasets, most borrowers have between 0 and 4 recent inquiries, with gradually fewer cases above that range. Before applying any plausibility or consistency checks, we first examine the raw frequency distribution to understand how the values are spread and whether implausible or unexpected patterns appear.
```{r}
# Count distinct values of inq_last_6mths
inq_counts <- cleaned_data %>%
  group_by(inq_last_6mths) %>%
  summarise(n = n()) %>%
  arrange(inq_last_6mths)

inq_counts
```
#### Inqueries in last 6 months (Inq_last_6mths)
The variable inq_last_6mths reports how many hard credit inquiries the borrower received within the past six months. Since this is a discrete count of bureau events, it should take small integer values. In typical consumer credit datasets, most borrowers have between 0 and 4 recent inquiries, with gradually fewer cases above that range. Before applying any plausibility or consistency checks, we first examine the raw frequency distribution to understand how the values are spread and whether implausible or unexpected patterns appear.
```{r}
# Count distinct values of inq_last_6mths
inq_counts <- cleaned_data %>%
  group_by(inq_last_6mths) %>%
  summarise(n = n()) %>%
  arrange(inq_last_6mths)

inq_counts
```

#### Months since last record (mths_since_last_record) & public records (pub_rec)
The variables mths_since_last_record and pub_rec describe the borrower’s public-record history and therefore must follow strict internal consistency rules. A non-missing value in mths_since_last_record (whether 0 or greater) indicates that the borrower has had at least one public record in the past, meaning that pub_rec must be at least 1. An NA in mths_since_last_record is not a missing value in the statistical sense but represents the meaningful case of “no public record on file,” so pub_rec must be 0. Furthermore, if mths_since_last_record equals 0, a new public record has been reported in the current cycle, which again requires that pub_rec is at least 1. Based on these rules, we first search for violations where a positive month value appears together with a zero or missing public-record count, which would indicate an impossible history.
```{r}
suspicious_pub_record_rows <- cleaned_data %>%
  filter(mths_since_last_record > 0 & (is.na(pub_rec) | pub_rec == 0))
suspicious_pub_record_rows # 0 rows
```
No contradictions of this type appeared in the dataset. Next, we examine rows where mths_since_last_record equals 0 but pub_rec is 0, which incorrectly implies a new public record with a record count of zero. These rows require correction by setting pub_rec to 1.
```{r}
suspicious_pub_record_rows_0 <- cleaned_data %>%
  filter(mths_since_last_record == 0 & pub_rec == 0)
suspicious_pub_record_rows_0 # 1'136 rows

cleaned_data <- cleaned_data %>%
  mutate(pub_rec = ifelse(mths_since_last_record == 0 & pub_rec == 0, 1, pub_rec))
```
We then verify that no cases exist where mths_since_last_record is NA but pub_rec is greater than 0, which would suggest that a public record exists but the time stamp is missing.
```{r}
suspicious_pub_record_rows_na <- cleaned_data %>%
  filter(is.na(mths_since_last_record) & pub_rec > 0)
suspicious_pub_record_rows_na # 0 rows
```
Since any non-missing value of mths_since_last_record must imply at least one public record, rows where the month value is present but pub_rec is NA are inconsistent as well. These NA values are corrected by assigning pub_rec = 1 because the time stamp indicates the presence of at least one public record.
```{r}
cleaned_data <- cleaned_data %>%
  mutate(pub_rec = ifelse(!is.na(mths_since_last_record) & is.na(pub_rec), 1, pub_rec))
```
Rows where both variables are NA cannot be reliably interpreted because it is unclear whether the borrower had no public records or whether the information is missing entirely. Such rows are removed to avoid introducing ambiguity into the data.
```{r}
cleaned_data <- cleaned_data %>%
  filter(!(is.na(mths_since_last_record) & is.na(pub_rec)))
```
After these corrections, the relationship between mths_since_last_record and pub_rec is logically coherent. We then inspect the distribution of values to ensure no remaining anomalies.
```{r}
cleaned_data %>%
  arrange(mths_since_last_record) %>%
  select(loan_amnt, mths_since_last_record, pub_rec, int_rate, term) %>%
  head(20)

cleaned_data %>%
  arrange(desc(mths_since_last_record)) %>%
  select(loan_amnt, mths_since_last_record, pub_rec, int_rate, term) %>%
  head(20)

cleaned_data %>%
  arrange(pub_rec) %>%
  select(loan_amnt, mths_since_last_record, pub_rec, int_rate, term) %>%
  head(20)

cleaned_data %>%
  arrange(desc(pub_rec)) %>%
  select(loan_amnt, mths_since_last_record, pub_rec, int_rate, term) %>%
  head(20)

```
This process ensures that these two variables are mutually consistent and correctly represent the borrower’s public-record history for downstream modeling.

#### open accounts (open_acc)
The variables open_acc, total_acc, revol_bal, total_rev_hi_lim, and revol_util describe the structure and utilization of the borrower’s current revolving credit profile. These fields must satisfy a set of strict logical rules based on how credit files are constructed. First, the number of currently open accounts (open_acc) cannot exceed the total number of accounts ever opened (total_acc), because an account cannot be open without having existed in the first place. We therefore search for cases where open_acc > total_acc, which should be impossible.
```{r}
suspicious_open_acc_rows <- cleaned_data %>%
  filter(open_acc > total_acc)
suspicious_open_acc_rows # 0 rows

```
Zero such rows were identified. If any would have been identified, modifying the historical count of total_acc would risk fabricating a credit history, whereas open_acc is the value that violates a hard structural constraint. For this reason, the safest correction would have been to set open_acc to NA and let the general imputation strategy handle this variable later.


Rows where both values are NA provide no meaningful information about the borrower’s credit file and cannot be logically reconstructed. Since only a small number of rows fall into this category, they are removed.
```{r}
# Amount of rows with both open_acc and total_acc as NA
sum(is.na(cleaned_data$open_acc) & is.na(cleaned_data$total_acc))

cleaned_data <- cleaned_data %>%
  filter(!(is.na(open_acc) & is.na(total_acc)))
```
To verify the integrity of the corrected variables, we inspect the lowest and highest values for both open_acc and total_acc, ensuring that extreme cases remain plausible and that no further contradictions appear.
```{r}
cleaned_data %>%
  arrange(open_acc) %>%
  select(loan_amnt, open_acc, total_acc, int_rate, term) %>%
  head(20)

cleaned_data %>%
  arrange(desc(open_acc)) %>%
  select(loan_amnt, open_acc, total_acc, int_rate, term) %>%
  head(20)

cleaned_data %>%
  arrange(total_acc) %>%
  select(loan_amnt, open_acc, total_acc, int_rate, term) %>%
  head(20)

cleaned_data %>%
  arrange(desc(total_acc)) %>%
  select(loan_amnt, open_acc, total_acc, int_rate, term) %>%
  head(20)
```
We then examine the revolving credit fields. The variable revol_bal represents the total outstanding balance on revolving lines, while total_rev_hi_lim reflects the total available revolving credit limit across all open accounts. Because a borrower cannot legally owe more than the total available limit, revol_bal must never exceed total_rev_hi_lim. Any row where the balance is greater than the maximum allowable credit limit represents a hard structural inconsistency in the data. Such cases must be corrected by setting revol_bal to NA, since the balance figure is the one that violates a fixed upper boundary.
```{r}
suspicious_revol_bal_rows <- cleaned_data %>%
  filter(revol_bal > total_rev_hi_lim)
suspicious_revol_bal_rows # 159 rows

cleaned_data <- cleaned_data %>%
  mutate(revol_bal = ifelse(revol_bal > total_rev_hi_lim, NA, revol_bal))
```
We also check whether the reported utilization ratio (revol_util) is consistent with the balance and credit limit. Utilization should satisfy the relationship 
revol_util = (revol_bal / 100 * total_rev_hi_lim. 
A very large number of rows fail this check, which suggests that inconsistencies are not isolated data-entry errors but instead reflect imperfect source reporting or timing differences. Because utilization is the most stable and meaningful of the three variables, the most defensible correction strategy is to recalculate utilization from the cleaned revol_bal and total_rev_hi_lim values.
```{r}
suspicious_revol_util_rows <- cleaned_data %>%
  filter((revol_util/100) * total_rev_hi_lim < revol_bal)
suspicious_revol_util_rows # 59'187 rows

cleaned_data <- cleaned_data %>%
  mutate(revol_util = ifelse(!is.na(revol_bal) & !is.na(total_rev_hi_lim) & total_rev_hi_lim > 0,
                             pmin(100, (revol_bal / total_rev_hi_lim) * 100),
                             revol_util))
```
Finally, because revolving balances cannot exist without at least one open revolving account, we verify that all rows with revol_bal > 0 also have open_acc > 0. No violations of this rule were found in the dataset.
```{r}
suspicious_revol_bal_open_acc_rows <- cleaned_data %>%
  filter(revol_bal > 0 & open_acc == 0) 
suspicious_revol_bal_open_acc_rows # 0 rows
```
Together, these steps ensure that the structural relationships between open accounts, total accounts, revolving balances, credit limits, and utilization rates are coherent and compliant with established credit-bureau logic, while avoiding any manipulations that would artificially alter a borrower’s reported credit history.


#### Collections and major derogatory events
The field collections_12_mths_ex_med records the number of collection events in the past 12 months and therefore must be a non-negative integer. Since collection events are discrete events recorded by credit bureaus, the value can never be negative or fractional. We identify any rows that violate these conditions and correct them by assigning NA, as such values cannot occur in a real credit file.
```{r}
invalid_coll <- cleaned_data %>%
  filter(collections_12_mths_ex_med < 0 | collections_12_mths_ex_med %% 1 != 0)
invalid_coll # 0 rows
```
The variable mths_since_last_major_derog indicates how many months have passed since the last major derogatory event such as bankruptcy, charge-off, foreclosure, or serious delinquency. As with other “months since event” fields, the interpretation is structural: if the borrower has never experienced a major derogatory, the value must be NA; if the borrower has such a history, the value must be zero or positive. Negative months or fractional months are impossible and must be removed. Although delinquency fields do not directly determine whether a major derogatory event occurred, the variable pub_rec (public records) does: any non-missing value in mths_since_last_major_derog implies at least one corresponding public record. Therefore, the following hard rules apply: if pub_rec == 0, then mths_since_last_major_derog must be NA; if mths_since_last_major_derog is zero or greater, then pub_rec must be at least 1. These checks ensure the two fields remain consistent.
```{r}
invalid_major_derog <- cleaned_data %>%
  filter(mths_since_last_major_derog < 0 |
         mths_since_last_major_derog %% 1 != 0)
invalid_major_derog # 0 rows

suspicious_major_derog_missing <- cleaned_data %>%
  filter(pub_rec == 0 & !is.na(mths_since_last_major_derog))
suspicious_major_derog_missing # 0 rows

suspicious_major_derog_present <- cleaned_data %>%
  filter(pub_rec >= 1 & is.na(mths_since_last_major_derog))
suspicious_major_derog_present # 80'403 rows 
```
If pub_rec == 0 but mths_since_last_major_derog is present, the timing field is incorrect and is set to NA. If pub_rec ≥ 1 but mths_since_last_major_derog is missing, the timing field is restored by assigning NA as representing “unknown months since event,” because the presence of a public record confirms that a major derogatory exists but its recency cannot be reconstructed precisely.
```{r}
# count how often pub_rec == 0 and mths_since_last_major_derog is not NA
suspicious_major_derog_missing <- cleaned_data %>%
  filter(pub_rec == 0 & !is.na(mths_since_last_major_derog))
suspicious_major_derog_missing  # 0 rows

cleaned_data <- cleaned_data %>%
  mutate(mths_since_last_major_derog =
           ifelse(pub_rec == 0, NA, mths_since_last_major_derog))
```
Finally, because mths_since_last_major_derog is not a linearly meaningful numeric variable and NA carries semantic information (“no major derogatory on file”), the variable is transformed into a categorical factor with recency buckets. This follows standard credit-modeling practice, where the recency of serious derogatory events is highly predictive but does not behave like a continuous metric. Very recent major derogatories (0–6 months) are associated with sharply elevated risk and therefore require finer granularity, whereas the predictive strength of a derogatory decreases gradually as it ages, with distinct risk breaks around 12, 24, 48, and 84 months. These cut points reflect how bureaus and scorecard models treat derogatory events: an event occurring this month (0) is the most severe signal; 1–6 and 7–12 months represent the critical recent-risk window; 13–24 and 25–48 months reflect medium-term aging; 49–84 months captures late-aging derogatories that still influence risk; and values above 84 months are typically close to falling off the credit report and contribute only minimal predictive value. This structure aligns with FICO guidelines, common bank scorecard binning strategies, and the natural decay of credit-risk signals over time, making a categorical transformation more appropriate than preserving the raw numeric values.
```{r}
cleaned_data <- cleaned_data %>%
  mutate(major_derog_recency = case_when(
    is.na(mths_since_last_major_derog) ~ "none",
    mths_since_last_major_derog == 0 ~ "0",
    mths_since_last_major_derog <= 6 ~ "1-6",
    mths_since_last_major_derog <= 12 ~ "7-12",
    mths_since_last_major_derog <= 24 ~ "13-24",
    mths_since_last_major_derog <= 48 ~ "25-48",
    mths_since_last_major_derog <= 84 ~ "49-84",
    TRUE ~ ">84"
  )) %>%
  mutate(major_derog_recency = factor(major_derog_recency))
```
The variables tot_coll_amt and tot_cur_bal also require structural validation. The field tot_coll_amt represents the total historical dollar amount ever assigned to collections. As a monetary field recorded in whole dollars, it must be zero or positive; negative values are impossible and are corrected by setting the field to NA.
```{r}
invalid_tot_coll_amt <- cleaned_data %>%
  filter(tot_coll_amt < 0)
invalid_tot_coll_amt # 0 rows
```
Similarly, tot_cur_bal reflects the borrower’s total current balance across all active credit lines and must also be zero or positive. If this balance is negative, the value is replaced by NA. In addition, if a positive total balance is reported, the borrower must have at least one open account. Because open_acc was previously cleaned, we verify that no rows show a positive balance with zero open accounts, which would be structurally impossible in credit reporting.
```{r}
invalid_tot_cur_bal <- cleaned_data %>%
  filter(tot_cur_bal < 0)
invalid_tot_cur_bal # 0 rows

contradiction_tot_cur_bal <- cleaned_data %>%
  filter(tot_cur_bal > 0 & open_acc == 0)
contradiction_tot_cur_bal # 0 rows

```











### Check the distribution of the target
To ensure that the target variable int_rate is reliable for modeling, the distribution is first examined for impossible values, unexpected patterns, or structural inconsistencies. Interest rates in LendingClub data are always expressed as percentages, typically between 5% and 30%, depending on loan grade and risk. They must therefore be non-negative, non-zero, and should not exceed realistic consumer-credit bounds. We begin by summarizing and plotting the distribution.
```{r}
summary(cleaned_data$int_rate)

ggplot(cleaned_data, aes(x = int_rate)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.8) +
  labs(title = "Distribution of Interest Rate",
       x = "Interest Rate (%)",
       y = "Count") +
  theme_minimal()

```
We then check for values that fall outside the plausible range of consumer loan interest rates. Negative interest rates, values equal to zero, or extremely high rates (e.g., above 40%) are structurally impossible in this lending context and indicate a data-entry or import error.
```{r}
invalid_int_rate <- cleaned_data %>%
  filter(int_rate < 0 | int_rate == 0 | int_rate > 40)

invalid_int_rate
```
Interest rates should also be numeric and typically recorded with at most two decimal places. If values contain long decimal expansions or parsing issues, they would appear in the summary above and require correction, but none were detected in this dataset.
Although these are not hard inconsistencies, we also examine whether extremely low or high interest rates appear in combinations that contradict typical underwriting outcomes. Very low interest rates (for example, below 5%) or very high rates (above 30%) are possible but unusual. If they occur together with very risky or very strong credit indicators, this can signal stale data or mismatches between borrower characteristics and the risk-based pricing system. We therefore inspect these rows in context using variables available in the dataset, such as recent delinquencies, revolving utilization, number of recent inquiries, annual income, and open account counts.
```{r}
# Soft plausibility check: extremely low or high interest rates
cleaned_data %>%
  filter(int_rate < 5 | int_rate > 30) %>%
  select(int_rate, delinq_2yrs, revol_util, inq_last_6mths) %>%
  head(20)
```
The structural checks for the target variable yielded no impossible or inconsistent values, and the resulting distribution of int_rate confirms that the data align with expected lending patterns. The histogram shows a smooth, unimodal distribution centered between roughly 10% and 15%, with values ranging from approximately 6% to just under 30%, which is fully consistent with typical unsecured consumer loans. No negative or zero interest rates, no implausibly high values, and no irregular spikes indicative of parsing errors were detected. Together, these results indicate that the interest-rate variable is clean, internally coherent, and suitable to be used as the modeling target without requiring further preprocessing.








```{r}
# Boxplot & Histogram for numerical columns to visualize outliers with ggplot2
for (col_name in numerical_cols) {

  p1 <- ggplot(cleaned_data, aes_string(x = col_name)) +
    geom_histogram(bins = 30) +
    ggtitle(paste("Histogram of", col_name)) +
    theme_minimal()

  p2 <- ggplot(cleaned_data, aes_string(y = col_name)) +
    geom_boxplot() +
    ggtitle(paste("Boxplot of", col_name)) +
    theme_minimal()

  # Combine side-by-side
  print(p1 + p2)
}
```
```{r}
summary(cleaned_data)
```





### Imputation


## Data Analysis

### Feature Selection

### Statistical Analysis

### Model Selection/Engineering
