---
title: "Assignment 1: LendingClub Training"
author: "Group 6"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: paged
---
```{r}
library(readr)
library(dplyr)
```


# Loading The preprocessed data

```{r}
data <- read_csv("cleaned_data.csv")
```

## Order of levels


```{r}
categorical_cols <- names(data)[
  sapply(data, function(x) is.factor(x) || is.character(x))
]

categorical_cols
```


```{r}
num_categories <- sapply(
  data[categorical_cols],
  function(x) length(unique(na.omit(x)))
)

num_categories
```

### emp_length

```{r}
# introduce order for emp_length
data$emp_length <- factor(
  data$emp_length,
  levels = c(
    "< 1 year",
    "1 year",
    "2 years",
    "3 years",
    "4 years",
    "5 years",
    "6 years",
    "7 years",
    "8 years",
    "9 years",
    "10+ years"
  ),
  ordered = TRUE
)
```

```{r}
levels(data$emp_length)
is.ordered(data$emp_length)
```



### verification_status

```{r}
data$verification_status <- factor(
  data$verification_status,
  levels = c(
    "Not Verified",
    "Source Verified",
    "Verified"
  ),
  ordered = TRUE
)
```


```{r}
levels(data$verification_status)
is.ordered(data$verification_status)
```

### home_ownership

```{r}
data$home_ownership <- factor(
  data$home_ownership,
  levels = c("RENT", "MORTGAGE", "OWN"),
  ordered = TRUE
)
```

```{r}
levels(data$home_ownership)
is.ordered(data$home_ownership)
```


### delinq_2yrs_cat

```{r}
data$delinq_2yrs_cat <- factor(
  data$delinq_2yrs_cat,
  levels = c("0", "1", "2", "3+"),
  ordered = TRUE
)
```


```{r}
levels(data$delinq_2yrs_cat)
is.ordered(data$delinq_2yrs_cat)
```

### delinq_recency

```{r}
data$delinq_recency <- factor(
  data$delinq_recency,
  levels = c(
    "0",
    "1-6",
    "7-12",
    "13-24",
    "24+",
    "none"
  ),
  ordered = TRUE
)
```

```{r}
table(data$delinq_recency)
is.ordered(data$delinq_recency)
```

### major_derog_recency

```{r}
data$major_derog_recency <- factor(
  data$major_derog_recency,
  levels = c(
    "0",
    "1-6",
    "7-12",
    "13-24",
    "25-48",
    "49-84",
    "none"
  ),
  ordered = TRUE
)
```

```{r}
table(data$major_derog_recency)
is.ordered(data$major_derog_recency)
```


```{r}
data_ordered <- data
```


# Imputation of missing values
```{r}
missing_summary <- data.frame(
  column = names(data),
  missing_count = sapply(data, function(x) sum(is.na(x))),
  missing_pct = sapply(data, function(x) mean(is.na(x)) * 100)
) %>%
  arrange(desc(missing_pct))

knitr::kable(
  head(missing_summary, 10),
  digits = 4,
  caption = "Top 10 features by missing-value percentage and count (after preprocessing)"
)
```


```{r}
library(tidymodels)
library(xgboost)

set.seed(123)

n <- nrow(data_ordered)
idx_all <- sample.int(n)

n_train <- floor(0.70 * n)
n_val   <- floor(0.15 * n)

train_idx <- idx_all[1:n_train]
val_idx   <- idx_all[(n_train + 1):(n_train + n_val)]
test_idx  <- idx_all[(n_train + n_val + 1):n]

train <- data_ordered[train_idx, ]
val   <- data_ordered[val_idx, ]
test  <- data_ordered[test_idx, ]

val_split <- rsample::make_splits(
  list(analysis = train_idx, assessment = val_idx),
  data = data_ordered
)

val_rs <- rsample::manual_rset(
  splits = list(val_split),
  ids    = "train_val"
)
```

The dataset was split into three mutually exclusive subsets:
- Training set: used to learn model parameters and preprocessing statistics
- Validation set: used for model selection and tuning
- Test set: held out for final, unbiased evaluation

Splitting was performed before any imputation to prevent leakage.



```{r}
ggplot(bind_rows(
  train %>% mutate(split="train"),
  val   %>% mutate(split="val"),
  test  %>% mutate(split="test")
), aes(int_rate, fill=split)) +
  geom_density(alpha=0.4)
```




```{r}
# 2) Recipe (ALL preprocessing here; trained on analysis only)
# ----------------------------
rec <- recipe(int_rate ~ ., data = train) %>%
  # --- special handling for major_derog_recency (do this before generic nominal steps) ---
  step_mutate(
    major_derog_recency = if_else(
      is.na(major_derog_recency),
      "none",
      as.character(major_derog_recency)
    )
  ) %>%
  step_string2factor(major_derog_recency) %>%   # ensures it's a factor again
  step_novel(major_derog_recency) %>%
  step_unknown(major_derog_recency, new_level = "Unknown") %>%

  # --- emp_length: explicit Missing level ---
  step_unknown(emp_length, new_level = "Missing") %>%

  # --- generic protection for *other* categoricals ---
  step_novel(all_nominal_predictors(), -major_derog_recency) %>%
  step_unknown(all_nominal_predictors(), -major_derog_recency, new_level = "Unknown") %>%

  # --- numeric imputation learned on analysis only ---
  step_impute_median(all_numeric_predictors()) %>%

  # --- encoding ---
  step_integer(all_ordered_predictors(), zero_based = FALSE) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors())
```


```{r}
# 3) Model + workflow
# ----------------------------
xgb_spec <- boost_tree(
  trees          = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  sample_size    = tune(),
  mtry           = tune(),
  min_n          = tune()
) %>%
  set_engine(
    "xgboost",
    tree_method = "hist",
    device      = "cuda",
    verbosity   = 2     # 0=silent, 1=warning, 2=info
  ) %>%
  set_mode("regression")

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(xgb_spec)

```


```{r}
# 4) Tuning
# ----------------------------
grid <- grid_latin_hypercube(
  trees(range = c(300, 1500)),
  tree_depth(range = c(2, 10)),
  learn_rate(range = c(log10(0.01), log10(0.3))),
  loss_reduction(),
  sample_size = sample_prop(),
  mtry(range = c(10, 80)),
  min_n(range = c(2, 30)),
  size = 
)

metrics <- metric_set(rmse, mae, rsq)

set.seed(123)
tuned <- tune_grid(
  wf,
  resamples = val_rs,
  grid      = grid,
  metrics   = metrics,
  control   = control_grid(save_pred = TRUE, verbose = TRUE)
)

best <- select_best(tuned, metric = "rmse")
final_wf <- finalize_workflow(wf, best)

best
```

```{r}
final_wf <- finalize_workflow(wf, best)
train_val <- dplyr::bind_rows(train, val)

final_fit <- fit(final_wf, data = train_val)
```

```{r}
train_preds <- predict(final_fit, train) %>%
  bind_cols(train %>% select(int_rate))

metrics(train_preds, truth = int_rate, estimate = .pred)
```

```{r}
val_preds <- predict(final_fit, val) %>%
  bind_cols(val %>% select(int_rate))

metrics(val_preds, truth = int_rate, estimate = .pred)
```


```{r}
test_preds <- predict(final_fit, new_data = test) %>%
  dplyr::bind_cols(test %>% dplyr::select(int_rate))

metrics(test_preds, truth = int_rate, estimate = .pred)
```

```{r}
nrow(test_preds)
summary(test_preds$.pred)
summary(test_preds$int_rate)
```

```{r}
library(tidymodels)
library(xgboost)

# recipe: rec (wie bei Ihnen)
rec_prep <- prep(rec, training = train, retain = TRUE)

train_b <- bake(rec_prep, new_data = train)
val_b   <- bake(rec_prep, new_data = val)
test_b  <- bake(rec_prep, new_data = test)

# X / y trennen
y_train <- train_b$int_rate
y_val   <- val_b$int_rate
y_test  <- test_b$int_rate

x_train <- train_b %>% select(-int_rate)
x_val   <- val_b %>% select(-int_rate)
x_test  <- test_b %>% select(-int_rate)
```


```{r}
p <- ncol(x_train)
colsample_bynode <- min(1, 22 / p)
colsample_bynode
```
```{r}
dtrain <- xgb.DMatrix(data = as.matrix(x_train), label = y_train)
dval   <- xgb.DMatrix(data = as.matrix(x_val),   label = y_val)

params <- list(
  booster          = "gbtree",
  objective        = "reg:squarederror",
  eval_metric      = c("rmse", "mae"),

  # Ihre Parameter:
  max_depth        = 8,                 # tree_depth
  eta              = 0.03,        # learn_rate    # loss_reduction
  subsample        = 0.32,         # sample_size
  colsample_bynode = colsample_bynode,  # mtry (als Anteil)
  min_child_weight = 7,                 # min_n (Mapping in tidymodels/xgb)
  
  tree_method      = "hist"
  # device = "cuda"  # optional, siehe Hinweis unten
)

watchlist <- list(train = dtrain, val = dval)

set.seed(123)
fit_xgb <- xgb.train(
  params  = params,
  data    = dtrain,
  nrounds = 500,         # trees
  watchlist = watchlist,
  verbose = 2
)
```

starts to overfit after 500
