---
title: "Neural_Network"
output: html_document
date: "2025-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Set personal working directory
setwd("~/Documents/Repos/GRP-6_DS-Project")
renv::activate()
```
## Type of Network
The dataset is tabular data with numerical features and a categorical target variable. A feedforward neural network is the most suitable type of network for this classification task.

A convolutional neural network (CNN) is typically used for image data, while a recurrent neural network (RNN) is more suited for sequential data such as time series or text. Since our data does not have spatial or sequential characteristics, a feedforward neural network is the best choice.

```{r libraries}
library(tfruns)
library(keras3)
```

## Execute one experiment
For easier handling the experiment is outsourced to a separate R file "Assignment_2_MLP.R"
```{r run the experiment}
training_run("Assignment_2_MLP.r")
```


# --- Architecture Search ---
# Goal: Find the balance between Model Size (Units) and Overfitting (Dropout)

runs_grid <- list(
  # 1. Model Capacity: Try a "Lean" model vs. a "Wide" model
  units1 = c(128, 256),
  units2 = c(64, 128),
  
  # 2. Regularization: Try Low vs. High Dropout
  # If the model learns poorly, Low dropout helps.
  # If the model overfits, High dropout helps.
  dropout1 = c(0.2, 0.5), 
  dropout2 = c(0.2, 0.5),
  
  # 3. Constants: Keep these fixed for now to reduce noise
  learning_rate = c(0.001),
  batch_size = c(128)
)

# Execute the Tuning Run
# This creates 2 * 2 * 2 * 2 = 16 combinations
# Note: Since our script does 5-Fold CV, this equals 80 total training sessions.
tuning_run("credit_mlp.R", 
           runs_dir = "tuning_architecture", 
           flags = runs_grid,
           echo = FALSE) # echo=FALSE speeds up the console output

# --- View Results ---
# List top 5 models sorted by Validation Accuracy
results <- ls_runs(runs_dir = "tuning_architecture", 
                   order = metric_val_accuracy, 
                   decreasing = TRUE)

# Display the winning hyperparameters
head(results[, c("flag_units1", "flag_units2", "flag_dropout1", "flag_dropout2", "metric_val_accuracy", "metric_val_loss")], 5)














## Preprocessing
Before training the neural network, we need to preprocess the data. Cleaning the data has been done in the previous steps. For a network to work efficiently, we need to scale the numerical features and encode the categorical target variable.

## Splitting the data 
```{r}  
library(dplyr)

set.seed(42) # Answer to everything

# -------------------------------
# Define target and feature matrix
# -------------------------------
target_col <- "status"   # your target column name
data <- cleaned_data

# Shuffle rows
data <- data[sample(nrow(data)), ]

n <- nrow(data)
n_train <- floor(0.70 * n)
n_val   <- floor(0.15 * n)

train_data <- data[1:n_train, ]
val_data   <- data[(n_train + 1):(n_train + n_val), ]
test_data  <- data[(n_train + n_val + 1):n, ]
```

## Prepare the Data for the network
The neural network with keras expects the input features to be in a matrix format and the target variable to be one-hot encoded for multi-class classification.
```{r}
# Features = all non-target columns
feature_cols <- setdiff(names(data), target_col)

x_train <- as.matrix(train_data[, feature_cols])
y_train <- train_data[[target_col]]

x_val   <- as.matrix(val_data[, feature_cols])
y_val   <- val_data[[target_col]]

x_test  <- as.matrix(test_data[, feature_cols])
y_test  <- test_data[[target_col]]
```

