---
title: "EDA_2"
output: html_document
date: "2025-12-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Set personal working directory
#setwd("~/Documents/Repos/GRP-6_DS-Project")
#renv::activate()
```

# Exploratory data analysis
Before we can start training a model, we need to understand the data. This means we need to do some exploratory data analysis (EDA). In this section, we will explore the dataset to understand its structure, identify any missing values, and visualize the data to gain insights. It is important not to look at the target variable during this phase to avoid biasing our analysis. Even when training a neural network, which is a black box method, we should still perform EDA to understand the data better.

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
```

```{r}
# Load the dataset
path <- "/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_data/Dataset-part-2.csv"
data <- read.csv(path)

# Display the dimensions of the dataset
dim(data)
cat("Number of rows:", nrow(data), "\n")
cat("Number of columns:", ncol(data), "\n")

# Display the structure of the dataset
str(data)

# Display the first few rows of the dataset
head(data)
```
We correct the data types of the columns and reload the data file to ensure no information was lost.
```{r}
# 1. Define Data Types (The 'dtype dictionary' equivalent in R)
col_types <- c(
  ID = "character",                  # Prevent ID from being parsed as math number
  CODE_GENDER = "factor",
  FLAG_OWN_CAR = "character",        # Contains "Y"/"N", read as char first
  FLAG_OWN_REALTY = "character",     # Contains "Y"/"N", read as char first
  CNT_CHILDREN = "integer",
  AMT_INCOME_TOTAL = "numeric",
  NAME_INCOME_TYPE = "character",
  NAME_EDUCATION_TYPE = "factor",
  NAME_FAMILY_STATUS = "factor",
  NAME_HOUSING_TYPE = "factor",
  DAYS_BIRTH = "integer",
  DAYS_EMPLOYED = "integer",
  FLAG_MOBIL = "integer",            # Read 0/1 as integer first, then convert
  FLAG_WORK_PHONE = "integer",
  FLAG_PHONE = "integer",
  FLAG_EMAIL = "integer",
  OCCUPATION_TYPE = "factor",        # Factors handle NAs automatically
  CNT_FAM_MEMBERS = "integer",
  status = "factor"
)

# 2. Reload the CSV with strict types
# na.strings includes empty strings to catch missing values in factors
data <- read.csv(path, colClasses = col_types, na.strings = c("NA", ""))

# 3. Post-Load Logic (Conversions that cannot be done directly in read.csv)

# Convert "Y"/"N" to TRUE/FALSE
data$FLAG_OWN_CAR <- data$FLAG_OWN_CAR == "Y"
data$FLAG_OWN_REALTY <- data$FLAG_OWN_REALTY == "Y"

# Convert 0/1 integers to Logical Booleans
flags_numeric <- c("FLAG_MOBIL", "FLAG_WORK_PHONE", "FLAG_PHONE", "FLAG_EMAIL")
data[flags_numeric] <- lapply(data[flags_numeric], as.logical)
str(data)
```

We can check for duplicated rows. These can be entirely duplicated rows or rows with the same ID.
```{r}
# Check for duplicated rows
duplicated_rows <- data[duplicated(data),]
cat("Number of duplicated rows:", nrow(duplicated_rows), "\n") # 0

# Check for duplicated IDs
duplicated_ids <- data[duplicated(data$ID),]
cat("Number of duplicated IDs:", nrow(duplicated_ids), "\n") # 0
```

Now we can check the summary of all columns to identify missing values and get an overview of the data.
```{r}
# Summary of the dataset
summary(data)
```

## Check Flag columns for variance
```{r}
unique(data$FLAG_OWN_CAR)
unique(data$FLAG_OWN_REALTY)
table(data$FLAG_OWN_CAR) 
table(data$FLAG_OWN_REALTY) 
table(data$FLAG_MOBIL) # Only True
table(data$FLAG_WORK_PHONE)
table(data$FLAG_PHONE)
table(data$FLAG_EMAIL)
```
The FLAG_MOBIL column has no variance, therefore we will drop these column.
```{r}
# Drop columns with no variance
data <- data %>%
  select(-FLAG_MOBIL)
```

## Check categorical columns
```{r}
# Check unique values in categorical columns
table(data$CODE_GENDER)
table(data$NAME_INCOME_TYPE)
table(data$NAME_EDUCATION_TYPE)
table(data$NAME_FAMILY_STATUS)
table(data$NAME_HOUSING_TYPE)
table(data$OCCUPATION_TYPE)
```

## Analysis of days_birth and days_employed
Of the numeric columns, we can see using str(data) strange behavior for 'DAYS_BIRTH' and 'DAYS_UNEMPLOYED'. We make a histogram using Hmisc-library to understand the distribution of days_birth (probably the age).
```{r}
# Load Hmisc library for enhanced histogram
library(Hmisc)

# Histogram of DAYS_BIRTH
hist(data$DAYS_BIRTH, 
     main="Histogram of DAYS_BIRTH", 
     xlab="DAYS_BIRTH", 
     col="lightblue")

# Using the summary we can see the distribution of the data
summary(data$DAYS_BIRTH)
```

On the first glance we do not see any outliers. However, we can see that the age is displayed as a negative value. We correct this by dividing by 365 and taking the absolute value. This computes the age in years, which is easier to understand whether or not the data makes sense. Furthermore, we will compute rounded numbers. The model most likely won't benefit from a day-level precision, furthermore most financial relationships are tied to age ranges, which is why this method is sufficient.
```{r}
# Correct DAYS_BIRTH to positive years
data$AGE <- abs(data$DAYS_BIRTH) / 365.25
# Histogram of AGE_YEARS
hist(data$AGE, 
     main="Histogram of AGE", 
     xlab="AGE (years)", 
     col="lightgreen")

# Using the summary we can see the distribution of the data
summary(data$AGE)
```

The second strange column is 'DAYS_UNEMPLOYED'. We will make a histogram to understand the distribution.
```{r}
# Histogram of DAYS_EMPLOYED
hist(data$DAYS_EMPLOYED, 
     main="Histogram of DAYS_EMPLOYED", 
     xlab="DAYS_EMPLOYED", 
     col="lightblue")

# Using the summary we can see the distribution of the data
summary(data$DAYS_EMPLOYED)
```

```{r}
#Show heighest values of days employed excluding pensioned people
top_days_employed <- data %>%
  filter(DAYS_EMPLOYED != 365243) %>%
  arrange(desc(DAYS_EMPLOYED)) %>%
  head(20)
print(top_days_employed)
```

```{r}
# Number of rows where days_employed == 365243
print(nrow(data[data$DAYS_EMPLOYED == 365243, ])) # 11'855 rows

# Look for occupation type where days_employed is 365243
unique(data$OCCUPATION_TYPE[data$DAYS_EMPLOYED == 365243]) # They are ALL NA meaning that they do not work

# Count how many rows have NA in occupation type
nrow(data[is.na(data$OCCUPATION_TYPE), ]) # 20'699 rows --> not all are pensioned

# Show name_income_type where there is an NA in occupation type
table(data$NAME_INCOME_TYPE[is.na(data$OCCUPATION_TYPE)]) 

# Show the amount of name_income_type where days_employed is 365243
table(data$NAME_INCOME_TYPE[data$DAYS_EMPLOYED == 365243])
```
As we see, all people with DAYS_EMPLOYED = 365243 are pensioned, therefore we could create a flag whether or not a person is working and change the sentinel to zero (they are not actively working). In OCCUPATION_TYPE, we know that pensioned people are represented with an NA-value. Yet not all NA in that column are pensioned. There are 8'844 NA values in OCCUPATION_TYPE which are true NA-Values. Due to the high cardinality of the column and the scope of building a neural network, we will impute these values with 'Unknown'. We do not use another method of imputation because of the high cardinality, hoping that the network will figure out the relationships itself.

```{r}
# There are 22 pensioners with NA in occupation_type but have another amount in days_employed which is not the sentinel
weird_pensioners <- data %>%
  filter(NAME_INCOME_TYPE == "Pensioner",
         is.na(OCCUPATION_TYPE),
         DAYS_EMPLOYED != 365243)

nrow(weird_pensioners)     # should be 22
summary(weird_pensioners$DAYS_EMPLOYED)
print(weird_pensioners)
```
There are 22 rows where NAME_INCOME_TYPE is 'Pensioner', yet with a normal amount of employment years. Furthermore, the age of the people in these rows are below the expected pension age. Therefore we will change their NAME_INCOME_TYPE to 'Working'.
```{r}
data <- data %>%
  mutate(
    NAME_INCOME_TYPE = ifelse(
      NAME_INCOME_TYPE == "Pensioner" & DAYS_EMPLOYED != 365243,
      "Working",
      NAME_INCOME_TYPE
    )
  )
```

```{r}
# We reclasify the NA values in OCCUPATION_TYPE
data <- data %>%
  mutate(
    OCCUPATION_TYPE = as.character(OCCUPATION_TYPE),
    OCCUPATION_TYPE = case_when(
      DAYS_EMPLOYED == 365243 ~ "Retired",
      is.na(OCCUPATION_TYPE)  ~ "Unknown",
      TRUE ~ OCCUPATION_TYPE
    ),
    OCCUPATION_TYPE = as.factor(OCCUPATION_TYPE)
  )
```

```{r}
# Create a flag for pensioned people
#data$FLAG_WORKING <- ifelse(data$DAYS_EMPLOYED == 365243, 0, 1)

# Set DAYS_EMPLOYED to 0 for pensioned people
data$DAYS_EMPLOYED[data$DAYS_EMPLOYED == 365243] <- 21915 # This keeps the ordinal order in days_employed, showing that pensioners have worked for a long time, giving more security

# Create a new categorical level for missing occupation
#data$OCCUPATION_TYPE <- as.character(data$OCCUPATION_TYPE)
#data$OCCUPATION_TYPE[is.na(data$OCCUPATION_TYPE)] <- "Unknown"
#data$OCCUPATION_TYPE <- as.factor(data$OCCUPATION_TYPE)
# Pensioners will also get Unknown flag, yet are identified by the FLAG_WORKING column. Therefore we do not have to make another category for them and increase the cardinality even more.
```

Now we can transform the DAYS_EMPLOYED column to represent years employed and set pensioned people to zero.
```{r}
# Calculate ACTIVE_EMPLOYMENT_YEARS
data$ACTIVE_EMPLOYMENT_YEARS <- abs(data$DAYS_EMPLOYED) / 365.25

# Make Histogram of active employment years
hist(data$ACTIVE_EMPLOYMENT_YEARS,
     main="Histogram of ACTIVE_EMPLOYMENT_YEARS", 
     xlab="ACTIVE_EMPLOYMENT_YEARS (years)", 
     col="lightblue")
```
As we see a very skewed distribution we will make a log-transformation to use in the NN.
```{r}
# Log-transform ACTIVE_EMPLOYMENT_YEARS
data$ACTIVE_EMPLOYMENT_YEARS_LOG <- log1p(data$ACTIVE_EMPLOYMENT_YEARS)
# Histogram of log-transformed ACTIVE_EMPLOYMENT_YEARS
hist(data$ACTIVE_EMPLOYMENT_YEARS_LOG,
     main="Histogram of Log-Transformed ACTIVE_EMPLOYMENT_YEARS",
     xlab="Log(ACTIVE_EMPLOYMENT_YEARS + 1)",
     col="lightblue")
```

```{r}
library(ggplot2)

# 1. Create a plot-specific column for coloring
# We determine the group based on your sentinel value of 60
data$Employment_Status <- ifelse(data$ACTIVE_EMPLOYMENT_YEARS == 60, "Pensioner", "Working")

# 2. Generate the Plot
ggplot(data, aes(x = AGE, y = ACTIVE_EMPLOYMENT_YEARS, color = Employment_Status)) +
  
  # A. The Scatter Points
  geom_point(alpha = 0.6, size = 1.5) + 
  
  # B. The "Biological Limit" Line (y = x)
  # Points above this dashed line represent the "Impossible Zone" (Working longer than alive)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  annotate("text", x = 25, y = 30, label = "Biological Limit (y=x)", 
           color = "black", angle = 45, vjust = -1) +

  # C. Colors & Styling
  scale_color_manual(values = c("Pensioner" = "#E41A1C", "Working" = "#377EB8")) +
  
  labs(
    title = "Age vs. Active Employment Years",
    subtitle = "Pensioners fixed at 60 years (Visual Artifact above the limit line)",
    x = "Age (Years)",
    y = "Active Employment Years",
    color = "Group"
  ) +
  
  theme_minimal() +
  theme(legend.position = "top")
```

Is there a strong relationship between income and occupation? We explore this relationship so we can impute all the NAs for occupation by their income. We will impute later so we can compare the different dataset in similar models to see how they perform.
```{r}
Occupation_Income<-data%>%
  filter(!is.na(OCCUPATION_TYPE))%>%
  group_by(OCCUPATION_TYPE)%>%
  summarise(MEAN_INCOME = mean(AMT_INCOME_TOTAL))%>%
  arrange(desc(MEAN_INCOME))#, .by_group=TRUE

Occupation_Income
```

## Analysis of CNT_FAM_MEMBERS and CNT_CHILDREN
```{r}
# show rows where CNT_FAM_MEMBERS > 10
outlier_fam_members <- data[data$CNT_FAM_MEMBERS > 10, ]
print(outlier_fam_members)
```

```{r}
# show rows where CNT_CHILDREN > 10
outlier_children <- data[data$CNT_CHILDREN > 6, ]
print(outlier_children)
```
We see heavy outliers in the family members column. A 38-year-old- with 12 children is statistically extremely improbable and likely a data error. Since it are only 2 rows we will drop them.
```{r}
#  Remove rows with CNT_FAM_MEMBERS > 10
data <- data[data$CNT_FAM_MEMBERS <= 10, ]
```

```{r}
# show rows where CNT_FAM_MEMBERS < CNT_CHILDREN
outlier_children <- data[data$CNT_FAM_MEMBERS < data$CNT_CHILDREN, ]
print(outlier_children)
print(nrow(outlier_children))
```

A logical consistency check was performed to validate the relationship between CNT_FAM_MEMBERS (Total Family Size) and CNT_CHILDREN. This validation identified 32 instances where the reported number of family members was strictly less than the number of children. Since the count of family members must mathematically include all children plus at least one adult, these records were flagged as erroneous.

```{r}
# Identify anomalous rows
anomaly <- data$CNT_FAM_MEMBERS < data$CNT_CHILDREN # 32 rows

# Apply corrections based on marital status
data$CNT_FAM_MEMBERS[anomaly & data$NAME_FAMILY_STATUS %in% c("Married", "Civil marriage")] <-
  data$CNT_CHILDREN[anomaly & data$NAME_FAMILY_STATUS %in% c("Married", "Civil marriage")] + 2

data$CNT_FAM_MEMBERS[anomaly & data$NAME_FAMILY_STATUS %in% c("Separated", "Widow", "Single / not married")] <-
  data$CNT_CHILDREN[anomaly & data$NAME_FAMILY_STATUS %in% c("Separated", "Widow", "Single / not married")] + 1

# Recheck anomalies, show rows where CNT_FAM_MEMBERS < CNT_CHILDREN
outlier_children_corrected <- data[data$CNT_FAM_MEMBERS < data$CNT_CHILDREN, ]
print(outlier_children_corrected)
```
To rectify these inconsistencies without discarding data, a deterministic imputation logic was applied based on the individual's NAME_FAMILY_STATUS.
- Couples: For individuals listed as "Married" or "Civil marriage," the family member count was corrected to equal the number of children plus two (representing the two partners).
- Single Adults: For individuals listed as "Separated," "Widow," or "Single / not married," the family member count was corrected to equal the number of children plus one (representing the single parent/adult).

This correction resolved the logical anomalies in all 32 identified cases, ensuring that CNT_FAM_MEMBERS is always greater than or equal to CNT_CHILDREN + 1 across the entire dataset.


## Analysis of AMT_INCOME_TOTAL
Make histogram of amt_income_total and look for outliers.
```{r}
# Histogram of AMT_INCOME_TOTAL
hist(data$AMT_INCOME_TOTAL,
     main="Histogram of AMT_INCOME_TOTAL", 
     xlab="AMT_INCOME_TOTAL", 
     col="lightblue")

# Make Boxplot of AMT_INCOME_TOTAL
boxplot(data$AMT_INCOME_TOTAL,
        main="Boxplot of AMT_INCOME_TOTAL",
        ylab="AMT_INCOME_TOTAL",
        col="lightgreen")

# Using the summary we can see the distribution of the data
summary(data$AMT_INCOME_TOTAL)
```
Show the top 20 rows with highest income.
```{r}
# Show top 20 rows with highest AMT_INCOME_TOTAL
top_income <- data %>%
  arrange(desc(AMT_INCOME_TOTAL)) %>%
  head(20)
print(top_income)
```

```{r}
boxplot(data$AMT_INCOME_TOTAL ~ data$NAME_EDUCATION_TYPE, main="Income by Education Type", ylab = "Income", xlab = "Cut",col="blue")
unique(data$NAME_EDUCATION_TYPE)
```

```{r}
boxplot(data$AMT_INCOME_TOTAL ~ data$OCCUPATION_TYPE, main="Income by Occupation Type", ylab = "Income", xlab = "Cut",col="blue")
```

```{r}
#2.Numerical v Numerical
data %>% plot(AMT_INCOME_TOTAL ~ AGE, data = ., ylab="Total Income", xlab="Age", main="Age Versus Income")
```


We see that there are heavy outliers in the income data. Let's calculate the threshold of the 99.9% percentile.

```{r}
# 1. Define the list of jobs that should NOT be earning > 1 Million
impossible_jobs <- c("Laborers", "Cleaning staff", "Secretaries", 
                     "Low-skill Laborers", "Drivers", "Waiters/barmen staff")

# 2. Filter logic:
# Remove row IF (Income > 1M  AND  Job is in impossible_jobs)
data <- data %>%
  filter( !(AMT_INCOME_TOTAL > 1000000 & OCCUPATION_TYPE %in% impossible_jobs) )
```

```{r}
# Calculate 99.9% percentile threshold
#threshold <- quantile(data$AMT_INCOME_TOTAL, 0.999)
#cat("99.9% percentile threshold for AMT_INCOME_TOTAL:", threshold, "\n")
# Identify outliers above the threshold
#outliers_income <- data[data$AMT_INCOME_TOTAL > threshold, ]
#cat("Number of outliers above 99.9% percentile:", nrow(outliers_income), "\n") # 58 rows
```
--Analysis of the AMT_INCOME_TOTAL feature identified 58 extreme outliers exceeding the 99.9th percentile threshold of 900,000, representing just 0.08% of the dataset. These instances included probable data entry --errors, such as a "Laborer" with a recorded income of 6.75 million. Consequently, these rows were removed rather than imputed or capped.

This exclusion was necessary to strictly maintain the numerical stability required for Neural Network training. Retaining such extreme values would compress the vast majority of useful data into a microscopic range during normalization, preventing the model from learning meaningful distinctions between standard income levels. Furthermore, dropping these rows was preferred over imputation because the data points were likely erroneous rather than merely missing; preserving them with guessed values could have introduced spurious relationships into the model. Similarly, capping the values was avoided to prevent the creation of an artificial "ceiling effect" at the upper bound that could bias the model's interpretation of high-income entities.

```{r}
# Remove all rows with AMT_INCOME_TOTAL above the 99.9% percentile threshold
#data <- data[data$AMT_INCOME_TOTAL <= threshold, ]
#cat("New number of rows after removing income outliers:", nrow(data), "\n")

# transform column by applying a log transformation
data$AMT_INCOME_TOTAL_LOG <- log1p(data$AMT_INCOME_TOTAL)

# Histogram of log-transformed AMT_INCOME_TOTAL
hist(data$AMT_INCOME_TOTAL_LOG,
     main="Histogram of Log-Transformed AMT_INCOME_TOTAL",
     xlab="Log(AMT_INCOME_TOTAL + 1)",
     col="lightblue")
```

We create an interaction term giving the model context of family size
```{r}
data$INCOME_PER_FAMILY_MEMBER <- data$AMT_INCOME_TOTAL / data$CNT_FAM_MEMBERS

# Histogram of INCOME_PER_FAMILY_MEMBER
hist(data$INCOME_PER_FAMILY_MEMBER,
     main="Histogram of INCOME_PER_FAMILY_MEMBER",
     xlab="INCOME_PER_FAMILY_MEMBER",
     col="lightblue")

# Make log transformation of income_per_family_memberS
data$INCOME_PER_FAMILY_MEMBER_LOG <- log1p(data$INCOME_PER_FAMILY_MEMBER)
```


## Feature Engineering
```{r}
data$EMPLOYMENT_RATIO <- data$ACTIVE_EMPLOYMENT_YEARS / data$AGE
# (Higher = Spent more of their life working)
hist(data$EMPLOYMENT_RATIO,
     main="Histogram of EMPLOYMENT_RATIO",
     xlab="EMPLOYMENT_RATIO",
     col="lightblue")
data$CREDIT_MATURITY <- data$AGE - data$ACTIVE_EMPLOYMENT_YEARS

# (Approximate age when they started working)
hist(data$CREDIT_MATURITY,
     main="Histogram of CREDIT_MATURITY",
     xlab="CREDIT_MATURITY",
     col="lightblue")
data$INCOME_PER_AGE <- data$AMT_INCOME_TOTAL / data$AGE
# (Earning power relative to life stage)
hist(data$INCOME_PER_AGE,
     main="Histogram of INCOME_PER_AGE",
     xlab="INCOME_PER_AGE",
     col="lightblue")

# Log scaling for income_per_Age
data$INCOME_PER_AGE_LOG <- log1p(data$INCOME_PER_AGE)

# 3. Check the histogram
hist(data$INCOME_PER_AGE_LOG, main="Log-Income Per Age", col="lightblue")
```

# Check Target distribution
```{r}
# Look at the distribution of the target variable histogram
ggplot(data, aes(x = status)) + 
  geom_bar(fill = "steelblue", color = "black") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5) + # Add count labels on top
  labs(
    title = "Distribution of Target Variable (Status)",
    subtitle = "Majority Class (0) vs. Minorities (1-5) vs. Special (C, X)",
    x = "Status Category",
    y = "Count"
  ) +
  theme_minimal()
```

## Clean dataset so only the necessary columns remain
Show all levels of the categorical columns and correct if necessary.
```{r}
str(data)
```

```{r}
library(dplyr)
library(fastDummies)
# We drop all rows with target X
data <- data %>% filter(status != "X")

# TARGET ENCODING
# Convert status to character to avoid factor issues
data$status <- as.character(data$status)

# Map status levels to class IDs 0..7 (unordered)
data$target_class <- recode(data$status,
                            "C" = 0,
                            "0" = 1,  # Was 2
                            "1" = 2,  # Was 3
                            "2" = 3,  # Was 4
                            "3" = 4,  # Was 5
                            "4" = 5,  # Was 6
                            "5" = 6
                            ) |> as.numeric()


# BOOLEAN FLAGS → NUMERIC
bool_cols <- c(
  "FLAG_OWN_CAR", "FLAG_OWN_REALTY",
  "FLAG_WORK_PHONE", "FLAG_PHONE", "FLAG_EMAIL"
)

# Convert TRUE/FALSE or Y/N encoded booleans into numeric 0/1
data[bool_cols] <- lapply(data[bool_cols], as.numeric)


# EDUCATION (ORDINAL)
# The ONLY truly ordinal categorical variable.

data$EDUCATION_ENCODED <- case_when(
    data$NAME_EDUCATION_TYPE == "Lower secondary" ~ 0,
    data$NAME_EDUCATION_TYPE == "Secondary / secondary special" ~ 1,
    data$NAME_EDUCATION_TYPE == "Incomplete higher" ~ 2,
    data$NAME_EDUCATION_TYPE == "Higher education" ~ 3,
    data$NAME_EDUCATION_TYPE == "Academic degree" ~ 4,
    TRUE ~ NA_real_
)

# Drop original education variable
data <- data %>% select(-NAME_EDUCATION_TYPE)

# ONE-HOT ENCODING (NOMINAL)
# No ranking between categories → must be one-hot encoded.

nominal_cols <- c(
  "CODE_GENDER",
  "NAME_INCOME_TYPE",
  "NAME_FAMILY_STATUS",
  "NAME_HOUSING_TYPE",
  "OCCUPATION_TYPE"
)

# Ensure factors
data[nominal_cols] <- lapply(data[nominal_cols], as.factor)

# One-hot encode
data <- dummy_cols(
  data,
  select_columns = nominal_cols,
  remove_selected_columns = TRUE,
  remove_first_dummy = FALSE   # keep full set
)


# 5. DROP UNUSED OR IGNORED COLUMNS (OPTIONAL)
# ID is not a feature.
data <- data %>% select(-ID, -DAYS_BIRTH, -DAYS_EMPLOYED, -AMT_INCOME_TOTAL, -ACTIVE_EMPLOYMENT_YEARS, -INCOME_PER_FAMILY_MEMBER, -INCOME_PER_AGE, -Employment_Status, -status)

str(data)
```

```{r}
dim(data)
```



## Save the dataframe
```{r}
write.csv(
  data,
  file = "/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_data/Intermediate/Assignment_2_cleaned_v3.csv",
  row.names = FALSE
)
```

