---
title: "Assignment_2_Documentation"
output: pdf_document
date: "2025-12-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

# Introduction

The goal of this assignment was to build a neural network to predict the pay-back behavior of a new customer. The prediction of this behavior is categorical in nature, having 8 different possible levels:

-   no loan for the month

-   paid of that month

-   1-29 days past due

-   30-59 days past due

-   60-89 days overdue

-   90-119 days overdue

-   120-149 days overdue

-   Overdue or bad debts, write-offs for more than 150 days.

The method to be used is a neural network, a supervised non-parametric learning method. Supervised learning means that the data is labeled. A non-parametric method is a statistical technique that does not assume the data follows a specific probability distribution (like the "bell curve" or normal distribution), but instead relies on the data's rank or order to draw conclusions.

The data is an adjusted form of an open source kaggle-dataset, provided by the lecturers. No extra data was used to enhance this dataset.

# Workflow

In a data science project, adhering to a structured workflow is critical. Before starting the project, a GitHub repository was initialized to ensure reproducibility and, most importantly, to maintain a transparent history of all changes and experiments.

Furthermore, a virtual environment was established to manage dependencies and ensure software consistency across different machines.

## Preprocessing (Data Cleaning)

This is the most time-consuming phase of a data science project. Raw data is rarely ready for modeling immediately, and thorough cleaning is essential to ensure high-quality output. If noisy or poor-quality data is fed into a model, the outcome will inevitably be of low value (a concept known as "Garbage in, Garbage out").

Preprocessing entails handling missing values, correcting inconsistencies, removing duplicate entries, formatting data types, and splitting the data into training, validation, and testing sets to prevent data leakage.

## Exploratory Data Analysis (EDA)

Before building a model, the data scientist must understand the data's behavior. This is achieved through visualization methods, statistical checks, and outlier detection. Findings in this phase often lead to a subsequent round of preprocessing. Consequently, the workflow frequently becomes an iterative back-and-forth between EDA and preprocessing, consuming the majority of the project's time.

## Feature Engineering

In this phase, domain knowledge is typically applied to maximize model performance. Although the team's background in life sciences meant limited prior exposure to specific business domain cases, general data transformation principles were effectively applied. The goal is to transform the data to make it easier for the machine learning algorithm to interpret patterns. Depending on the model's initial performance, new features or different transformations are often introduced to optimize results based on emerging findings.

## Model Selection and Training

For this project, the model type was predefined as a neural network. However, critical decisions remained regarding the specific architecture (e.g., MLP, CNN, RNN) and hyperparameter configuration. The process begins with a baseline model—a simple architecture used to establish a performance benchmark. By testing different architectures and tuning hyperparameters, one attempts to optimize the applicable evaluation metrics. Crucially, this optimization process relies exclusively on the training and validation sets created during preprocessing.

## Evaluation

Finally, the model's performance is assessed using metrics relevant to the problem—specifically loss and accuracy, as defined by the project requirements. This assessment is conducted using the test data set, which was set aside at the beginning of the workflow to ensure an unbiased evaluation of the model's generalization capability.

```{r}
# Set personal working directory 
# Change directory as needed
setwd("~/Documents/Repos/GRP-6_DS-Project")
renv::activate()
```

# Preprocessing and EDA

In this chapter, the preprocessing pipeline is documented step-by-step. Crucially, this section reflects the iterative nature of the project. It details not only the final code but also the initial attempts that were later discarded or refined due to poor model accuracy.

```{r, eval = FALSE}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(fastDummies)
library(Hmisc)
library(corrplot)
library(tfruns)
library(tictoc)
library(beepr)
library(keras3)
```

## Data Import and Initial Inspection

The first thing we do is load the csv into a dataframe and inspect the amount of rows and columns, as well as the datatypes needed for each column.

```{r}
# Load the dataset
path <- "/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_data/Dataset-part-2.csv"
original <- read.csv(path)

# Display the dimensions of the dataset
dim(original)
cat("Number of rows:", nrow(original), "\n")
cat("Number of columns:", ncol(original), "\n")

# Display the structure of the dataset
str(original)

# Display the first few rows of the dataset
head(original)
```

The dataset has an initial of 67'614 rows and 19 columns (18 features and 1 label). We reload the dataset with the correct datatypes using the knowledge on the columns found on kaggle. It is crucial to load datatypes like ID as a string since sometimes the programm would change the value due to its formatting.

Flag columns are changed to integers with TRUE being 1 and FALSE being 0.

```{r}
# 1. Define Data Types (The 'dtype dictionary' equivalent in R)
col_types <- c(
  ID = "character",                  # Prevent ID from being parsed as math number
  CODE_GENDER = "factor",
  FLAG_OWN_CAR = "character",        
  FLAG_OWN_REALTY = "character",     
  CNT_CHILDREN = "integer",
  AMT_INCOME_TOTAL = "numeric",
  NAME_INCOME_TYPE = "character",
  NAME_EDUCATION_TYPE = "factor",
  NAME_FAMILY_STATUS = "factor",
  NAME_HOUSING_TYPE = "factor",
  DAYS_BIRTH = "integer",
  DAYS_EMPLOYED = "integer",
  FLAG_MOBIL = "integer",            
  FLAG_WORK_PHONE = "integer",
  FLAG_PHONE = "integer",
  FLAG_EMAIL = "integer",
  OCCUPATION_TYPE = "factor",        
  CNT_FAM_MEMBERS = "integer",
  status = "factor"
)

# 2. Reload the CSV with strict types
# na.strings includes empty strings to catch missing values in factors
data <- read.csv(path, colClasses = col_types, na.strings = c("NA", ""))

# 3. Post-Load Logic (Conversions that cannot be done directly in read.csv)
# Manually convert "Y" to 1 and "N" to 0
data$FLAG_OWN_CAR <- ifelse(data$FLAG_OWN_CAR == "Y", 1, 0)
data$FLAG_OWN_REALTY <- ifelse(data$FLAG_OWN_REALTY == "Y", 1, 0)

# (Optional) Verify that all flags are now technically integers
flags_list <- c("FLAG_MOBIL", "FLAG_WORK_PHONE", "FLAG_PHONE", "FLAG_EMAIL", "FLAG_OWN_CAR", "FLAG_OWN_REALTY")

# Force them all to be integers to be safe
data[flags_list] <- lapply(data[flags_list], as.integer)

str(data)
```

After each transformation a visual check using str(data) is used to check whether changes have been done as wished.

Next, we check for duplicated rows or rows with the same ID

```{r}
# Check for duplicated rows
duplicated_rows <- data[duplicated(data),]
cat("Number of duplicated rows:", nrow(duplicated_rows), "\n") # 0

# Check for duplicated IDs
duplicated_ids <- data[duplicated(data$ID),]
cat("Number of duplicated IDs:", nrow(duplicated_ids), "\n") # 0

# Check for duplicated rows when excluding ID-column
duplicates_mask <- duplicated(data[, !names(data) %in% "ID"])

# Count the number of duplicates
num_duplicates <- sum(duplicates_mask)

cat("Number of duplicate rows (excluding ID):", num_duplicates, "\n")
cat("Percentage of total dataset:", round((num_duplicates / nrow(data)) * 100, 2), "%\n")
```

Since ID is an identifier of a customer, it will not be fed into the model. Furthermore, it is highly unlikely that two distinct people have the same age, years worked, etc. We find 23 rows with similar information (excluding the ID), which can be dropped.

```{r}
# Optional: Remove the duplicates found above
data <- data[!duplicates_mask, ]
```

## Data summary

Now we can check the summary of all columns to identify missing values and get an overview of the data.

```{r}
# Summary of the dataset
summary(data)
```

The`summary()`function acts as a critical diagnostic tool in the early stages of preprocessing by providing a five-number summary for numerical variables and data type information for categorical ones. This quick overview helps identify scale differences between variables, such as the vast difference between`CNT_CHILDREN`and`AMT_INCOME_TOTAL`, which indicates a need for normalization. Furthermore, it allows for the immediate detection of outliers or physically impossible values, as well as the identification of constant variables with zero variance that add no predictive value to the model.

**Observations from Output:** The summary output reveals several critical data quality issues that require handling. Most notably,`DAYS_EMPLOYED` contains a significant artifact with a maximum value of 365'243, which is roughly 1,000 years. This is likely a placeholder for pensioners or unemployed individuals and severely skews the mean compared to the median, requiring correction before training. Additionally,`FLAG_MOBIL`shows a minimum, mean, and maximum of 1, indicating that every row contains the same value; this feature offers no discriminative information and should be removed. It is also observed that `DAYS_BIRTH` and `DAYS_EMPLOYED` are recorded as negative values, which should be converted to positive figures for better interpretability. Finally, many categorical variables like `CODE_GENDER` and `NAME_EDUCATION_TYPE` are currently stored as characters and must be converted to factors to properly utilize them in the modeling phase.

## Univariate columns

As we already seen in the summary(), FLAG_MOBIL shows a mean, minimum and max of 1, indicating that there is only one value present. We check this with the table() function. We do this simultaneously for the other FLAG-columns to see whether there is a strong imbalance in the column.

```{r}
unique(data$FLAG_OWN_CAR)
unique(data$FLAG_OWN_REALTY)
table(data$FLAG_OWN_CAR) 
table(data$FLAG_OWN_REALTY) 
table(data$FLAG_MOBIL) # Only True
table(data$FLAG_WORK_PHONE)
table(data$FLAG_PHONE)
table(data$FLAG_EMAIL)
```

```{r}
# Drop columns with no variance
data <- data %>%
  select(-FLAG_MOBIL)
```

The column FLAG_MOBIL is redundant since it only has one distinct value and can therefore be dropped.

## Analysis of days_of_birth and days_employed

Both of these columns show in the summary negative values. In the datadictionary we learn that DAYS_BIRTH is the age of the customer. It is a backwards count from the current day (0) where -1 would mean yesterday. DAYS_EMPLOYED is the start date of employment, it holds the same logic as DAYS_BIRTH. If the value is positive, it means that the person is currently unemployed.

```{r}
# Histogram of DAYS_BIRTH
hist(data$DAYS_BIRTH, 
     main="Histogram of DAYS_BIRTH", 
     xlab="DAYS_BIRTH", 
     col="lightblue")

# Using the summary we can see the distribution of the data
summary(data$DAYS_BIRTH)
```

On the first glance we do not see any outliers. However, we can see that the age is displayed as a negative value. We correct this by dividing by 365 and taking the absolute value. This computes the age in years, which is easier to understand whether or not the data makes sense. Furthermore, we will compute rounded numbers. The model most likely won't benefit from a day-level precision, furthermore most financial relationships are tied to age ranges, which is why this method is sufficient.

```{r}
# Correct DAYS_BIRTH to positive years
data <- data %>%
  mutate(AGE = round(abs(DAYS_BIRTH) / 365.25)) %>%
  select(-DAYS_BIRTH)
# Histogram of AGE_YEARS
hist(data$AGE, 
     main="Histogram of AGE", 
     xlab="AGE (years)", 
     col="lightgreen")

# Using the summary we can see the distribution of the data
summary(data$AGE)
```

The second strange column is 'DAYS_UNEMPLOYED'. We will make a histogram to understand the distribution.

```{r}
# Histogram of DAYS_EMPLOYED
hist(data$DAYS_EMPLOYED, 
     main="Histogram of DAYS_EMPLOYED", 
     xlab="DAYS_EMPLOYED", 
     col="lightblue")

# Using the summary we can see the distribution of the data
summary(data$DAYS_EMPLOYED)
```

We know that the unemployed have the sentinel value of 365'243. We therefore would expect the same amount of NA-values in OCCUPATION_TYPE.

```{r}
# Number of rows where days_employed == 365243
print(nrow(data[data$DAYS_EMPLOYED == 365243, ])) # 11'855 rows

# Look for occupation type where days_employed is 365243
unique(data$OCCUPATION_TYPE[data$DAYS_EMPLOYED == 365243]) # They are ALL NA meaning that they do not work

# Count how many rows have NA in occupation type
nrow(data[is.na(data$OCCUPATION_TYPE), ]) # 20'699 rows --> not all are pensioned

# Show name_income_type where there is an NA in occupation type
table(data$NAME_INCOME_TYPE[is.na(data$OCCUPATION_TYPE)]) 

# Show all occupation_type where name_income_type is Pensioner
table(data$OCCUPATION_TYPE[data$NAME_INCOME_TYPE == 'Pensioner'])

# Show the amount of name_income_type where days_employed is 365243
table(data$NAME_INCOME_TYPE[data$DAYS_EMPLOYED == 365243])

# Total amount of pensioners in the dataset
table(data$NAME_INCOME_TYPE == 'Pensioner')
```

We see that there are 11'855 rows with positive values, indicating 11'855 unemployed which happen all to be pensioners. Looking at the OCCUPATION_TYPE we can see that those with the sentinel value have an NA-value in OCCUPATION_TYPE as well.

-   Sentinel Group (11'855 rows)

    -   There are exactly 11'855 rows where DAYS_EMPLOYED is 365'243.

    -   Every single one of these rows has NAME_INCOME_TYPE as 'Pensioner'.

    -   Every single one of these rows has OCCUPATION_TYPE as NA.

    -   These represent the 'standard' retirees in the dataset.

-   Missing Occupation Group (20'694 rows)

    -   There are 20'694 rows total where OCCUPATION_TYE is NA.

    -   11'877 of these are Pensioners. This is slightly higher than the 11'855 sentinel count, implying that 22 pensioners have valid days but no job title.

    -   The remaining 8'817 rows are active workers who simply did not fill in their specific job title.

-   Hidden subgroup (127 rows)

    -   Total Pensioners (11'982) - Sentinel Pensioners (11'855) = 127 cases..

    -   These rows are labeled 'Pensioner' but have valid employment days and often specific job titles.

These inconsistencies necessitate further investigation into their age and employment history to determine if they are valid 'working retirees' or data entry errors.

NOTE: In the first 2 iterations of this sentinel value these specific findings were not identified. In the first iteration all NA-values in OCCUPATION_TYPE were changed to 'Unknown' and a Flag was made for those who were retired. In a second iteration the flag was dropped due to it's redundancy and the rows were imputed with 'Retired' instead of 'Unknown' where applicable. Due to a final data check and mediocre accuracy in the trained model, these new findings were identified.

```{r}
# 1. Prepare the specific subgroup data
working_pensioners_data <- data %>%
  filter(NAME_INCOME_TYPE == "Pensioner", DAYS_EMPLOYED != 365243) %>%
  mutate(
    EXPERIENCE_YEARS = abs(DAYS_EMPLOYED) / 365.25
  )

# Boxplot of Age for Working Pensioners specifically
# To see if they are actually 'Old' or if they are younger people on disability pension
working_pensioners <- working_pensioners_data %>% filter(NAME_INCOME_TYPE == 'Pensioner')

ggplot(working_pensioners, aes(x = NAME_INCOME_TYPE, y = AGE)) +
  geom_boxplot(fill = "orange", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Age Distribution of the 'Working Pensioners' Subgroup") +
  theme_minimal()

```

```{r}
# Scatter Plot: Focused on the Anomaly
ggplot(working_pensioners_data, aes(x = AGE, y = EXPERIENCE_YEARS, color = OCCUPATION_TYPE)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "The 'Working Pensioner' Subgroup",
    subtitle = "Age vs. Experience (N = ~127).",
    x = "Age (Years)",
    y = "Years Employed",
    color = "Occupation Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8)
  )
```

The boxplot above isolates the age distribution of the 'working pensioner' subgroup. Here we see a median age of approximately 42 years old, with a range from 25 to 60 years old. There is a complete absence of individuals over the standard retirement age (60+), which is statistically impossible for a genuine random sample of pensioners.

The age profile of this group is identical to that of the active workforce. It does not reflect a demographic of retirees or even typical disability pensioners (who would likely show a wider age variance including older individuals).

The scatterplot maps age against years employed, coloring points by there occupation type. T he subgroup is not limited to 'consultancy' or 'specialized' post-retirement roles. Instead, we see physically demanding and standard roles such as laborers, drivers, core staff, and security staff. Furthermore, the years-employed scales naturally with age, mimicking the standard accumulation of experience found in regular employees.

The "Working Pensioners" do not form a distinct cluster. They are buried within the main distribution of the "Working" class. If they were a distinct demographic (e.g., wealthy early retirees), we would expect them to cluster differently in terms of experience-to-age ratio. Their perfect alignment with the standard workforce further suggests they are standard workers, simply mislabeled.

Based on the evidence above, we conclude that these 127 rows represent classification errors. Due to the small amount of rows, we decide to drop them.

NOTE: This was done in the final iteration of data cleaning. The previous iteration including these rows got a maximum validation accuracy in the neural network of 84%.

```{r}
print(paste("Total rows before:", nrow(data)))

# 2. Filter out the specific anomaly
# Logic: Keep rows that are NOT (Pensioner AND have Valid Employment Days)
data <- data %>%
  filter(!(NAME_INCOME_TYPE == "Pensioner" & DAYS_EMPLOYED != 365243))

# 3. Verify the drop
print(paste("Total rows after:", nrow(data)))

# Double check: Should be 0
remaining_anomalies <- nrow(data[data$NAME_INCOME_TYPE == "Pensioner" & data$DAYS_EMPLOYED != 365243, ])
print(paste("Anomalies remaining:", remaining_anomalies))
```

We can now handle the remaining NA-values in OCCUPATION_TYPE.

```{r}
# 1. Reclassify OCCUPATION_TYPE
data <- data %>%
  mutate(
    OCCUPATION_TYPE = as.character(OCCUPATION_TYPE),
    OCCUPATION_TYPE = case_when(
      DAYS_EMPLOYED == 365243 ~ "Retired", # Only the valid pensioners remain now
      is.na(OCCUPATION_TYPE)  ~ "Unknown", # The missing workers
      TRUE ~ OCCUPATION_TYPE
    ),
    OCCUPATION_TYPE = as.factor(OCCUPATION_TYPE)
  )

# 2. Calculate ACTIVE_EMPLOYMENT_YEARS cleanly in one step
# We set sentinel values (365243) directly to 0 to avoid the "1000 year" glitch
data <- data %>%
  mutate(
    ACTIVE_EMPLOYMENT_YEARS = ifelse(
      DAYS_EMPLOYED == 365243, 
      0, # in a previous iteration this was set so that the years_worked was 60
      abs(DAYS_EMPLOYED) / 365.25
    )
  )

# 3. Histogram 
hist(data$ACTIVE_EMPLOYMENT_YEARS,
     breaks = seq(0, max(data$ACTIVE_EMPLOYMENT_YEARS, na.rm=TRUE) + 1, by=1),
     main = "Histogram of Employment Duration (Pensioners = 0)", 
     xlab = "Years Employed", 
     col = "lightblue",
     border = "white")
```

As we see a very skewed distribution we will make a log-transformation to use in the Neural Network.

Note: Empirical results from our previous neural network iterations confirmed that preserving 'Unknown' as a distinct category, rather than applying statistical imputation, significantly improved the model's ability to differentiate risk. This suggests that the absence of occupation data is Missing Not At Random (MNAR) and contains a latent 'informational signal', representing a specific risk profile among applicants who choose not to disclose their professional details. By avoiding methods like KNN or Mode imputation, we prevent the dilution of this signal and allow the network to learn the predictive weight of non-disclosure itself.

```{r}
# Log-transform ACTIVE_EMPLOYMENT_YEARS and drop the original
data <- data %>%
  mutate(ACTIVE_EMPLOYMENT_YEARS_LOG = log1p(ACTIVE_EMPLOYMENT_YEARS)) %>%
  select(-c(DAYS_EMPLOYED, ACTIVE_EMPLOYMENT_YEARS))
# Histogram of log-transformed ACTIVE_EMPLOYMENT_YEARS
hist(data$ACTIVE_EMPLOYMENT_YEARS_LOG,
     main="Histogram of Log-Transformed ACTIVE_EMPLOYMENT_YEARS",
     xlab="Log(ACTIVE_EMPLOYMENT_YEARS + 1)",
     col="lightblue")
```

## Analysis of family members and children

We next examine CNT_FAM_MEMBERS and CNT_CHILDREN.

```{r}
# Identify extreme outliers
outlier_fam_members <- data[data$CNT_FAM_MEMBERS > 10, ]
print(outlier_fam_members)
```

A record showing a 38-year-old with 12 children (and a family size of 14) is statistically improbable within this demographic context and likely represents a data entry error. Given the extreme rarity (2 rows), these were removed to ensure the model is not biased by impossible family-to-income ratios.

```{r}
data <- data[data$CNT_FAM_MEMBERS <= 10, ]
```

A mathematical constraint must exist where CNT_FAM_MEMBERS ≥ CNT_CHILDREN + 1. A validation check identified 31 instances where the total family count was strictly less than the number of children reported.

```{r}
# Identify anomalous rows where children outnumber total family members
anomaly_rows <- data[data$CNT_FAM_MEMBERS < data$CNT_CHILDREN, ]
print(nrow(anomaly_rows)) # 31 rows
```

To rectify inconsistencies without losing data, we applied a deterministic correction based on NAME_FAMILY_STATUS. This assumes the child count is the "source of truth" and adjusts the total family count to include the necessary adults.

-   If 'married' or 'civil marriage', the formula is CNT_CHILDREN + 2

-   if 'separated', 'widowed', or 'single', the formula is CNT_CHILDREN + 1

```{r}
# Identify logic mask
anomaly_mask <- data$CNT_FAM_MEMBERS < data$CNT_CHILDREN

# Apply corrections
data <- data %>%
  mutate(CNT_FAM_MEMBERS = case_when(
    anomaly_mask & NAME_FAMILY_STATUS %in% c("Married", "Civil marriage") ~ CNT_CHILDREN + 2,
    anomaly_mask & NAME_FAMILY_STATUS %in% c("Separated", "Widow", "Single / not married") ~ CNT_CHILDREN + 1,
    TRUE ~ CNT_FAM_MEMBERS
  ))

# Final verification
nrow(data[data$CNT_FAM_MEMBERS < data$CNT_CHILDREN, ]) # Should be 0
```

This correction resolved the logical anomalies in all 31 cases. By using the family status to determine the number of adults, we maintain the integrity of the family-unit data which is often a strong predictor for debt-to-income ratios in credit scoring.

## Analysis of income amount

This chapter outlines the analysis and treatment of AMT_INCOME_TOTAL variable, focusing on outlier detection, logical consistency with occupation, and data normalization.

```{r}
summary(data$AMT_INCOME_TOTAL)
```

The initial summary of \`AMT_INCOME_TOTAL\` reveals a highly skewed distribution with a significant gap between the median (\$157'500) and the maximum (\$6'750'000).

Visual inspection through histograms and boxplots confirms a heavy right-tail distribution dominated by extreme outliers.

```{r}
# Standard Histogram
hist(data$AMT_INCOME_TOTAL, main="Histogram of AMT_INCOME_TOTAL", col="lightblue")
```

```{r}
# Boxplot to visualize the distance of outliers from the interquartile range
boxplot(data$AMT_INCOME_TOTAL, main="Boxplot of AMT_INCOME_TOTAL", ylab="Income", col="lightgreen")
```

To determine if the high-income values are legitimate or data entry errors, we analyzed income against NAME_EDUCATION_TYOE and OCCUPATION_TYPE.

```{r}
# Income by Education
boxplot(data$AMT_INCOME_TOTAL ~ data$NAME_EDUCATION_TYPE, 
        main="Income by Education Type", ylab = "Income", xlab = "Education", col="blue")

```

```{r}
# Income by Occupation
boxplot(data$AMT_INCOME_TOTAL ~ data$OCCUPATION_TYPE, 
        main="Income by Occupation Type", ylab = "Income", xlab = "Occupation", col="blue")
```

```{r}
# Age vs Income
plot(data$AMT_INCOME_TOTAL ~ data$AGE, main="Age Versus Income", 
     ylab="Total Income", xlab="Age", pch=19, col=rgb(0,0,0,0.2))
```

```{r}
data %>%
  arrange(desc(AMT_INCOME_TOTAL)) %>%
  head(10)
```

Examination of the top earners reveals critical inconsistencies. In a standard economic context, it is highly improbable for low-skill or service-oriented roles to command multi-million dollar annual salaries. Records where individuals in low-skill roles (e.g., Laborers, Cleaning staff) earn over \$1,000,000 are treated as entry errors. Removing these ensures that the model does not learn false correlations between low-skill occupations and extreme wealth.

```{r}
impossible_jobs <- c("Laborers", "Cleaning staff", "Secretaries", 
                     "Low-skill Laborers", "Drivers", "Waiters/barmen staff")

# Filter logic: Remove high earners in low-skill categories
data <- data %>%
  filter( !(AMT_INCOME_TOTAL > 1000000 & OCCUPATION_TYPE %in% impossible_jobs) )
# Visualizing the normalized distribution
hist(data$AMT_INCOME_TOTAL, 
     main="Histogram of AMT_INCOME_TOTAL", 
     xlab="AMT_INCOME_TOTAL", col="lightblue")
```

After removing the high earners in low-skill categories, we still see outliers for managers, which should be removed.

```{r}
# Income by Occupation
boxplot(data$AMT_INCOME_TOTAL ~ data$OCCUPATION_TYPE, 
        main="Income by Occupation Type", ylab = "Income", xlab = "Occupation", col="blue")
```

```{r}
data <- data %>%
  filter(AMT_INCOME_TOTAL <= 3000000)

# Visualizing the normalized distribution
hist(data$AMT_INCOME_TOTAL, 
     main="Histogram of AMT_INCOME_TOTAL", 
     xlab="AMT_INCOME_TOTAL", col="lightblue")
```

We check the people who earn more than 1 million. Yet looking at their profile, we do not see any strange behavior in the data.

```{r}
# Count rows with Income > 1,000,000
count_high_earners <- data %>%
    filter(AMT_INCOME_TOTAL > 1000000) %>%
    nrow()

print(paste("Number of customers earning > 1M:", count_high_earners))

# view them to see if they look like valid data or errors
data %>%
    filter(AMT_INCOME_TOTAL > 1000000) %>%
    select(ID, AMT_INCOME_TOTAL, NAME_INCOME_TYPE, OCCUPATION_TYPE, AGE) %>%
    arrange(desc(AMT_INCOME_TOTAL))
```

Even after removing logical outliers, the income variable remains highly skewed. To stabilize variance and minimize the impact of legitimate high-income earners, a Log Transformation (log1p) was applied.

```{r}
# Applying log transformation to achieve a more Gaussian distribution
data <- data %>%
  mutate(AMT_INCOME_TOTAL_LOG = log1p(AMT_INCOME_TOTAL))

# Visualizing the normalized distribution
hist(data$AMT_INCOME_TOTAL_LOG, 
     main="Histogram of Log-Transformed AMT_INCOME_TOTAL", 
     xlab="Log(AMT_INCOME_TOTAL + 1)", col="lightblue")
```

The resulting log-transformed distribution is much closer to a normal distribution, providing a more stable input for the neural network and preventing the gradient from being dominated by extreme values.

Note: In previous iterations all values above the 99% percentile were dropped without checking any domain logic.

## Correlation

In this chapter we check for mutlicollinearity. Thfis is for different reasons than we would for a linear model. While Neural Networks are famously capable of handling non-linear relationships and redundant features, a correlation matrix is still av ital diagnostic tool during the data cleaning phase.

Even though NNs can technically learn to ignore redundant features, high multicollinearity can make the model's loss landscape 'flat' or unstable.

In terms if data-leakage this is only allowed to be done using the train data. Throughout this entire project we use seed 42 for obvious reasons. The data is splitted into train (70%), validation (15%) and test (15%).

```{r}
set.seed(42) # ANSWER TO EVERYTHING
# Ensure target_col is defined
target_col <- "status" 

# Create splits based on the target
y_factor <- factor(data[[target_col]])
test_idx <- caret::createDataPartition(y_factor, p = 0.15, list = FALSE)
test_data <- data[test_idx, ]
remain_data <- data[-test_idx, ]

val_p <- 0.15 / 0.85
val_idx <- caret::createDataPartition(factor(remain_data[[target_col]]), p = val_p, list = FALSE)
val_data   <- remain_data[val_idx, ]
train_data <- remain_data[-val_idx, ]
```

```{r}
# Calculate correlation matrix for numerical columns
# Note: Use 'pearson' for linear and 'spearman' for non-linear rank correlation
cor_matrix <- cor(train_data %>% select_if(is.numeric), use="complete.obs", method="pearson")

# Visualize with a heatmap
corrplot(cor_matrix, method="color", tl.cex=0.7, title="Correlation Matrix of Cleaned Features")
```

In the provided Correlation Matrix, we see a heatmap representing the strength and direction of linear relationships between numerical variables. The graph uses a color scale where dark blue indicates a strong positive correlation, dark red indicates a strong negative correlation, and white indicates no linear correlation.

The most prominent dark blue square occurs at the intersection of CNT_CHILDREN and CNT_FAM_MEMBERS. This indicates that as athe number of children increases, the total family size increases almost in lockstep.

In traditional statistics, high intercorrelation is problematic because it suggests that one feature is a linear combination of another. Dropping a redundant feature reduces the dimensionality of the problem, making the model faster to train and easier to interpret.

While Neural Networks are more robust to multicollinearity than linear models, it still influences the training process. High correlation can make the "loss landscape" flat in certain directions, which can cause the optimizer to oscillate or converge more slowly toward the best solution. The network might assign very large weights to one feature and very small weights to the other, even though they carry the same signal. This can make the model less stable when faced with new, unseen data.

Despite the high correlation between CNT_FAM_MEMBERS and CNT_CHILDREN seen in the matrix, we keep both for the following reasons:

-   Latent Information:By keeping both, we allow the Neural Network to implicitly calculate the number of adults (Adults=Family Members−Children). This distinguishes between a single-parent household and a dual-parent household of the same total size.
-   Interaction Effects:The network can learn complex, non-linear interactions between these two variables that a simple correlation coefficient cannot capture, such as how the financial burden of children might change depending on the total number of adults present.
-   Data Integrity:Since we have already performed deterministic imputation to ensure these columns are logically consistent (Total Family≥Children + 1), providing both features gives the network the "full picture" of the household structure.

## Final Data Audit and Chapter Conclusion

This section summarizes the final state of the dataset following the multi-stage cleaning process. We have successfully resolved structural anomalies in employment data, corrected logical errors in family units, and mitigated the influence of extreme income outliers.

## Summary of Engineered Features

The table below provides the final audit of transformed variables:

| Original Variable | Cleaned/Engineered Variable | AGE |
|------------------------|------------------------|------------------------|
| DAYS_BIRTH | AGE | Absolute value conversion to positive years. |
| DAYS_EMPLOYED | ACTIVER_EMPLOYMENT_YEARS_LOG | Sentinel values set to 0; log1p transformation applied |
| OCCUPATION_TYPE | OCCUPATION_TYPE | NA values imputed as 'Retired' for pensioners and 'Unknown' for workers. |
| CNT_FAM_MEMBERS | CNT_FAM_MEMBERS | Deterministic correction where FAM_MEMBERS \< CHILDREN |
| AMT_INCOME_TOTAL | AMT_INCOME_TOTAL_LOG | Removal of 'impossible' job/income pairs and log1p scaling |

## Dataset Dimensions

```{r}
dim(data)
```

# Feature Engineering

The feature engineering phase was initiated during the third iteration of the data cleaning and modeling process. Despite significant efforts in fine-tuning the Neural Network architecture, including adjustments to layer depth, dropout rates, and activation functions, validation accuracy remained persistently low and stagnant.

Analysis of the misclassifications suggested that the raw features lacked sufficient "signal" to distinguish between target classes, particularly for minority classes. To overcome this, we transitioned from basic data cleaning to active feature engineering. The goal was to create interaction terms and domain-specific ratios that provide the model with better context regarding a borrower's financial stability and life stage.

## Financial context: family and life stage ratios

A borrower’s raw income is less informative than their "disposable" income or earning power relative to their age and dependents. We created interaction terms to provide the model with this missing context.

### Income per Family Member

Total income does not account for the financial burden of dependents. We created INCOME_PER_FAMILY_MEMBER to capture the per-capita resource availability within the household. Given the high skewness of financial data, we applied a log transformation to normalize the distribution.

```{r}
# Interaction term for family-adjusted income
data$INCOME_PER_FAMILY_MEMBER <- data$AMT_INCOME_TOTAL / data$CNT_FAM_MEMBERS

# Log transformation to stabilize variance
data <- data %>%
  mutate(INCOME_PER_FAMILY_MEMBER_LOG = log1p(INCOME_PER_FAMILY_MEMBER)) %>%
  select(-c(AMT_INCOME_TOTAL,INCOME_PER_FAMILY_MEMBER))


# Histogram of INCOME_PER_FAMILY_MEMBER
hist(data$INCOME_PER_FAMILY_MEMBER_LOG,
     main="Histogram of INCOME_PER_FAMILY_MEMBER_LOG",
     xlab="Income per Member",
     col="lightblue")
```

## Target Distributions

```{r}
# Look at the distribution of the target variable histogram
ggplot(data, aes(x = status)) + 
  geom_bar(fill = "steelblue", color = "black") +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.5
  ) +
  labs(
    title = "Distribution of Target Variable (Status)",
    subtitle = "Majority Class (0) vs. Minorities (1–5) vs. Special (C, X)",
    x = "Status Category",
    y = "Count"
  ) +
  theme_minimal()
```

The target variable distribution reveals a severe class imbalance, with the majority class (0) overwhelmingly outnumbering minority delinquency categories (1–5) and special status codes (C, X). Such extreme skewness can cause a neural network to develop a strong majority-class bias, where it achieves high overall accuracy by simply predicting the most frequent class while failing entirely to identify high-risk borrowers. To mitigate this, common sampling techniques like oversampling minority instances via SMOTE or undersampling the majority class can be employed to balance the training signal. Alternatively, algorithmic adjustments such as applying class weights or utilizing Focal Loss can penalize errors in minority classes more heavily, forcing the network to prioritize difficult, underrepresented examples. Beyond sampling, architectural strategies like delayed generalization or monitoring for grokking, where the model suddenly learns to generalize long after traditional convergence, may help the network move beyond simple rote memorization of the majority distribution toward meaningful pattern recognition across all risk tiers.

# Preprocessing 2

This final phase of data preparation transitions the cleaned dataset into a machine-ready format. The primary objective is to transform categorical strings and boolean flags into numerical representations that the neural network can process, while removing redundant original variables to prevent multi-collinearity.

## Target Recoding

```{r}
# TARGET RECODING
# Mapping status levels to class IDs 0..7
data$status <- as.character(data$status)
data$target_class <- recode(data$status,
                            "X" = 0, "C" = 1, "0" = 2, "1" = 3,  
                            "2" = 4, "3" = 5, "4" = 6, "5" = 7) |> as.numeric()

data <- data %>% select(-status)
```

## Categorical Encoding: Ordinal vs. Nominal

We distinguish between two types of categorical data to preserve as much information as possible:

-   Ordinal Encoding: Education level is treated as a ranked variable (EDUCATION_ENCODED), as there is a clear logical progression from secondary education to an academic degree.

-   One-Hot Encoding: Nominal variables like CODE_GENDER, NAME_INCOME_TYPE, and our engineered OCCUPATION_GROUP have no inherent ranking. These are expanded into binary dummy columns to ensure the model does not assume a false mathematical relationship between categories.

```{r}
# EDUCATION (ORDINAL ENCODING)
data$EDUCATION_ENCODED <- case_when(
    data$NAME_EDUCATION_TYPE == "Lower secondary" ~ 0,
    data$NAME_EDUCATION_TYPE == "Secondary / secondary special" ~ 1,
    data$NAME_EDUCATION_TYPE == "Incomplete higher" ~ 2,
    data$NAME_EDUCATION_TYPE == "Higher education" ~ 3,
    data$NAME_EDUCATION_TYPE == "Academic degree" ~ 4,
    TRUE ~ NA_real_
)
data <- data %>% select(-NAME_EDUCATION_TYPE)

# NOMINAL ENCODING (ONE-HOT)
nominal_cols <- c("CODE_GENDER", "NAME_INCOME_TYPE", "NAME_FAMILY_STATUS", 
                  "NAME_HOUSING_TYPE", "OCCUPATION_TYPE")

data[nominal_cols] <- lapply(data[nominal_cols], as.factor)

data <- dummy_cols(
  data,
  select_columns = nominal_cols,
  remove_selected_columns = TRUE,
  remove_first_dummy = FALSE 
)
```

## Drop ID-column

Since ID is just an identifier and not a real feature, it should be dropped.

```{r}
data <- data %>%
  select(-ID)

# Final Structure Check
str(data)
```

## Note on Feature Scaling

Throughout the iterative development of the neural network, the strategy for feature scaling evolved to better align with the Rectified Linear Unit (ReLU) activation function and the specific distribution of our engineered features.

### Initial Iterations: Robust Scaler

In the first two iterations, Robust Scaler was employed as the primary normalization tool.

-   Mechanism: Unlike standard scaling which uses the mean and variance, the Robust Scaler removes the median and scales the data according to the Interquartile Range (IQR).
-   ustification: This method is specifically designed to handle datasets with significant outliers, such as the initial versions of our income and employment data. It ensures that extreme values do not exert undue influence on the feature's scale, preserving the relative relationships within the majority of the data.
-   Compatibility with ReLU: Robust Scaler typically produces a centered distribution with both positive and negative values. This works with ReLU because the network can learn to utilize the "active" positive region for significant signals while the negative region is zeroed out, effectively acting as a natural threshold for the neurons.

### Transition to Min-Max Scaling

In the final iteration, the scaling strategy was shifted to Min-Max Scaling to optimize performance with the ReLU activation function.

-   Mechanism: Min-Max Scaling transforms all features into a fixed range, typically [0, 1], by subtracting the minimum value and dividing by the range.
-   Superiority with ReLU:
    -   Avoiding Dead Neurons: ReLU outputs zero for any negative input ($f(x) = \max(0, x)$). By shifting the entire feature set to a positive [0, 1] range, we ensure that every feature starts within the active gradient region of the ReLU function. This reduces the risk of "dying ReLU" problems during the initial phases of training.
    -   Consistent Input Space: Standardizing all inputs to the same positive bounds allows for more stable weight initialization and prevents certain features from dominating the gradient updates simply due to their numerical magnitude.
-   Sparsity Preservation: Min-Max scaling is particularly effective at preserving the zero-values in sparse data, which is essential given our use of one-hot encoding for the OCCUPATION_GROUP.

### Leakage Prevention and Implementation

While Min-Max Scaling is a required step for neural network convergence, it is intentionally excluded from this preprocessing script.

Data Leakage Prevention: To prevent data leakage, the scaling parameters (minimum and maximum values) must be calculated solely on the training split. Implementation Flow: If parameters were calculated using the entire dataset, information from the test set would "leak" into the training process. Consequently, scaling is implemented within the neural network training script immediately following the data split.

# Neural Network

## Neural Network Type: The Multi-Layer Perceptron (MLP)

For this classification task, we selected a Multi-Layer Perceptron (MLP), the canonical form of a feed-forward neural network. While modern deep learning often focuses on specialized architectures like Convolutional Neural Networks (CNNs) for image data or Recurrent Neural Networks (RNNs) for sequential data, the MLP remains the optimal choice for tabular datasets for several key reasons:

1.  Data Independence: Our dataset consists of independent customer records. Unlike time-series data (e.g., stock prices), there is no temporal dependency between Row i and Row i+1. Unlike image data, there is no spatial correlation between Column A and Column B. CNNs and RNNs rely on these dependencies to function; applying them here would introduce unnecessary computational overhead without theoretical benefit.

2.  Universal Approximation: The Universal Approximation Theorem states that a feed-forward network with a single hidden layer containing a finite number of neurons can approximate any continuous function. Given the complex, non-linear relationships between financial variables (e.g., the interaction between Age, Income, and Family Size), an MLP provides the necessary capacity to map these inputs to the 8 target classes.

3.  Feature Interaction: Unlike linear models (e.g., Logistic Regression), which assume additive relationships, an MLP learns to combine features. A neuron in the second hidden layer might activate only when Income is Low AND Family Size is High, effectively automating the feature engineering process.

## Defining the Architecture

Designing the network topology is a balancing act between Capacity (the ability to learn complex patterns) and Generalization (the ability to perform well on unseen data). Our final architecture was derived through a systematic Grid Search, but guided by the following principles:

### Input Layer

The input layer size corresponds to the number of features after preprocessing. Since we employed One-Hot Encoding for categorical variables (like Occupation and Gender), our input vector is sparse and high-dimensional.\
We tried grouping the column OCCUPATION_TYPE. The goal was to ensure that, when looking at the training set, occupations with the same risk-category could be found in all target classes. Yet grouping efforts actually lost important information for the model, leading to lower class-specific accuracy.

```{r}
dim(data) # We have 50 features and 1 target
```

The size of the input layer of the neural network is the amount of columns, in this case 50.

### Hidden Layers (Depth and Width)

We experimented with depths ranging from 1 to 3 hidden layers. These layers often are represented by concepts:

-   First layer: Extraction of the first layer.

-   Second layer: Interactions between features.

-   Third layer: Abstractions, in this use case represented by risk profiles.

The Risk of Depth: The deeper a network gets, the more complex interactions become. Yet the signal is diluted in every layer. Similarly, the signal from the loss function dilutes as it propagates back through many layers. Furthermore, deep networks on small datasets (like ours) are prone to overfitting, memorizig the noise in the training set rather than the signal.

The Width Strategy (Cubic vs. Funnel): We tested two shape configurations. Funnel (Pyramid): Layers decrease in size (e.g., 512 to 256 to 128). This assumes the network should compress information into higher-level abstractions. Cubic (Rectangular): Layers maintain constant width (e.g., 256 to 256 to 256).

```{r}
# Define the Search Grid
runs_grid <- list(
  # --- 1. ARCHITECTURE ---
  # Search Size: Wide (1024) vs. Narrow (256)
  units1 = c(128, 256, 512, 1024),
  
  # Search Funneling: Tapered (128) vs. Cylindrical (256)
  units2 = c(64, 128, 256, 512),
  
  # Search Depth: 2 Layers (0) vs. 3 Layers (64)
  units3 = c(0, 64, 128, 256), 
  
  # --- 2. OPTIMIZATION ---
  # Search Learning Speed: Standard (1e-3) vs. Precise (3e-4)
  learning_rate = c(0.01),
  
  # Search Gradient Noise: Noisy (128) vs. Stable (256)
  batch_size = c(64),
  
  # --- 3. REGULARIZATION  ---
  dropout1 = c(0.0),
  dropout2 = c(0.0),
  dropout3 = c(0.0),
  
  # Search momentum
  momentum = c(0.9),
  # --- CONSTANTS ---
  patience = c(100),
  epochs = c(1000)
)
```

The goal of this grid search is to only see how high validation accuracy can get depending solely on the network architecture.

```{r}
tic()
tuning_run("Assignment_2_MLP_minmax.r", 
           runs_dir = "tuning_architecture_sgd", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = TRUE)

beep(sound=5)
toc()
```

From all the models, we will search for the model with the lowest validation loss since this is the most stable model. A model with higher accuracy and higher loss score is more unstable. While it gets more class labels correct, it is very unconfident or very wrong on the outliers. A high loss usually signals that the model has overfit the cross-entropy objective and is essentially "memorizing" the decision boundary. Therefore it leaves no room to improve with hyperparameter tuning later because the gradients are already exploding.

```{r}
# Load all runs from the directory
all_results <- ls_runs(runs_dir = "tuning_architecture_sgd")

# View them in the RStudio data viewer (Sortable table)
View(all_results)

# 3. Print just the top 5 models sorted by Validation Accuracy
top_models <- all_results %>%
  arrange((metric_val_loss)) %>%
  select(metric_val_accuracy, metric_val_loss, flag_units1, flag_units2, flag_units3) %>%
  head(5)

print(top_models)
```

We see that a simpler model is the more stable model. The danger of going too small though is underfitting (low accuracy and low loss). A microgrid with one-layer experiments was made to see if even simpler models still can find the patterns in the data. Then we would base the decision on Razor's principle \`The simplest model is the best'.

Based on multiple experiments, always in combination with different scalers and optimisers due to the iterative process. The best architectures found was:

-   Number of hidden layers: 2 layers

-   Number of units per layer: 128/64

### Activation Functions

We utilized the Rectified Linear Unit for our hidden layers. ReLU is the standard for deep learning because it introduces non-linearity without suffering from the vanishing gradient problem common in Sigmoid or Tanh functions. However, we discovered a critical dependency between our choice of scaler and this activation function. Initial Failure (Robust Scaler): In our early iterations, we used a Robust Scaler, which centers data and often produces negative values for inputs below the median. Since ReLU defines the output of any negative input as exactly 0, a significant portion of our input signal was immediately zeroed out (deactivated) upon entering the first hidden layer. This led to dead neurons and significant information loss. The Fix (Min-Max Scaler): To resolve this, we switched to Min-Max Scaling, which forces all input features into the range [0, 1]. By ensuring all input values are positive, we guarantee that they fall into the active linear region of the ReLU function (x \> 0) rather than the dead region. This alignment between our preprocessing and our activation function resulted in an immediate and significant increase in model accuracy.

Output Layer (Softmax): Since this is a multi-class classification problem with 8 mutually exclusive classes, the final layer uses the Softmax function. This converts the raw logits into a probability distribution summing to 1.0, allowing us to interpret the output as confidence percentages.

$$
\text{Softmax}(z_i) = \frac{e^{z_i}}{\sum_{j} e^{z_j}}
$$

Here we demonstrate the difference of both scalers with the relu activation function. In the following experiment a robustscaler was used with the same architecture as above.

```{r}
options(tfruns.runs_dir = "tuning_scaler")
tic()
training_run("Assignment_2_MLP_robustscaler.r")

beep(sound=5)
toc()
# Reset options
options(tfruns.runs_dir = NULL)
```

```{r}
run_path <- "tuning_architecture_minmax/2025-12-22T10-12-20Zs"

view_run(run_path)
```

## Class Imbalance Handling

Our target variable is severely imbalanced (approx. 99% Good vs 1% Bad/Late). A standard neural network minimizes global error by predicting the majority class for every instance, achieving 99% accuracy but 0% utility.

```{r}
tic()
# Set the "Bucket" 
options(tfruns.runs_dir = "tuning_imbalance")

#  Run SMOTE Script
training_run("Assignment_2_MLP_SMOTE.r")

# Run Undersampling Script
training_run("Assignment_2_MLP_undersampling.r")

# Run Weighted Class Scripts
training_run("Assignment_2_MLP_weighted.r")
training_run("Assignment_2_MLP_man_weight.r")

# Run Focal Loss Script
training_run("Assignment_2_MLP_focal_loss.r")


# 4. Finish
beep(sound=5)
toc()

# 5. Reset options (good practice)
options(tfruns.runs_dir = NULL)
```

Note: These tests were made with a different amount of layers and depth, compared to the optimal chosen architecture. Still, these tests can be compared among eachother to show whether or not the new method is superior to the raw-data.

### A. Data Level: SMOTE and Real Augmentation

Random oversampling, which involves simply duplicating rows, frequently leads to overfitting because the model memorizes specific examples rather than learning general patterns. To address this, we employed SMOTE. Instead of duplication, SMOTE creates new synthetic examples by interpolating between existing minority samples in the feature space. This forces the model to learn a more generalized decision boundary. We noted that ADASYN (Adaptive Synthetic Sampling) could represent a potential enhancement. ADASYN functions similarly to SMOTE but prioritizes generating synthetic data for examples located near the decision boundary, which are harder to classify. However, this variation was not implemented in this project.

**Analysis of Results**

The model achieved an overall accuracy of 85.9%, significantly outperforming the No Information Rate (NIR) of 77.1%. The NIR is the baseline accuracy in machine learning classification. It represents how well a model performs if it always guesses the most frequent class (majority class 2) in the dataset, without learning anything. The substantial gap between Accuracy and NIR, along with a Kappa score of 0.61 (indicating "substantial agreement"), confirms that the SMOTE-enhanced model has successfully learned distinctive patterns beyond just memorizing the majority class distribution.

The goal of using SMOTE was to improve the recognition of minority classes (Classes 4, 5, 6 and 7), which are often ignored during standard training.

The most telling metric here is Balanced Accuracy, which averages recall across classes. While the standard sensitivity (recall) for minority classes ranges from roughly 41% to 53%, the balanced accuracy remains consistently high, between 70% and 76% for these rare groups. this indicates that SMOTE successfully pushed the decision boundary to "pay attention" to these classes, preventing them from dropping to near-zero recognition.

For the rarest groups, the model exhibits a conservative bias. For example, in class 5, the model only identified 41% of the actual cases (sensitivity), but when it did predict class 5, it was correct 80% of the time (Precision). This suggests that SMOTE helped create high-confidence regions for these classes, even if those regions don't cover the entire spread of the minority data.

The confusion matrix reveals where the model still struggles despite oversampling:

-   Class 2 remains the dominant anchor of the model. However, significant "leakage" occurs between Class 2 and it's neighbors.

-   Classes 4 and 6 show lower Precision compared to Class 5. Specifically, Class 6 suffers from a very small sample size (only \~17 true instances total), yet the model successfully captured 9 of them.

In conclusion, the application of SMOTE resulted in a Macro F1 score of 0.616.

### Cherry picking

```{r}
# Set the "Bucket" 
options(tfruns.runs_dir = "tuning_imbalance")

# 1. Clean the Data
print(">>> Step 1/3: Running Data Cleaning...")
source("Kaggle_dataset_clean.r")
print(">>> Data Cleaning Complete.")

# 2. Run Exploratory Data Analysis
print(">>> Step 2/3: Running EDA...")
source("Kaggle_EDA.r")
print(">>> EDA Complete.")

# 3. Train Neural Network (Cherry Picking Strategy)
print(">>> Step 3/3: Running Neural Network (Cherry Picking)...")
source("NN_Cherrypicking_Kaggle.r")
print(">>> Pipeline Finished Successfully.")

# Reset options (good practice)
options(tfruns.runs_dir = NULL)
```

To address the severe class imbalance without relying solely on synthetic data (like SMOTE), we implemented a strategy of Selective Data Augmentation. We leveraged the original Kaggle source dataset to find "organic" examples of minority classes that were excluded from the assignment subset.

First, we reconstructed the target labels for the full Kaggle dataset by joining application_record.csv with credit_record.csv and identifying the worst delinquency status for each unique ID.

A critical challenge was ensuring no data leakage occurred between this new source and our validation sets. Since the assignment IDs were scrambled, we performed a rigorous content-based exclusion. We compared the two datasets based on their full feature vectors (dropping the ID column) and filtered the Kaggle dataset to retain only those rows that did not exist in the assignment data. This effectively created a "Rest of World" dataset containing unique, unseen customers.

From this pool of unique rows, we "cherry-picked" only the samples belonging to underrepresented classes (Status 1, 2, 3, 4, 5, X, and C), deliberately excluding the majority class (Status 0). This allowed us to inject real-world patterns of delinquency into the training set without further inflating the majority baseline.

This approach significantly enriched the training data with genuine minority examples. However, it introduced a distribution shift: by artificially boosting the prevalence of defaulters in the training set, we altered the prior probabilities compared to the validation set (which retained the natural, imbalanced distribution). Consequently, while the model’s Sensitivity (ability to detect risk) improved, the overall Accuracy decreased, as the model became more aggressive in predicting defaults than the natural base rate would suggest.

**Analysis of Results**

Contrary to the hypothesis that adding real-world minority examples would improve detection rates, the Selective Data Augmentation strategy resulted in a degradation of model performance across nearly all critical metrics.

-   The overall validation accuracy dropped from 86.04% (in the baseline Raw Model) to 84.77%. While a slight drop in accuracy is often an acceptable trade-off for higher sensitivity in imbalanced datasets, we did not observe the corresponding gain in minority detection.

-   The most significant failure occurred in the sensitivity (recall) of the high-risk classes, which the strategy was specifically designed to boost.

    -   Class 3 (Status 1): Sensitivity dropped precipitously from 61.0% to 49.1%.

    -   Class 4 (Status 2): Sensitivity declined from 51.5% to 44.7%.

    -   Class 0 (Status X): Sensitivity decreased from 55.8% to 53.6%.

-   The confusion matrix reveals that the model became less discriminative and more prone to classifying actual defaulters as "Good" clients (Class 2/Status 0). Specifically, the number of Class 3 (Status 1) defaulters misclassified as Class 2 rose from 335 in the raw model to 448 in this augmented model.

These results strongly suggest the presence of a covariate shift between the University assignment dataset and the raw Kaggle source. Although the external rows represented the same target labels (e.g., "Status 1"), their underlying feature distributions likely differed from the "slightly altered" University data. Consequently, the augmented data acted as noise rather than signal, blurring the decision boundaries and causing the model to revert to predicting the majority class more frequently. Due to this negative impact, this external augmentation strategy was discarded in favor of the baseline model.

### B. Data Level: Undersampling majority classes

To counter the dominance of the majority class, we implemented Random Undersampling. Unlike SMOTE, which adds synthetic data, undersampling balances the dataset by randomly removing examples from the majority class in the training set until its size matches that of the minority classes, or untill a threshold is reached. The theoretical advantage of this approach is that it forces the neural network to treat every class as, more or less, equally probable, preventing the optimization algorithm from getting stuck in a local minimum where it simply predicts "class 2" for everything to minimize loss.

However, this technique introduces a significant risk known as distribution shift. By artificially equalizing the class counts, we fundamentally alter the "prior probability" the model sees during training. In our training set, a rare class (like Class 5) might appear 12.5% of the time, but in the real-world test set, it only appears 0.2% of the time. Consequently, the model learns to expect these minority classes far more often than they actually occur. This leads to a high rate of False Positives: the model becomes "trigger-happy," guessing minority classes too frequently because it assumes they are common. Furthermore, by discarding a vast amount of majority class data, we risk losing valuable information about the variance and sub-clusters within the majority group.

**Analysis of Results**

The results of the undersampling experiment strictly confirm the theoretical risks described above. The most immediate observation is the drop in Overall Accuracy to 72.3%, which is notably lower than the No Information Rate (NIR) of 77.1%. Statistically, this model performs worse on a raw accuracy basis than a "dumb" model that simply predicts Class 2 for every instance.

However, accuracy is misleading here. The goal was to improve minority detection, and by that metric, the model succeeded, albeit at a heavy cost:

-   The distribution shift caused the model to drastically over-predict minority classes.

    -   For Class 0, the model predicted it 1'618 times, but only 644 were correct. A massive 846 instances of Class 2 were incorrectly classified as Class 0. This results in a very low Precision of 39.8%.

    -   Similarly, for Class 4, the Precision is only 33.8%, meaning two-thirds of the time the model predicted Class 4, it was wrong.

-   Conversely, the model is excellent at finding the minority classes when they do exist.

    -   Balanced Accuracy is exceptionally high across the board, ranging from 74% to 82%.

    -   Class 0 Sensitivity rose to 74.3%, and even the extremely rare Class 6 achieved a Sensitivity of 64.7%.

-   The trade-off is clear in the Class 2 column. The model only correctly identified 73.2%of the Class 2 instances (Sensitivity), whereas standard models usually achieve over 95%.

In summary, undersampling successfully "woke up" the model to the existence of rare classes, achieving high balanced accuracy. However, it did so by aggressively hallucinating minority labels on majority class data, resulting in a model that is sensitive but not precise.

### C. Algorithm Level: Class Weights

To address the imbalance problem without altering the training data itself (as in SMOTE or Undersampling), we applied Cost-Sensitive Learning. This technique modifies the loss function during training, assigning a higher penalty (cost) to misclassifications of minority classes. By telling the model that "missing a rare class is 50 times worse than missing a common class," we force the optimization algorithm to pay attention to the underrepresented groups.

We explored two strategies for assigning these weights:

1.  Computed Weights (Inverse Frequency): We first calculated weights mathematically to perfectly counterbalance the class frequencies. The weight wj​ for class j was calculated as$$w_j = \frac{N_{samples}}{N_{classes} \times N_j}$$

    This resulted in extreme penalties for the rarest classes (e.g., Class 6 received a weight of ca. 74.7, while the majority Class 2 received ca. 0.16).

2.  Manual Tuning: We hypothesized that the computed weights might be too aggressive, leading to overfitting or excessive false positives. Therefore, we performed a second experiment with Manually Tuned Weights, capping the maximum penalty at 5.0 and slightly down-weighting the majority class (0.5) to find a more stable "business-logic" balance.

**Analysis of Results**

The comparison between the two weighting strategies highlights a clear trade-off between detection sensitivity (catching the minority) and precision (avoiding false alarms).

#### 1. Computed Weights: The "Aggressive" Approach

The mathematically computed weights (ranging up to 74.7) succeeded in forcing the model to recognize minority classes but did so indiscriminately.

-   High Sensitivity, Low Precision: The model became "trigger-happy." For the rarest group, Class 6, the Sensitivity was 64.7%, which is excellent. However, the Precision was only 17.7%. This means that for every 100 times the model predicted Class 6, it was wrong 82 times.

-   Impact on Accuracy: Because the model was aggressively predicting rare classes where they didn't exist, the Overall Accuracy (78.5%) barely exceeded the No Information Rate (77.1%).

-   High Balanced Accuracy: The Balanced Accuracy of \~81% was the highest among methods, proving the model was strictly prioritizing average recall over total accuracy.

#### 2. Manual Tuning: The "Balanced" Approach

By dampening the weights (setting Class 6/7 to 5.0 and Class 2 to 0.5), we achieved a far more robust model.

-   Restored Accuracy: The Overall Accuracy jumped to 84.9%, a significant improvement over the computed weights. This indicates the model stopped "hallucinating" minority labels on clear majority cases.

-   Precision Recovery: The precision for minority classes improved drastically. For Class 6, Precision rose from 17.7% to 42.9%, and for Class 5, it rose from 25.4% to 68.2%.

-   The Trade-off: We accepted a slight drop in Balanced Accuracy (from \~81% down to \~78%). The Sensitivity for Class 6 dropped from 64.7% to 52.9%, meaning we miss a few more actual cases. However, the cases we do flag are much more likely to be correct.

While the computed weights provided the highest theoretical recall for minority groups, the manual weights offered a superior practical solution. The manual approach (Accuracy 84.9%, Macro F1 0.61) successfully identifies underrepresented classes without overwhelming the user with false positives, making it the more viable candidate for deployment.

### D. Loss Function: Focal Loss

In our final experiments, we replaced standard Categorical Cross-Entropy with Focal Loss.

Standard Cross-Entropy loss struggles with extreme imbalance because it is overwhelmed by the sheer volume of "easy" examples. In our dataset, Class 2 (Non-Default) comprises 77% of the data. Even if the model assigns a decent probability (e.g., 0.6) to these easy examples, the cumulative loss from thousands of them drowns out the valuable signal from the few misclassified minority examples.

Focal Loss adds a modulating factor, (1−pt​)γ, to the standard loss equation.

-   Effect: If the model is 99% confident in a prediction (an "easy" example), the modulating factor approaches 0. This effectively silences the loss contribution from the majority class.

-   Result: The network is forced to focus its weight updates almost entirely on the "hard" examples—specifically, the minority class errors where the model is currently failing.

**Analysis of Results**

Focal Loss emerged as the most "surgical" of all the methods tested. While Undersampling used a sledgehammer (deleting data) and Class Weights used a megaphone (amplifying penalties), Focal Loss acted as a scalpel, ignoring what was already learned to focus on what was difficult.

1\. Best-in-Class Precision The most striking result is the Macro F1 Score of 0.631, the highest among all tested methods (compared to \~0.61 for SMOTE and Manual Weights). This improvement is driven by superior Precision.

-   Unlike Undersampling, which hallucinated minority classes everywhere, Focal Loss remained conservative and accurate.

-   Class 7: Precision of 83.3% (it rarely guessed Class 7, but when it did, it was right).

-   Class 5: Precision of 75.0%.

-   Class 0: Precision of 70.1%. This indicates that Focal Loss successfully learned the distinctive features of these minority classes rather than just guessing them based on inflated priors.

2\. Preserving the Majority: Crucially, Focal Loss did not sacrifice the majority to save the minority.

-   Class 2 Sensitivity remained extremely high at 94.4%.

-   Overall Accuracy was 85.9%, effectively tying with SMOTE and significantly beating Undersampling (72%) and Computed Weights (78%).

3\. The Limit of "Hard" Examples (Class 6) However, the method hit a wall with Class 6.

-   Sensitivity: Only 35.3% (detecting 6 out of 17 instances).

-   Balanced Accuracy: Dropped to 67.6% for this class. Because Class 6 is so incredibly rare (only \~0.1% of data), it is possible these examples were treated as "outliers" rather than "hard examples," or simply that the model requires more representative features to distinguish them.

Focal Loss provided the best balance of the imbalance rectifying attempts. It achieved the highest F1 score, maintained high overall accuracy, and delivered reliable predictions (high precision) for the minority classes. It proves that intelligently re-weighting the loss gradients is often superior to artificially manipulating the training data.

#### Conclusion and Recommendation for Class Imbalance

We systematically evaluated four techniques to address the class imbalance against a baseline model trained on raw data. Our hypothesis was that the baseline would fail to detect the minority classes (late payers). However, the empirical results refuted this.

|  |  |  |  |  |
|---------------|---------------|---------------|---------------|---------------|
| **Model** | **Overall Accuracy** | **Macro F1 Score** | **Class 6 Sensitivity (Recall)** | **Class 6 Precision** |
| Baseline (Raw Data) | 86.04% | 0.648 | 58.8% | 62.5% |
| Focal Loss | 85.88% | 0.631 | 35.3% | 50.0% |
| SMOTE | 85.86% | 0.616 | 52.9% | 40.9% |
| Manual Class Weights | 84.94% | 0.612 | 52.9% | 42.9% |
| Cherry Picking | 84.77% | 0.606 | 52.9% | 64.3% |
| Undersampling | 72.34% | 0.518 | 64.7% | 0.1% (Approx) |

The table below summarizes the key performance metrics across all experiments:

-   **Undersampling** achieved the highest "detection rate" for rare classes but failed catastrophically on precision, generating so many false alarms that the model became practically unusable.

-   **Focal Loss** was excellent at increasing confidence (Precision) but surprisingly conservative, missing too many of the "hard" examples (lowest Recall for Class 6).

-   **The Baseline Model** struck the best balance. It not only achieved the highest Overall Accuracy and F1 Score but, crucially, it outperformed SMOTE and Focal Loss in detecting the rarest instances (Class 6 Recall of 58.8%).

**Business Implications: The Cost of a Miss vs The cost of a False Alarm**

For this specific business case, identifying payment behavior, the two critical errors have different costs:

1.  **False Negative (Missed Late Payer):** The bank fails to intervene, leading to a financial loss (Default). This is the most expensive error.

2.  **False Positive (False Alarm):** The bank flags a good customer as risky. This incurs a small administrative cost (checking the file) or a customer service risk (annoyance).

While we want to maximize the detection of late payers (high Recall), we cannot afford a model like Undersampling, where \~80-90% of the flagged customers are actually innocent. This would overwhelm the collections department and damage customer relationships.

The Baseline Model offers the most viable path for deployment. It catches the majority of late payers (high sensitivity) without drowning the operations team in false positives.

**Final Decision for Hyperparameter Tuning**

Based on these findings, we will proceed with the Baseline (Raw Data) approach for the final hyperparameter tuning phase.

The data suggests that the "imbalance" is not the primary blocker; the minority classes have distinctive enough features that a standard neural network can separate them if tuned correctly. Adding complexity via SMOTE or Focal Loss provided diminishing returns. Therefore, our final optimization efforts will focus on maximizing the architecture (layers, neurons, learning rate) of the standard model to squeeze out the remaining performance.

```         
```

### Choosing the Optimizer: The Shift from Adam to SGD

After choosing the architecture and the type of imbalance-fixing script (in this case going for the raw, cleaned data), we checked the different optimizers. Initially we trained all our models with Adam since this is the industry standard for rapid convergence. Though after frustrating results, we learned following:

"I trained for a long time, where nothing changed on the validation accuracy, yet suddenly the model learned and the accuracy increased".

This is a typical description for delayed learning or grokking. Grokking is a concept where researchers overfit the model, so the loss is minimal. With special optimisers and/or weight decay, the model downscales the weights over epochs, simplifying the model, and adjusts into a specific local minimum. Based on this assumption to be correct, we changed to delayed-learning optimisers like AdamW and SGD with Momentum. We therefore restarted the cleaning process to be sure of good data quality and conducted new grid searches for SGD with Momentum and AdamW. Once again the grid search's goal was to see how high the validation accuracy could get. Overfitting was ignored for the time being

```{r}
# Define the Search Grid
runs_grid <- list(
  # --- 1. ARCHITECTURE ---
  # Search Size: Wide (512) vs. Narrow (256)
  units1 = c(128),
  
  # Search Funneling: Tapered (128) vs. Cylindrical (256)
  units2 = c(64),
  
  # Search Depth: 2 Layers (0) vs. 3 Layers (64)
  units3 = c(0), 
  
  # --- 2. OPTIMIZATION ---
  # Search Learning Speed: Standard (1e-3) vs. Precise (3e-4)
  learning_rate = c(0.01, 0.001, 0.0001),
  
  # Search Gradient Noise: Noisy (128) vs. Stable (256)
  batch_size = c(32, 64, 128),
  
  # --- 3. REGULARIZATION  ---
  dropout1 = c(0.0),
  dropout2 = c(0.0),
  dropout3 = c(0.0),
  
  # Search weight_decay
  weight_decay = c(0.0001),
  # --- CONSTANTS ---
  patience = c(100),
  epochs = c(1500)
)


tic()
tuning_run("Assignment_2_MLP_adamw.r", 
           runs_dir = "tuning_adamw_v2", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = FALSE)
beep(sound=5)
toc()
```

### Adam and AdamW

We began with Adam (Adaptive Moment Estimation), the industry standard for rapid convergence. Adam computes individual adaptive learning rates for different parameters. This is theoretically advantageous for our sparse, One-Hot encoded data, as it boosts the learning rate for rarely occurring features (like specific job titles). These can be found in the 'tunin_architecture_search'.

Yet with the goal of delayed generalization, a grid search for the learning rate and weight decay for AdamW optimiser was made. AdamW is a variation of the Adam optimiser that fundamentally changes how weight decay is applied. To understand the difference, we have to look at how they handle regularization. In standard Adam (and most other optimisers like SGD), L2 regularization is implemented by adding a penalty term directly to the loss function. When the optimiser calculates the gradient of this total loss, the penalty gets mixed in with the gradient of the actual task. In Adam, this mixed gradient is then adapted by the moving averages (momentum and variance). This means the decay of the weights gets scaled by the same adatpive factor as the learning steps, which inadvertently shrinks weights with large gradients less than intended and weights with small gradients more than intended.

AdamW fixes this by decoupling the weight decay from the gradient update. Instead of adding the penalty to the loss function, AdamW applies the decay directly to the weights themselves after the main gradient update is calculated. It calculates the adaptive step based only on the original loss gradient, updates the weight, and then subtracts a tiny portion of the weight value (the decay) separately. This ensures that every weight decays at the same consistent rate regardless of how large or small its gradient is, which leads to better generalization and more stable training, especially in deep networks.

The hyperparameters control this decoupled process. Learning Rate determines the step size of the gradient update, just like in standard Adam. Weight Decay becomes a standalone parameter that strictly controls how much the weights shrink at each step, independent of the learning rate. Finally, Batch Size indirectly affects the noise level of the gradient estimates; smaller batches introduce more noise which can act as a regularizer, while larger batches provide more stable but potentially less generalizable updates.

```{r}
#
```

```{r}

```

The best model found in the grid search is following:

-   Architecture: 2 layers (128 and 64 neurons).

-   Learning rate: 0.0001

-   Batch size: 32

The decision is based on a combination of lowest loss and good accuracy. A loss score of 0.5341 was reached, which shows the most "calibrated" model. It makes fewest confident errors and is therefore most stable engine for future fine-tuning. Furthermore, the accuracy is one of the highest found within the grid, yet with a significantly better loss.

### Regularization tuning

```{r}
# Define the Search Grid
runs_grid <- list(
  # --- 1. ARCHITECTURE ---
  # Search Size: Wide (512) vs. Narrow (256)
  units1 = c(128),
  
  # Search Funneling: Tapered (128) vs. Cylindrical (256)
  units2 = c(64),
  
  # Search Depth: 2 Layers (0) vs. 3 Layers (64)
  units3 = c(0), 
  
  # --- 2. OPTIMIZATION ---
  # Search Learning Speed: Standard (1e-3) vs. Precise (3e-4)
  learning_rate = c( 0.0001),
  
  # Search Gradient Noise: Noisy (128) vs. Stable (256)
  batch_size = c(32),
  
  # --- 3. REGULARIZATION  ---
  dropout1 = c(0.0, 0.2, 0.4),
  dropout2 = c(0.0, 0.2, 0.4),
  dropout3 = c(0.0),
  
  # Search weight_decay
  weight_decay = c(0.1, 0.01, 0.001, 0.0001),
  # --- CONSTANTS ---
  patience = c(100),
  epochs = c(1500)
)


tic()
tuning_run("Assignment_2_MLP_adamw.r", 
           runs_dir = "tuning_adamw_v3", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = FALSE)
beep(sound=5)
toc()
```

```{r}
# Load all runs from the directory
all_results <- ls_runs(runs_dir = "tuning_adamw_v3")

# View them in the RStudio data viewer (Sortable table)
View(all_results)

# 3. Print just the top 5 models sorted by Validation Accuracy
top_models <- all_results %>%
  arrange((metric_val_loss)) %>%
  select(metric_val_accuracy, metric_val_loss, flag_units1, flag_units2, flag_units3) %>%
  head(5)

print(top_models)
```

To decide what model using AdamW optimiser is the best, we look at the generalization gap (the difference between Training Loss and Validation Loss). Based on this criteria, the best model using AdamW has following hyperparameter and architecture:

-   Architecture: 2 layers (128 and 64 neurons).

-   Learning rate: 0.0001

-   Batch size: 32

-   Weight decay: 0.01

-   Dropout: 0.2 / 0.0

```{}
```

### Threshold Tuning

```{r}
# Define the "Best Model" Configuration
runs_grid <- list(
  # --- 1. ARCHITECTURE ---
  units1 = c(128),
  units2 = c(64),
  units3 = c(0), 
  
  # --- 2. OPTIMIZATION ---
  learning_rate = c(0.0001),
  batch_size = c(32),
  
  # --- 3. REGULARIZATION ---
  dropout1 = c(0.2),
  dropout2 = c(0.0),
  dropout3 = c(0.0),
  weight_decay = c(0.01),
  
  # --- CONSTANTS ---
  patience = c(300), # Ensure your script uses this flag!
  epochs = c(5000)
)

tic()

# Launch the run
run_info <- tuning_run(
  file = "Assignment_2_MLP_adamw.r", 
  runs_dir = "tuning_adamw_v4", 
  flags = runs_grid, 
  sample = 1.0, 
  echo = FALSE
)

toc()
beep(sound=5)

# --- FIND YOUR MODEL ---
# The run is now saved in a timestamped folder inside "tuning_adamw_v4".
latest_run_dir <- run_info$run_dir[1]
message("Your final model is located in: ", latest_run_dir)
```

```{r}
# 1. Set the Exact Flags of your Winner
# (I am inferring these from your "Row 1" result earlier. 
#  Double check if your specific run had different values!)
FLAGS <- list(
  units1 = 128,
  units2 = 64,
  dropout1 = 0.0,
  dropout2 = 0.0,
  learning_rate = 0.01,      
  weight_decay = 0.001,     
  batch_size = 32,
  epochs = 1000,
  patience = 100
)

# 2. Build the Model Architecture (Exact copy of your script)
model <- keras_model_sequential() %>%
  # Layer 1
  layer_dense(units = FLAGS$units1, input_shape = c(ncol(x_train)), use_bias = FALSE) %>%
  layer_batch_normalization() %>%  
  layer_activation("relu") %>%
  layer_dropout(FLAGS$dropout1) %>%
  # Layer 2
  layer_dense(units = FLAGS$units2, use_bias = FALSE) %>%
  layer_batch_normalization() %>%
  layer_activation("relu") %>%
  layer_dropout(FLAGS$dropout2)
# Output Layer
model <- model %>%
  layer_dense(units = 8, activation = "softmax")

# 6. Compile -------------------------------------------------------------
model %>% compile(
  loss = "categorical_crossentropy",
  optimizer = tf$keras$optimizers$AdamW(
    learning_rate = FLAGS$learning_rate,
    weight_decay = FLAGS$weight_decay
  ),
  
  metrics = list("accuracy")
)

# 4. Train (Reproducing the winner)
history <- model %>% fit(
  x_train, y_train,
  epochs = FLAGS$epochs,
  batch_size = FLAGS$batch_size,
  validation_data = list(x_val, y_val),
  callbacks = list(callback_early_stopping(monitor = "val_loss", patience = FLAGS$patience, restore_best_weights = TRUE)),
  verbose = 1
)
```

**get raw scores**

gen proba for val set

val_probs \<- model %\>% predict(x_val)

verify schape

dim(val_probs)

**Find the Best Threshold for "Class 0" (The Hardest Class)**

library(pROC)

class_0_true \<- (apply(y_val, 1, which.max) == 1) \# TRUE if actually Class 0 class_0_prob \<- val_probs[, 1] \# Model's confidence for Class 0

roc_0 \<- roc(class_0_true, class_0_prob)

best_thr_0 \<- coords(roc_0, "best", best.method = "closest.topleft", ret = c("threshold", "sensitivity", "specificity"))

print(best_thr_0) \# Look at the 'threshold' value (e.g., might be 0.23 instead of 0.5)

**Apply handicap**

tune_vec \<- c(0.25, 0.25, 0.70, 0.30, 0.50, 0.50, 0.15, 0.40)

tuned_probs \<- sweep(val_probs, 2, tune_vec, FUN = "/")

tuned_preds \<- apply(tuned_probs, 1, which.max) - 1

**Check**

val_true_labels \<- apply(y_val, 1, which.max) - 1

library(caret) cm_tuned \<- confusionMatrix( factor(tuned_preds, levels = 0:7), factor(val_true_labels, levels = 0:7), mode = "everything" )

print(cm_tuned)

### SGD with Momentum

```{r}
# Define the Search Grid
runs_grid <- list(
  # --- 1. ARCHITECTURE ---
  # Search Size: Wide (512) vs. Narrow (256)
  units1 = c(128),
  
  # Search Funneling: Tapered (128) vs. Cylindrical (256)
  units2 = c(64),
  
  # Search Depth: 2 Layers (0) vs. 3 Layers (64)
  units3 = c(0), 
  
  # --- 2. OPTIMIZATION ---
  # Search Learning Speed: Standard (1e-3) vs. Precise (3e-4)
  learning_rate = c(0.01, 0.001, 0.0005, 0.0001),
  
  # Search Gradient Noise: Noisy (128) vs. Stable (256)
  batch_size = c(32, 64),
  
  # --- 3. REGULARIZATION  ---
  dropout1 = c(0.0),
  dropout2 = c(0.0),
  dropout3 = c(0.0),
  
  # Search momentum
  momentum = c(0.9, 0.95, 0.99),
  weight_decay = c(0.0001),
  # --- CONSTANTS ---
  patience = c(100),
  epochs = c(1500)
)

tic()
tuning_run("Assignment_2_MLP_minmax.r", 
           runs_dir = "tuning_sgdmomentum_v2", 
           flags = runs_grid, 
           sample = 1.0, # Run ALL combinations (set to e.g. 0.5 to run random 50%)
           echo = FALSE)
beep(sound=5)
toc()
```

Following the evaluation of adaptive methods, we turned to Stochastic Gradient Descent (SGD) with Momentum. While adaptive optimizers like Adam often provide faster initial convergence, SGD remains a crucial baseline in deep learning. It is often cited in literature for achieving better long-term generalization and finding flatter minima, provided the hyperparameters are tuned precisely.

To understand the necessity of the "Momentum" term, we must look at the limitations of standard SGD. In standard SGD, the weight update is determined solely by the gradient of the current batch. This can lead to two issues:

1.  Noisy Updates: Because the batch is a random sample, the gradient can fluctuate wildly, causing the optimization path to "zigzag" rather than moving directly toward the minimum.

2.  Ravines: In areas where the surface curves much more steeply in one dimension than another, standard SGD tends to oscillate across the narrow ravine without making significant progress along the bottom.

Momentum addresses these issues by introducing the physical concept of inertia to the optimization process. Instead of calculating the update based only on the current gradient, the optimizer maintains a running average of past gradients (the velocity). At each step, the new update is a combination of the current gradient and the previous velocity vector.

Mathematically and intuitively, this acts like a ball rolling down a hill. As it descends, it accumulates speed (momentum). If it encounters small bumps or noisy gradients (local irregularities), the accumulated velocity carries it through, smoothing out the optimization path. Consequently, dimensions with consistent gradient directions accelerate, while dimensions with oscillating gradients cancel each other out.

```{r}
# Load all runs from the directory
all_results <- ls_runs(runs_dir = "tuning_sgdmomentum")

# View them in the RStudio data viewer (Sortable table)
View(all_results)

# 3. Print just the top 5 models sorted by Validation Accuracy
top_models <- all_results %>%
  arrange((metric_val_loss)) %>%
  select(metric_val_accuracy, metric_val_loss, flag_units1, flag_units2, flag_units3) %>%
  head(5)

print(top_models)
```

A grid search was conducted to optimize the learning rate, momentum, and weight decay. The best performing configuration for SGD was:

-   Architecture: 2 layers (128 and 64 neurons)

-   Learning Rate: 0.0001

-   Momentum: 0.9

-   Weight Decay: 0.001

-   Batch Size: 64

The model achieved a validation loss of 0.5635 and a validation accuracy of 82.8%.

While the accuracy is competitive, the loss is notably higher than that of the AdamW model (0.5046). This result highlights the trade-offs inherent in non-adaptive optimizers. The high momentum value of 0.9 helped stabilize the updates, but the fixed learning rate (1×10−4) likely limited the model's ability to navigate the sparse, One-Hot encoded features as effectively as Adam's per-parameter adaptive rates. The larger batch size of 64 was chosen to provide a more accurate estimate of the gradient, compensating for the lack of adaptive scaling, but ultimately, the model was less "calibrated" than its AdamW counterpart.

##### regularization

weight_decay = c(0, 0.0001, 0.001), # Search Dropout # None vs. Light vs. Moderate dropout1 = c(0.0, 0.2, 0.4), dropout2 = c(0.0, 0.2, 0.4), dropout3 = c(0.0),

### Threshold Tuning

First we retrain the winning model architecture for SGD with momentum.

```{r}
# 1. Set the Exact Flags of your Winner
# (I am inferring these from your "Row 1" result earlier. 
#  Double check if your specific run had different values!)
FLAGS <- list(
  units1 = 128,
  units2 = 64,
  units3 = 0,
  dropout1 = 0.0,
  dropout2 = 0.0,
  dropout3 = 0.0,
  learning_rate = 0.001,      # From your grid
  momentum = 0.9,             # From your grid
  weight_decay = 0.0001,      # From your grid (Assuming this was the best)
  batch_size = 64,
  epochs = 1500,
  patience = 100
)

# 2. Build the Model Architecture (Exact copy of your script)
l2_reg <- regularizer_l2(l = FLAGS$weight_decay)

model <- keras_model_sequential() %>%
  layer_dense(units = FLAGS$units1, input_shape = c(ncol(x_train)), use_bias = FALSE, kernel_regularizer = l2_reg) %>%
  layer_batch_normalization() %>% layer_activation("relu") %>% layer_dropout(FLAGS$dropout1) %>%
  layer_dense(units = FLAGS$units2, use_bias = FALSE, kernel_regularizer = l2_reg) %>%
  layer_batch_normalization() %>% layer_activation("relu") %>% layer_dropout(FLAGS$dropout2) %>%
  layer_dense(units = 8, activation = "softmax")

# 3. Compile
model %>% compile(
  loss = "categorical_crossentropy",
  optimizer = optimizer_sgd(learning_rate = FLAGS$learning_rate, momentum = FLAGS$momentum, nesterov = TRUE),
  metrics = list("accuracy")
)

# 4. Train (Reproducing the winner)
history <- model %>% fit(
  x_train, y_train,
  epochs = FLAGS$epochs,
  batch_size = FLAGS$batch_size,
  validation_data = list(x_val, y_val),
  callbacks = list(callback_early_stopping(monitor = "val_loss", patience = FLAGS$patience, restore_best_weights = TRUE)),
  verbose = 1
)
```

## Choosing the best experiment

The entire process is more iterative than this document shows. When finding the best architecture, different grid searches are made with changing hyperparameters and scripts to solve the imbalance. All tests are logged by tfruns and can be checked later.

Note: The tuning-folders have not been updated correctly during testing. Good experiment management would have likely make the understandability of what was changed in what run higher.

```{r}
# Get all runs
all_runs <- ls_runs(runs_dir = "/Users/Jujou/Documents/Repos/GRP-6_DS-Project/DS-Project_Part2_Scripts/tuning_architecture_search")

# View top models sorted by Validation Accuracy
head(ls_runs(order = metric_val_loss, decreasing = FALSE), 10)
```

## Hyperparameter Tuning and Training Strategy

### Grid Search Implementation

Given the computational cost of training for 5,000 epochs, standard K-Fold Cross-Validation was infeasible. We utilized a rigorous Grid Search on a fixed validation set.

Search Space: Neurons: [64, 128, 256, 512] Dropout: [0.1, 0.2, 0.3, 0.5] Batch Size: [32, 64, 128]

Why specific ranges? We avoided neurons \> 512 as the model size became disproportionate to the dataset size (leading to immediate overfitting). We avoided Batch Sizes \> 128 because smaller batches introduce noise into the gradient estimate, which, paradoxically, acts as a regularizer and helps escape sharp local minima.

### Regularization: Dropout

To further prevent overfitting, we inserted Dropout layers after each hidden layer. Dropout randomly deactivates a percentage of neurons (e.g., 30%) during each training pass.

Justification: This prevents neurons from co-adapting (relying on each other to correct errors). It forces each neuron to learn useful features independently, effectively training an ensemble of exponentially many thinner networks within a single model.

### Manual Learning Rate Scheduling and Grokking

We observed a phenomenon known as Grokking or delayed generalization, where the model's validation accuracy would stagnate for hundreds of epochs before suddenly improving. To exploit this, we abandoned automatic decay (like ReduceLROnPlateau) in favor of a Manual Scheduler.

Strategy: We trained with a relatively high learning rate (0.01) for the first 1,000 epochs to explore the loss landscape broadly.

Refinement: Once the loss stabilized, we manually intervened to drop the learning rate by a factor of 10 (0.001). This allowed the optimizer to settle into the deepest part of the global minimum without overshooting. This manual control gave us the stability required to train for extended durations (5,000 epochs).

# Choosing the best model
